% vim: tw=78 ai spell spelllang=en

\documentclass[10pt]{llncs}
\usepackage{color}
\usepackage{url}
\pagestyle{plain}

\newcommand\Color[2]{\textcolor{#1}{#2}}
\newcommand{\SM}{{\textit{SM}}}

\begin{document}

\title{Checking Properties Described by State Machines: On Synergy of
Instrumentation, Slicing, and Symbolic Execution}
\author{Jiri~Slaby, Jan~Strej\v{c}ek, and Marek~Trt\'{\i}k}
\institute{Faculty of Informatics, Masaryk University\\
  Botanick\'a 68a, 60200 Brno, Czech Republic\\
  \email{\{slaby,strejcek,trtik\}@fi.muni.cz}}

\maketitle

\begin{abstract}
  We introduce a novel technique for checking properties described by finite
  state machines. The technique is based on a synergy of three well-known
  methods: instrumentation, program slicing, and symbolic execution.
  The instrumentation works on the source code level, the other two on the top
  of \textsc{Llvm}.
  
  More precisely, the instrumentation step takes as input a state machine
  describing some behavior of programs and generates a code that corresponds
  to state machine transitions. There the error transitions of the state
  machine are represented as \texttt{assert} calls.  This code is then
  instrumented into a given program's source code.
  
  Having the \texttt{assert} calls in the code, we immediately have the
  slicing criteria too. The criteria are always a pair of program variables
  with a corresponding program location. These are exactly referenced
  variables in an \texttt{assert} condition and the line with the
  \texttt{assert} call proper.  Hence we can slice the program to reduce its
  size without affecting runs of state machines. In fact, the slicing step
  wipes out over 60\% of the code in average. Not only the reduced code size
  is an improvement for symbolic execution, but also the loops which are no
  longer present in the sliced code.
  
  After the slicing is finished, we can symbolically execute the sliced
  program to find real violations of the checked properties, i.e.~real bugs.
  For that purpose we use a freely available executor \textsc{Klee}. Again,
  slicing improves symbolic execution performance a lot. With no slicer used,
  the symbolic executor is able to finish the computation only on nearly 1\%
  of the files. With slicer, it is raised to over 21\%.
  
  Depending on the kind of symbolic execution, the technique can be applied as
  a stand-alone bug finding technique, or to weed out some false-positives
  from an output of another bug-finding tool. The latter stands as an
  improvement of techniques like abstract interpretation. It takes their
  output and filters false-positives out of that.

  If used as a stand-alone technique, it will take much longer to check the
  same files in comparison to other simpler techniques. On the other hand it
  will find only real bugs, i.e. no false-positives at all.
  
  Finally, we provide several examples demonstrating the practical
  applicability. They are taken from a real-world code, from the Linux kernel.
\end{abstract}

\end{document}
