% vim: tw=78 ai spell spelllang=en

\documentclass[10pt]{llncs}
\usepackage{bold-extra}
\usepackage{xcolor}
\usepackage{url}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{multirow}
\usepackage{xspace}
\pagestyle{plain}

%{{{ (La)TeX definitions

\newcommand\Color[2]{\textcolor{#1}{#2}}
\newcommand{\nb}[1]{\marginpar{\scriptsize #1}}
\renewcommand\note[1]{\nb{\Color{blue}{#1}}}
\newcommand\todo[1]{\nb{\Color{red}{#1}}}
\newcommand\xnote[1]{\Color{green!40!black}{#1}}
\newcommand\snote[1]{\nb{\Color{green!40!black}{#1}}}
\lstset{basicstyle=\ttfamily\scriptsize,escapeinside={(*@}{@*)}}

\newcommand\clang{\textsc{Clang}}
\newcommand\stanse{\textsc{Stanse}}
\newcommand\undetI{\textsc{Undetermined I}}

\newcommand\cdb{\textsc{ClabureDB}}

%}}}

\begin{document}

%{{{ Title + Author + Abstract

\title{\cdb: Classified Bug-Reports Database}
\author{Ji\v{r}\'\i~Slab\'y, Jan~Strej\v{c}ek, and Marek~Trt\'{\i}k}
\institute{Faculty of Informatics, Masaryk University\\
  Botanick\'a 68a, 60200 Brno, Czech Republic\\
  \email{\{slaby,strejcek,trtik\}@fi.muni.cz}}

\maketitle

\begin{abstract}
  We present a database that can serve as a tool for tuning and evaluation
  of various program analysis tools. The database contains a heap of bug
  reports produced by several tools applied to the Linux kernel. The reports
  are classified as real errors or false positives. The database has a web 
  frontend and it is also available for download. Using the database,
  developers of various bug-finding tools can easily evaluate performance of
  their tool on a large real-life software project.
\end{abstract}

%}}}

%{{{ Intro

\section{Introduction}
\label{sec:introduction}

Many successful bug-finding tools based on various program analysis methods
appeared during the last ten years. None of them is perfect. Each tool
either reports both real errors and false positives, or it discovers only a
part of real errors. To improve or evaluate such a tool, one needs a suitable
benchmark, i.e.~a set of programs with a (preferably complete) list of their
errors. There are several benchmarks for bug-finding tools (we refer
to~\cite{CHK09} for an overview). These benchmarks usually contain only
small and/or artificially generated programs. \textbf{(Sem dat
  odkazy na existujici benchmarky a vylepsit text)}.

As far as we know, there is no big real-life project with a significantly
long list of uniformly described errors. To fill this gap, we decided to
augment and open our database of Linux kernel (version 2.6.28) errors to the
public. The database contains bug reports produced by several bug-detection
tools. The reports have been manually classified by skilled programmers with
a help of Linux kernel developers. The reports have been marked as real errors
or false positives (only a few of them remain unclassified). The presence of
false positives in the database partly solves the problem that our database
likely does not contain all errors in the kernel.\note{Tu posledni vetu jsem
nezachytil...}

The database currently contains nearly 500 bug reports describing 11 kinds
of errors. The structure of the database and considered kinds of errors are
described in Section~\ref{sec:structure}. An overview of the current data
and their sources are presented in Section~\ref{sec:data}. 

The database can be accessed via a web interface or downloaded in
\textsc{SQLite~3} format for local use under the \emph{Open Database License
  (ODbL) v1.0}~\cite{ODbL}. The database (including documentation) is
available at:
\begin{center}
  \url{http://claburedb.fi.muni.cz/}
\end{center}

\textbf{Pridat sem neco konkretnejsiho, jak se databaze da vyuzit? Pridat
  sem vic related work, nebo informaci, ze zpusoby vyhodnoceni uspesnosti
  toolu byly navrzeny v~\cite{CHK09}?}

%}}}
%{{{ Database Structure

\section{Database Structure}\label{sec:structure}
The database is designed to conceive various kind of errors from various
projects and their version. However mixing errors from different projects or
version makes the work with the database hard. We decided to separate each
project and version. Hence the user selects project and version and then work
with a chosen sub-database. Such sub-database comprises information about
errors, tools, error types, users, and relation between them.

Namely we have several tables for each project and version:
\begin{itemize}
\item \textbf{error\_type} -- specifies various kinds of errors (cf.
Section~\ref{sec:error_type}). It contains name of the type, short description
and a reference CWE number if exists (see later).
\item \textbf{error} -- an error specified by the type above (and subtype if
needed), location (file, line), URL for reference, classification (false
positive, real error, unknown), confirmation (for real bugs, a commit ID for
instance), user who inserted the entry, and timestamp of the moment of
insertion.
\item \textbf{tool} -- the database contains errors from more tools, here they
are enumerated. We store name, version, URL and description for each of tools.
\item \textbf{user} -- each user, who inserts entries to the database, must
have name, affilitation and login in the database.
\item \textbf{error\_rel\_tool} -- since there is an m:n relation between
\texttt{error} and \texttt{tool} tables, we have this relation to provide the
mapping.
\end{itemize}


%Information stored in the database (snapshot?)

%user(name affilitation login password)

\subsection{Error Types} \label{sec:error_type}
All the tools produce only a limited kinds of errors. We investigated their
output and established a set of errors to put in the database. Since there
already exists a publicly available database of error types, we decided to
leverage that. It summarizes the error type, provides examples and also points
to real errors of a particular type. The database is called \emph{Common
Weakness Enumeration} (CWE)~\cite{cwe}. Some of our errors are still not
present in CWE though. For those we obviously do not provide a CWE number.
Here, we list all error types currently available in \cdb.

\noindent \textbf{BUG/WARNING} (CWE 617) \\ \hangindent=2em
Developers often inject \emph{assert}s to their code to ensure the input to
the code is as they expected.  Sometimes these assertions are hit and crashes
are enforced, thus we include them in \cdb.  The location is specified by the
line where the macro/function appears in the code.

\noindent \textbf{Double Lock} (CWE 764) \\ \hangindent=2em
Some lock is unintentionally locked twice in a sequence and it leads to an
inconsistent lock state. The location is specified by the line where the
second lock happens.

\noindent \textbf{Double Unlock} (CWE 765) \\ \hangindent=2em
Some lock is unintentionally unlocked twice in a sequence and it leads to an
inconsistent lock state. The location is specified by the line where the 
second unlock happens.

\noindent \textbf{Leaving function in locked state} \\ \hangindent=2em
Some lock is not unlocked on all paths of a function, so it is unintentionally
leaked. This often results in deadlocks on next invocation of such function.
The location is specified by the line of \texttt{return} statement or closing
brace if there is no \texttt{return}.

\noindent \textbf{Use After Free} (CWE 416) \\ \hangindent=2em
Dereferencing a pointer that points to an already freed memory. Often, the
code overwrites memory which is no longer owned. The location is specified by
the line of the use.

\noindent \textbf{Double Free} (CWE 415) \\ \hangindent=2em
Freeing function is called twice on the same address. Most allocators can
detect this and usually kill the program. The location is specified by
the line of the invalid free.

\noindent \textbf{Memory Leak} (CWE 401) \\ \hangindent=2em
Some code omits to free a memory which was allocated. The location is
specified by the line of the allocation.

\noindent \textbf{NULL Pointer Dereference} (CWE 476) \\ \hangindent=2em
A pointer which is NULL is being dereferenced. This typically happens when a
code allocates memory and dereferences it without prior checking it against
NULL. The location is specified by the line of the dereference.

\noindent \textbf{Double Resource Put} (CWE 763) \\ \hangindent=2em
The code requests one copy of a resource from a system, but there is a try to
return that to the system twice. The location is specified by the line of the
second put.

\noindent \textbf{Resource Leak} (CWE 404) \\ \hangindent=2em
The code requests a resource from a system, but omits to return that back to
the system for reuse. The location is specified by the line of the request.

\noindent \textbf{Calling function from invalid context} \\ \hangindent=2em
Some function is called at inappropriate place like calling sleep inside
critical sections or in interrupt handlers. Results are undefined, the system
may become unresponsive. The location is specified by the line of the function
call.

%}}}
%{{{ Current Data and Their Sources

\section{Current Data and Their Sources}\label{sec:data}
Currently, the database comprises 498 reports in 11 categories.  118 of the
reports are real, 353 false positives. The database contains also 27 entries
which were left unclassified. Table~\ref{tab:results} depicts how each
kind of bugs is represented in the database.

\begin{table}[tb]
\begin{center}
\begin{tabular}{|l|r|r|r|r|} \hline
\textbf{Error Type} & \textbf{Bugs} & \textbf{FP} & \textbf{Uncl.} &
	\textbf{Reports} \\ \hline

BUG/WARNING & 8 & 0 & 0 & 8 \\ \hline
Calling function from invalid context & 16 & 19 & 0 & 35 \\ \hline
Double Free & 0 & 1 & 0 & 1 \\ \hline
Double Lock & 16 & 60 & 0 & 76 \\ \hline
Double Resource Put & 3 & 4 & 0 & 7 \\ \hline
Double Unlock & 11 & 38 & 3 & 52 \\ \hline
Leaving function in locked state & 27 & 150 & 0 & 177 \\ \hline
Memory Leak & 7 & 13 & 0 & 20 \\ \hline
NULL Pointer Dereference & 17 & 14 & 0 & 31 \\ \hline
Resource Leak & 13 & 51 & 24 & 88 \\ \hline
Use After Free & 0 & 3 & 0 & 3 \\ \hline

\textbf{Overall Count} & 118 & 353 & 27 & 498 \\ \hline
\end{tabular}
\end{center}
\caption{Bugs in the database.}
\label{tab:results}
\end{table}

The data were collected thanks to the following tools: \clang\ 3\footnote{We
used modified \texttt{experimental.unix.PthreadLock} checker, where support
for kernel locking was added.}, \stanse, one tools which wanted to stay
in anonymity, and from kernel and Novell bug tracking systems and mailing
lists using our own web crawler.

%}}}
%{{{ Future Plans

\section{Future Plans}

We plan to extend the database in several ways. We will extend the data if we
find another source of bug reports. Further, we may extend the number of
considered kinds of errors. Finally, we think about building similar databases
also for newer versions of the Linux kernel or for other real-life projects.
The future of our database depends also on a feedback of the program analysis
community. Hence, if you have any suggestions, source of bug reports, or
another comments, please contact us \texttt{claburedb@fi.muni.cz}.

%}}}

\section{Poznamky}
\begin{itemize}
\item nasadit databazi
\item dat na stranky balik s analyzovanym jadrem
\item budou k dispozici statictiky pristupu k databazi?
\end{itemize}

\clearpage
\section{Related Work}

\xnote{Sem dodam behem pondelka strucny text (odstavecek) shrnujici informace v nasledujici subsekci~\ref{sec:relwork:notes}. Ten text, ktery sem dopisu, se pak bude dat kamkoliv presunout (napr. do sekce Introduction, jak jsme se domluvili...).}

\subsection{Notes (they are not meant to appear in the article)}\label{sec:relwork:notes}

\subsubsection{BegBunch -- Benchmarking for C Bug Detection Tools~\cite{CHK09}} This benchmark suite consists of two suites of different purpose. The \emph{Accuracy suite} consists of small benchmarks, called bug-kernels, which were extracted from existing buggy open source applications, like OpenSolaris or MySQL. The bug-kernels was extracted by inspecting bug tracing systems and mailing lists. The bug-kernels contains errors of the following types: buffer overflow, memory/pointer issues, integer overflow, and format string. Each bug in a bug-kernel is annotated with an XML-like markup of the same line that exposes the bug. The second suite \emph{Scalability suite} is composed of a particular versions of several open source applications (like MySQL 5.0.51a, Perl 5.8.4, Sendmail 8.12.3, etc.) of various size from small to large. There is no bug reports considered in this suite. Only size of benchmarks is considered. The purpose of this suite is to measure how well a program analysis tool scales in spite of size of an analysed code.

\subsubsection{BugBench: Benchmarks for Evaluating Bug Detection Tools~\cite{LLQ05}} Benchmarks are in the form of complete program distributions. It is composed of 17 programs (like gzip-1.2.4, cvs-1.11.4, msql-4.1.1, etc.) including 19 bugs (like stack smash, global buffer overflow, heap buffer overflow, uninitialised read, memory leak, double free, data race,  atomicity, and semantic bug) that were known to exist in these programs.

\subsubsection{On Establishing a Benchmark for Evaluating Static Analysis Alert Prioritization and Classification Techniques~\cite{HW08}} Benchmarks of our FaultBench benchmarks suite are in the form of complete Java program distributions (like jbook 1.4, jdom 1.1, xmlwriter 2.2.2, itrust fall 2007) including known bugs. FaultBench evaluates static analysis false pasitive mitigation techniques. The goal is to find best false positive mitigation technique with best rate of anomaly detection.

\subsubsection{Using a Diagnostic Corpus of C Programs to Evaluate Buffer Overflow Detection by Static Analysis Tools~\cite{KRA05}} The corpus consists of 291 small C programs generated in four versions: with no buffer overflow, and with minimum, medium, and large buffer overflows. This corpus is dedicated to check buffer overflow errors.

\subsubsection{Test-driving static analysis tools in search of C code vulnerabilities~\cite{CHK11}} The benchmarking test suite includes more than 700 programs, where each of them consists from 10 to 100 lines of code. All programs have a line with the tested flaw commented as /*ERROR*/ and another line commented as /*SAFE*/ that checks the tool capability to avoid reporting spurious errors. Error types used in the benchmarks have their links (IDs) to CWE.

\subsubsection{ABM: A Prototype for Benchmarking Source Code Analyzers~\cite{NCH05}} The benchmark test suite is composed of 91 C test cases. Whenever possible, test cases were constructed in matched pairs of OK and BAD test cases. Each BAD
test case has code with a vulnerability in it. Corresponding OK cases are similar to BAD cases but do not share the vulnerability. OK cases are generally constructed by patching. Each test case contains annotations specifying where a bug may be present. Test cases with vulnerabilities are tagged with a comment “/* BAD */” at any line that may be considered to contribute to the vulnerability described by the test case’s keywords. Those cases without vulnerabilities are tagged with a comment “/* OK */” at any line that could have contributed to the vulnerability if it had been present. Each test case is also annotated with information about valid and invalid inputs.

\subsubsection{CVE and CWE~\cite{cwe}} These are just errors taxonomies.

\subsubsection{NIST SAMATE Reference Dataset~\cite{SRD}} It is a suite of synthetic benchmarks  mainly for C, C++, and Java, that have been contributed by various groups. Ignore this...probably.

\subsubsection{SPEC} Ignore this...probably.

\subsubsection{TCP} Ignore this...probably.


\bibliographystyle{plain}
\bibliography{bd}

\end{document}

%{{{ Intro

\section{Introduction}
\label{sec:introduction}

databaze klasifikovanych chyb pro ladeni toolu urcenych pro analyzu realneho kodu.
availability, related work, url,

\xnote{
Nasledujici ctyri body predstavuji vecny obsah teto sekce. Dulezite je i jejich poradi. (nejen v techto bodech pouzivam tyto zkratky: Sorted Bug-reports Database (SBD), Program Analysis Tool (PAT)).
\\
(1) Problem pri vyvoji PATu, ktery zde hodlame resit:
%
Zamerujeme se na takove PATy, ktere mohou pri sve cinnosti produkovat falesna hlaseni. Uvest priklady takovych PATu (BLAST, SLAM, SDV, COVERITY, KLOCWORK, XGCC, CPPCHECK, PREFIX, PREFAST, ...).
%
Jde nam o pozdejsi stadia vyvoje techto PATu, kdy se ma nastroj zacit vyhodnocovat na realnych programech.
%
PAT pusteny na realnem programu casto generuje velke mnozstvi bug-reportu, o kterych predem nevime, ktere reporty predstavuji realne chyby, a ktere jsou jen falesna hlaseni. Zduraznit, ze overovani chyby/falesneho hlaseni v realnem, cizim a neznamem kodu je casto casove narocna a nezaziva prace.
%
Cilem SBD je tuto casove narocnou a nezazivnou cinnont znacne redukovat.
\\
(2) Jak konkretne SBD pomuze pri trideni bug-reportu:
%
SBD obsahuje pro analyzovany program velke mnozstvi jiz drive roztridenych bug-reportu. Vyvojari PATu tak staci, aby porovnal vystup z jeho toolu s obsahem SBD.
%
Krome vlastniho roztrideni (velke) casti vsech bug-reportu PATu na realne chyby a false positiva, vyvojar dale ziska prehled o znamych realnych chybach, ktere jeho PAT v analyzovane casti programu prehledl.
\\
(3) Prirozene zkvalitnovani SBD:
%
Muze se stat, ze vyvijeny PAT vyprodukuje na danem programu bug-report, ktery doposud neni v SBD. Vyvojar muze tento bug-report vlozit do SBD a muze k nemu pripojit informaci o tom, jestli jde o realnou chybu nebo o falesne hlaseni. Timto prispivanim do SBD si mohou vsichni vyvojari PATu vzajemne setrit cas.
%
SBD je navic navrzena tak, ze se do ni daji vkladat ruzne programy a dokonce i ruzne jejich verze. SBD tak mohou vyuzit i PATy specializovane na urcite typy programu (napr. na knihovny pro praci s retezci, sobory, GUI, apod.).
\\
(4) Statisticke udaje o cinnosti PATu na danem programu pomoci SBD: Uzivatel by mel mit moznost si v SBD oznacit moduly z daneho programu (verze), ktere jeho PAT analyzoval. Pak by mel mit moznost oznacit v SBD, ktere bug-reporty jeho PAT vyprodukoval. SBD by mu pak mela poskytnout statistiku, kolik nasel realnych chyb toho ktereho typu, kolik mel falesnych hlaseni toho ktereho typu, procentualni zastoupeni falesnych hlaseni, apod. 
}

There are several successful tools for finding errors in real code like
\textsc{Coverity}~\cite{coverity}, or \stanse~\cite{stanse}.
It is quite common that these tools generate, besides the errors which can
really occur during the execution, also false positives. Mostly, implementers
of such tools have to sort these reports manually to have an overview of how
the tool stands amongst others. To alleviate this burden, we augmented this
idea. We fixed a project and version and created a database of marked errors
where everybody may come and compare their tool. This will immediately return
count of errors the tool found along with false positives count. But more
importantly, it will tell the implementers also the count of errors that were
completely missed by their tool.

There were attempts to create a similar database previously like
in~\cite{IBM_database}. However those were either built upon a synthetic code
base, tailor-made exactly for that purpose, or they did not meet with success
because they have not got enough errors in the data base.

We built our database on the Linux Kernel. The idea behind is to have a real,
large-scale code base which might be also easily obtained. The version we
chose is a couple of years old version 2.6.28. The elderly version has several
reasons. First, we do not want to expose security holes of some new kernel to
wide public. Second, more importantly, we have an input from the kernel
developers on this particular version for all reports of chosen tools, so that
we are sure which errors are real and which are not.

In the end the database comprises 474 reports in 11 categories. The most
investigated categories from our side are locking errors like \emph{double
lock}, \emph{double unlock} or \emph{leaving functions in locked state}.
These are from the output of \stanse, \sparse\note{fakt?}, and
one tool from a North-American company dealing with static analysis of the
Linux kernel, we call this \undetI.

%}}}
%{{{ Marek

\section{\xnote{Sorted Bug-reports Database}}

\subsection{\xnote{Usage}}

\xnote{
V teto subsekci predvedeme, jak se konkretne s SBD pracuje. Jako PAT zvolime STANSE a budeme tridit jeho vystup (bug-reporty) pomoci SBD. Jednotlive ukazky budou zcela konkretni realne priklady pouziti SBD, ktere si kdokoliv muze sam hned odzkouset. Predvedeme nasledujici veci:
\\
(1) Trideni bug-reportu z PATu pomoci SBD: (a) Pomoci SQL dotazu: Vezmu nejakou vhodnou hlasku ze STANSE a ukazu SQL dotaz do SBD, ktery vrati, zda se jedna o chybu nebo falesne hlaseni, nebo ze hlaska zatim neni v SBD. Klidne lze udelat 3 priklady pro jednotlive mozne vysledky. Vhodne by mohly byt obrazky vystupu z konzole apod. (b) Pomoci weboveho rozhrani: Provedu totez co v (a), akorat zde zobrazim prislusne nastavena GUI okenka z weboveho rozhrani.
\\
(2) Pridani noveho bug-reportu do SBD: (a) Pomoci SQL dotazu: Vezmu nejakou vhodnou hlasku ze STANSE (nejlepe tu z tech 3 hlasek z bodu (1)) a ukazu SQL dotaz do SBD, ktery zajisti vlozeni bug-reportu do SBD. Mely by se odiskutovat jednotlive varianty dotazu (napr. vlozeni rovnou s klasifikaci bug/false alarm; dale typ chyby; pridani dalsich doplnujicich dat k zaznamu). (b) Pomoci weboveho rozhrani: Provedu totez co v (a), akorat zde zobrazim prislusne nastavena GUI okenka z weboveho rozhrani.
\\
(3) Pridani noveho programu, ci verze do SBD: (a) Pomoci SQL dotazu: Zvolim nejaky (mensi) open source a program ci knihovnu (pak se domluvime, co by to tak asi mohlo byt) a jeho/jeji nejakou verzi a formuluji SQL dotaz, ktery vlozi ten program do SBD. (b) Pomoci weboveho rozhrani: Provedu totez co v (a), akorat zde zobrazim prislusne nastavena GUI okenka z weboveho rozhrani. Mimochodem, tento bod (3) se da cely take zatim vyresit tak, ze rekneme, ze tato funkcionalita je prozatim ve vyvoji, ale bude zpristupnema v blizke budoucnosti.
\\
(4) Statisticke udaje o cinnosti PATu na danem programu pomoci SBD: (a) Pomoci SQL dotazu: Vzal bych vystup STANSE na urcitych par modulech. Pro tento vystup bych formuloval ruzne varianty SQL dotazu do SBD pro ziskani ruznych statistickych udaju o analyze. Mel by se take odiskutovat ten univerzalni format zadavani bug-reportu v dotazu a format seznamu analyzovanych modulu. (b) Pomoci weboveho rozhrani: Provedu totez co v (a), akorat zde zobrazim prislusne nastavena GUI okenka z weboveho rozhrani.
}

%}}}
%{{{ Results

%\section{Results}
\subsection{\xnote{Current State}}

\xnote{
Zde bych dal vsechny zajimave statisticke udaje o aktualnim stavu obsahu SBD. Nezapomenout uvest verzi kernelu, kterou tam mame. Take odiskutovat propojeni s tou databazi nejcastejsich chyb v SW (nemuzu si vzpomenout na tu zkratku) a diskuse zvolenych typu chyb v SBD. Sekci bych uzavrel nasimi plany s SBD do blizke budoucnosti (neco jako kratka Future Work).
}


\stanse, by its nature of overapproximating, reported all the reports.
\undetI failed to report 46\todo{exact count needed}. We
believe this is because of the default settings of the tool or due to filters
which are executed on the analysis results.

\subsection{\xnote{Availability}}

\xnote{
Krome vlastniho URL bych zde zminil moznost stazeni SBD (s licencnimi podminkami, pripadne) pro praci off-line. Pro tuto moznost stazeni SBD je ale nutne dodat na webove stranky instalacni manual. Zde muzeme navic pripadne dat odkaz na tuto stranku. Dale musime zminit, ze subsekce Usage obsahuje konkretni priklady pouziti, ktere je mozne na dane URL primo odzkouset. A ze z te sekce by take melo byt zcela jasne, jak pri zkouseni postupovat (tj. sekci lze tez chapat jako manual pro pouziti tech prikladu).
}
\snote{Ackoliv se mohou nektere ze zde pozadovanych veci zdat zbytecne, jedna se o explicitne zminene pozadavky pro tool-papers na ATVA2012}

\begin{table}[tb]
\begin{center}
\begin{tabular}{|l|c|c|c|c|c|c|c|c|} \hline

\textbf{Tool} & \textbf{Reports} & \textbf{Bugs} & \textbf{FP} &
	\textbf{Missed} \\ \hline
\sparse & & & & \\ \hline
\stanse & & & & \\ \hline
\undetI & & & & \\ \hline

\end{tabular}
\end{center}
\caption{Bugs in the database.}
\label{tab:results}
\end{table}


%}}}
%{{{ Related Work

\section{\xnote{Related Work}}

\xnote{
Musime dohledat clanky, ktere resi ruzne benchmarky a kolekce chyb apod. a radne se vuci nim vymezit. Hlavne je potreba zduraznit (oproti ostatnim) ze SBD neslouzi a ani nema slouzit pro porovnavani PATu (jednak vyvojari nemusi nektere vlozene bugy a falesna hlaseni znackovat svym PATem, dale nejsou uvadeny casy pro nalezeni dane chyby, neni zadavan HW na nemz PAT bezel, a take se neuvadi konfigurace, pri ktere ten ktery PAT chybu nasel). Proto statisticke udaje o obsahu SBD jsou prilis hrube na rozumne srovnani vykonu PATu prispivajicich do SBD. Z techto duvodu SBD tyto srovnavaci udaje vubec neposkytuje.
}

%}}}
%{{{ Conclusion

\section{\xnote{Conclusion}}
\label{sec:conclusion}

\paragraph{Acknowledgements}
\label{sec:acknowledgements}

Jan Strej\v{c}ek has been supported by The Czech Science Foundation
(GA\v{C}R), grants No.~P202/10/1469 and P202/12/G061. Ji\v{r}\'i Slab\'{y}
and Marek Trt\'{\i}k have been supported by The Czech Science Foundation
(GA\v{C}R), grant No.~102/09/H042.

%}}}

\bibliographystyle{plain}
\bibliography{bd}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% TeX-command-default: "pdfLaTeX"
%%% End: 
