% vim: textwidth=78 ai

\documentclass[draft,11pt]{fithesis}

\usepackage[utf8x]{inputenc}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{graphicx}

\thesislang{en}

\thesistitle{Rapid data transfers on COMBO platform}
\thesissubtitle{Master's thesis}
\thesisstudent{Jiří Slabý}
\thesisadvisor{Ing. Pavel Čeleda, Ph.D.}

\thesiswoman{false}
\thesisfaculty{fi}
\thesisyear{spring 2008}

%\hyphenation{flow-mon}

\newcommand{\lbr}{\emph{Liberouter}}
\newcommand{\com}{\emph{COMBO6}}
\newcommand{\comx}{\emph{COMBO6x}}
\newcommand{\nflow}{\emph{NetFlow}}
\newcommand{\fpga}{\emph{FPGA}}
\newcommand{\dma}{\emph{DMA}}

\begin{document}
\FrontMatter
\ThesisTitlePage
\begin{ThesisDeclaration}
  \DeclarationText

  \begin{flushright}
  In Brno, \today \\
  Jiří Slabý
  \end{flushright}
  \AdvisorName
\end{ThesisDeclaration}
\begin{ThesisThanks}
  I want to thank my thesis advisor Pavel čeleda. And least but not the last
  I would like to thank here my parents for their patience too.
\end{ThesisThanks}
\begin{ThesisAbstract}
  In this thesis we are concerned in fast transfers from hardware to computer
  memory and vice versa. Above all we specialize to scatter-gather direct
  memory access transfers from/to hardware through ring buffers. Also we
  discuss other transfer methods implemented in other projects and consider
  using them here. Finally we implement the method and test it with all
  available tools and describe the testing process.
  
\end{ThesisAbstract}
\begin{ThesisKeyWords}
  Linux, Driver, ring buffer, mmap, DMA, COMBO6x, FlowMon, Liberouter
\end{ThesisKeyWords}

\MainMatter
\tableofcontents

\chapter{Introduction}
  To even boot a computer, it's necessary to process many transfers inside the
  hardware among its components. It's a basic function which must work,
  however there are several methods of how it can be achieved. Especially
  the fast ones are very important in computer science.

% Many improvements have been done in this area in the past.

  \section{Linux}
  1992, Torvalds, fast development nowadays, well evolved API, licensing.

  \section{Liberouter project}
  Also in Liberouter project a new generation of the \comx\ card has evolved

  \section{NetCOPE project}
  why, how, szedata2

  \section{Devices interconnection}
  device -- transparent bus -- pci bus -- northbridge -- southbridge -- cpu

  \section{Bus Mastering and DMA}
  dma controllers, position in interconnection, purpose, how do they work\ldots

  \section{Data from kernelspace to userspace}
  mmio in short, pcap in short

% http://en.wikipedia.org/wiki/Bus\_mastering
% http://www.quatech.com/support/comm-over-pci.php

  \section{udev}
  userspace dev of whatever it stands for\ldots

  how does it work, what is it for, very fast comparison with the past devfs

  The way, how is it used in the \lbr is described in next chapter.

\chapter{Device nodes naming}
  2 kind of nodes -- combo x szedata, consistent naming needs

  udev\ldots

  The script (figure \ref{fig.udevscript}) and the rules (figure
  \ref{fig.udevrules}) file contents come here.

  Also the description of each entry in the lines.

  \begin{figure}
  \begin{verbatim}
KERNEL=="szedata*", ...
  \end{verbatim}
  \caption{.rules}
  \label{fig.udevrules}
  \end{figure}

  \begin{figure}
  \begin{verbatim}
#!/bin/bash
  \end{verbatim}
  \caption{script}
  \label{fig.udevscript}
  \end{figure}

\chapter{szedata2}
  What is it good for? Why is it arising? Diff betwwen szedata and version 2.

  \section{Introduction}
  \subsection{Ring buffer processing}
  Organistaion of data in the card memory.

  \subsection{Data path}
  card $\rightarrow$ host memory (kernelspace) $\rightarrow$ userspace (pcap,
  mmap)

  \subsubsection{Modified libpcap}
  ref Konik's work

% --
  \section{Development cycle}
  \subsection{Models and design}
  Bouml et al.

  \subsection{Code templates}
  + design patterns if any applies

  \subsection{Code development}
  combo6x driver as a base, new szedata2 module
    \subsubsection{Mmio space structure}
    locked : desc : rx ring : tx ring

  \subsection{Backporting to older kernels}
  Why we should do this? Macros used (irq handler vs. 2.6.19, mmap vs.
  remap\_pfn\_range, remap\_page\_range), workarounds (class+class\_simple
  vs. 2.6.16).

  \subsection{Final code}
  Nice, working, debugged kernel module as a result. License (\cite{gplv2}).

\chapter{Conclusion}
  XXX\cite{ldd}, \cite{ulk}

\bibliographystyle{plain} 
\bibliography{dp}

\include{app}

\end{document}
