% vim: tw=78 ai spell spelllang=en

\documentclass[10pt]{llncs}
\usepackage{color}
\usepackage{url}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{xspace}
\pagestyle{plain}

\lstset{basicstyle=\ttfamily\scriptsize,escapeinside={(*@}{@*)}}

%{{{ (La)TeX definitions

\newcommand{\nb}[1]{\marginpar{\scriptsize #1}}
\renewcommand\note[1]{\nb{\textcolor{blue}{#1}}}
\newcommand\todo[1]{\nb{\textcolor{red}{#1}}}
\newcommand\tool{\textsc{Symbiotic}\xspace}
\newcommand\klee{\textsc{Klee}\xspace}

%\renewcommand\note[1]{}
%\renewcommand\todo[1]{}

%}}}

\begin{document}

%{{{ Title + Author + Abstract

\title{\tool~2: More Precise Slicing}
% \thanks{This work has been supported by The Czech Science
%     Foundation (GA\v{C}R), grant No.~P202/12/G061.}}
\subtitle{(Competition Contribution)}
\author{Jiri~Slaby \and Jan~Strej\v{c}ek}
\institute{Faculty of Informatics, Masaryk University\\
  Botanick\'a 68a, 60200 Brno, Czech Republic\\
  \email{\{slaby,strejcek\}@fi.muni.cz}}

\maketitle

\begin{abstract}
  \tool~2 keeps the concept and the structure of the original bug-finding
  tool \tool, but it uses a more precise slicing based on a field-sensitive
  pointer analysis instead of field-insensitive analysis of the original
  tool. The paper discusses this improvement and its consequences.  We also
  briefly recall basic principles of the tool, its strong and weak points,
  installation, and running instructions. Finally, we comment the results
  achieved by \tool 2 in the competition.
\end{abstract}

%}}}

%{{{ Verification approach

\section{Verification Approach and Software Architecture}
\label{sec:approach}

Both \tool~\cite{SST13} and \tool 2 implement our verification approach
proposed in~\cite{sse}. The approach combines three standard techniques,
namely \emph{code instrumentation}, \emph{program slicing}~\cite{weiser},
and \emph{symbolic execution}~\cite{king}. While the approach is originally
designed for detection of bugs described by state-machines, \tool 2 still
supports only one kind of bugs: reachability of an \texttt{ERROR}
label. Hence, we briefly recall the approach restricted just to this simple
kind of errors. We explain the structure of the tool simultaneously.
\begin{enumerate}
\item Code instrumentation inserts \texttt{assert(0)} to each \texttt{ERROR}
  label. It is realized by a \texttt{bash} script calling \texttt{sed}. The
  instrumented code is translated to the \textsc{Llvm} bitcode by
  the \textsc{Clang} compiler.
\item Program slicing removes instructions of the instrumented code that do
  not affect reachability of the inserted \texttt{assert(0)}
  statements. This code size reduction is crucial for the overall efficiency
  of \tool 2. The slicer is implemented in C++ as a plug-in for the
  \textsc{Llvm} optimizer \texttt{opt}.
\item Symbolic execution either reaches \texttt{assert(0)}, or correctly
  finishes the execution without reaching \texttt{assert(0)}, or it runs out
  of time or memory etc. These possibilites correspond to answers
  \texttt{FALSE}, \texttt{TRUE}, \texttt{UNKNOWN}, respectively.
  We use symbolic executor \klee~\cite{klee} whose outputs are translated
  to \texttt{TRUE/FALSE/UNKNOWN} by a simple \texttt{bash} script.
\end{enumerate}
The whole pipeline is executed stepwise by another \texttt{bash} script.

All improvements of \tool 2 over the tool \tool competing in SV-COMP 2013
are in the slicer.  We have fixed some bugs in the original slicer (invalid
treatment of several instructions and functions with variable number of
arguments). While the original fixpoint algorithm for slicing~\cite{weiser}
is relatively simple, it gets more complicated for programs with pointers as
one instruction can influence the following one without any syntactic
overlap. We need to use a pointer analysis to know which pointers can point
to the same target. Both \tool and \tool 2 use Andresen's pointer
analysis~\cite{And94}, \tool 2 replaces the original field-insensitive
analysis by a field-sensitive one. This means that every field of a
\texttt{struct} is now handled as a different pointer target. Similarly, we
handle the first 64 elements of each array as distinct targets. The
field-sensitive analysis is computationally more demanding. Thus we have
added also some type-filters that speed up the pointer analysis and make its
results more accurate. All these improvements of the slicer significantly
reduce the number of incorrect answers produced by \tool 2. For example,
\tool 2 produces only correct answers for category \emph{ProductLines} while
\tool produces 131 incorrect answers on the same category.
 
%}}} 
%{{{ Software architecture - Zakomentováno

% \section{Software Architecture}
% \paragraph{Libraries/External Tools} For the code translation and for
% operations with the \textsc{Llvm} bitcode we use \textsc{Llvm}/\textsc{Clang}
% 3.2\footnote{\url{http://llvm.org}}.  The symbolic executor is
% \klee~\cite{klee} which itself uses the \textsc{Stp} constraint
% solver~\cite{stp}. We obtained both \klee and \textsc{Stp} from the
% respective GIT repositories and used the snapshots.

% \paragraph{Software Structure and Architecture} The architecture described
% in Section~\ref{sec:approach} is summarized in Figure~\ref{fig:pipeline}.
% The slicer is written as a plug-in for the optimizer \texttt{opt} from the
% \textsc{Llvm} suite. It is publicly available in a separate repository at
% \url{https://github.com/jirislaby/LLVMSlicer/}.
% % The slicer improves the symbolic execution considerably.
% For ease of use, all parts of the tool pipeline are one by one run by a
% single script \texttt{runme}.

% \begin{figure}[tb]
% \begin{center}
% \input{pipeline2.pspdftex}
% \end{center}
%   \caption{Pipeline of the tool}
% \label{fig:pipeline}
% \end{figure}

% \paragraph{Implementation Technology} The instrumentation is performed by a
% \texttt{bash} script using \texttt{sed}. The final filter also uses
% \texttt{bash} with the help of \texttt{grep}. The rest of the toolchain is
% written in C++ and compiled using \texttt{gcc}.

%}}}
%{{{ Discussion

\section{Strengths and Weaknesses}

Our tool is applicable to all competition benchmarks satisfying two
restrictions: the studied property is \texttt{ERROR} label reachability and
the benchmark code is a sequential C program. Hence, the results of \tool 2
in the competition categories \emph{Concurrency} and \emph{MemorySafety}
should be ignored (we missed the opt-out deadline). The first restriction
can be removed by implementation of a more sophisticated code
instrumentation.  The second restriction comes directly from the approach as
symbolic execution and program slicing are primarily designed for sequential
programs.

\medskip We first discuss strong and weak aspects of the approach and then
we talk about additional strong and weak aspects of the tool. Our approach
is based on symbolic execution which should produce only correct answers. On
the other hand, symbolic execution suffers from the path explosion problem
and relies on expensive (and often even undecidable) SMT solving. Hence,
application of symbolic execution leads to many \texttt{UNKNOWN} answers,
which is also the main weakness of the approach. To reduce this weakness, we
combine symbolic execution with slicing which is the only theoretical source
of incorrect answers (namely false positives) of our approach. Indeed,
slicing can in some cases remove an infinite loop and a potential
unreachable \texttt{ERROR} label located below the cycle thus become
reachable. Howerver, this sitation is very rare in practice (e.g.~it does
not appear in the competition benchmarks) and we do not see it as a problem.
An orthogonal method to reduce the high cost of symbolic execution is to use
some of its variants suppressing the path explosion problem. For example, we
plan to apply \emph{compact symbolic execution}~\cite{CSE} instead of the
classic one.

The strong aspect of the tool is its simple architecture: it is a sequence
of scripts and standalone tools that are easy to replace (for example, if
there is a better symbolic executor for \textsc{Llvm} bitcode, we can deploy
it in few minutes). The main weakness of \tool 2 are incorrect results it
produces. Even if the number of incorrect results is substantially lower
than in the case of \tool, it is still relatively high. All the incorrect
results are due to imperfection of our implementation.

%}}}
%{{{ Tool setup and configuration

\section{Tool Setup and Configuration}
Before using \tool 2, ensure that the target system contains 32-bit
libraries and \texttt{llvm} with \texttt{clang}. \texttt{llvm} and
\texttt{clang} have to be in version 3.2 exactly.  Then, \tool 2 can be
downloaded from %:
%\begin{center}
\url{http://sf.net/projects/symbiotic/}.
%\end{center}
As \klee requires to be run from a pre-defined path, we are obliged to
change the current directory to \texttt{/opt/} and \emph{untar} the
downloaded \tool 2 archive there.  Running of the tool is then
straightforward. When the current directory is \texttt{/opt/symbiotic/}, the
tool can be invoked for each \texttt{<benchmark.c>} from the set by
\texttt{./runme <benchmark.c>}. For benchmarks intended for 64-bit, set
\texttt{MFLAG=-m64} environment variable.  The answers provided by \tool 2
are as required by the competition rules: \texttt{TRUE/FALSE/UNKWNOWN}. If
the result for \texttt{<benchmark.c>} is \texttt{FALSE}, discovered error
paths can be found in \texttt{<benchmark.c>-klee-out/}.

%}}}
%{{{ Software project and contributors

\section{Software Project and Contributors}
\tool 2 was contributed prevalently by the authors of this paper and Marek
Trt\'{\i}k. Jiri Slaby is a contact person. The tool is licensed under the
GNU GPLv2 License.

% We would like to thank our advisors prof. Anton\'in Ku\v{c}era and dr. Jan
% Strej\v{c}ek. Our work is supported by grant no.

%}}}

\bibliographystyle{plain}
\bibliography{sse2014}

\end{document}
