% vim: textwidth=78 ai spell spelllang=en
\documentclass[11pt,a4paper]{article}

\usepackage[utf8x]{inputenc}
\usepackage[IL2]{fontenc}
\usepackage[final]{hyperref}
\usepackage{multirow}

\setlength{\textwidth}{140mm}
\setlength{\textheight}{220mm}
\setlength{\oddsidemargin}{12mm}
\setlength{\topmargin}{8mm}

\font\logo fi-logo at 4cm

\setcounter{tocdepth}{2}

\addtolength{\oddsidemargin}{0.2in}
\addtolength{\evensidemargin}{0in}
\addtolength{\textwidth}{0in}
\addtolength{\topmargin}{0in}
\addtolength{\textheight}{0in}

\newcommand{\stan}{\textsc{Stanse}}

\begin{document}
\begin{titlepage}
\vspace{2cm}
\begin{center}
      {\sc Masaryk University\\Faculty of Informatics}

     \vspace{1cm}
     \logo SL

     \vspace{2cm}
     {\bf\huge Symbolic execution on OS kernels}\\[0.8in]

     {\sc\large Dissertation Thesis Topic}\\[0.4in]

     {\Large\bf Jiří Slabý}
     \vfill
     {\large Brno, spring 2010}
     \vfill
\end{center}
     {\bf Supervisor:} prof. RNDr. Antonín Kučera, Ph.D.\\
     {\bf Supervisor-specialist:} Mgr. Jan Obdržálek, PhD.\\[0.5cm]
     {\bf Supervisor's Signature: ......................}
\end{titlepage}

\pagestyle{empty}

\tableofcontents

\pagestyle{plain}

\newpage
\setcounter{page}{1}

\section{Introduction}
Static analysis is currently considered to be a powerful approach used for
finding errors in a source code. The first real application was performed by
Engler et al. in \textsc{Metal}~\cite{metal} in 2000.

Static analysis might have a form of finding some patterns in the code in some
specified manner (similarly to \textsc{Metal}) or test cases generation which
try to cover the most of the code (covering both true and false branches of
each branching).

Many tools were developed based on either approach however they are commercial
with closed source code, limited in what they can do or both. We are currently
working under Institute of Theoretical Informatics at the Faculty of
Informatics on a free (open-source) tool called \stan. It aims at finding
errors with a use of static analysis.

\subsection{Symbolic Execution}
One method of static analysis, \emph{Symbolic execution} is currently widely
used for checking code properties. It stems from~\cite{select} and many
studies tried to apply the principle on a variety of userspace programs.
\textsc{Klee}~\cite{klee} primarily aims at tests generation that would cover
as many program instructions as possible. \textsc{Pex}~\cite{pex} uses
symbolic execution in cooperation with standard execution on processor
to address program coverage as well, but from another point of view.
Section \ref{current-se} is deserved for detailed description of the work done
in this area.

To the best of my knowledge nobody has ever tried to apply symbolic execution
on an operation system kernel (further only \emph{kernel}). This task
incorporates not only the execution
itself, but some kind of code preparation, compilation, linking and dynamic
linking. These terms will be discussed further in the text.

\subsection{Justification}
Symbolic execution has shown its power in finding errors in userspace
programs. The idea of applying the approach to the kernel is straightforward
extension, however it is not as easy task as it might seem. The kernel needs
to be processed, the result then executed and checked for some properties.
All of
\begin{itemize}
\item processing the kernel code,
\item finding the right properties to check and
\item finding them in the processed result
\end{itemize}
was not done by anybody yet. The dissertation should lead to complete
implementation of the mentioned and finding errors with minimum false
positives.

Dissertation Thesis Intent; Methods of Solving the Problem; Comparison of Key
Results

\newpage

\section{Report on Current Results}
Static analysis is a form of software testing and/or finding errors. It
incorporates many branches such as \emph{model checking} and \emph{abstract
interpretation}. Further in the text I will aim specifically at abstract
interpretation, because it is a field of my doctoral study.

\subsection{Abstract Interpretation}
Abstract interpretation is a relatively old approach, first summarized and
formalized by Cousot \& Cousot in~\cite{cousot} in 1977. There they specify a
formal model used in interpretation to (a) construct a correct abstraction of
the code; (b) be able to compute a solution of some properties defined on the
abstracted code (by fixpoints). Note that checking properties in programs is
in general undecidable, hence they also define narrowing and widening
techniques if the system of equations does not approach convergence when
computing a fixpoint.

Cousots laid down the foundation for others, however the principle was brought
to interest even after 2000. It was thanks to work of Engler et al. on a meta
compilation based technique described in~\cite{metacomp}. System developers
there specify properties of the system in \textsc{Metal}~\cite{metal} language,
compile it into internal representation and the tool applies the specification
on all executable paths in a particular function. Violation of the
specification on a path is then reported as error.

By improving the technique\footnote{According to the press release at
\href{http://www.coverity.com/html/press_story04_02_02_05.html}%
{http://www.coverity.com/html/press\_story04\_02\_02\_05.html}, it produces
less false positives, can run in parallel on the code and contains a
statistical engine.} they developed a commercial derivation of
\textsc{Metal/xgcc}
named \textsc{Coverity Prevent}~\cite{coverity}, according to developers, the
current world leader in static analysis. Currently it analyses many production
systems including the Linux kernel on a daily
basis\footnote{\url{http://marc.info/?l=linux-kernel&m=125856033627134&w=2}}.

Programming languages supported by the mentioned Engler's tools are C and C++.

Aside from Engler's work, Klocwork~\cite{klocwork} started to develop their
bug-finding static analysis tools in 2003. Currently they, as same as
Coverity, find errors in many production projects. In addition to C/C++,
Klocwork products are able to parse Java and C\# programs as well. They seem
to be world number two in code analysis.

\paragraph{Open-source}
All the mentioned tools till now are commercial products without an access to
their source code. However there are also many open-source tools available.
First, let me mention \textsc{Lint} derivatives. They belong to the
\emph{first static analysis tools} used whatsoever. \textsc{Lint} used to find
suspicious and non-portable code constructs. Mostly, the checks are currently
superseded by compilers which perform the checking during standard compilation
phase, so that \textsc{Lint} is not used anymore.

There was also a popular successor \textsc{LCLint}~\cite{lclint} later renamed
to \textsc{Splint}. The development stalls for more than five
years\footnote{\url{http://www.cs.virginia.edu/pipermail/%
splint-discuss/2007-July/000994.html}} now though. Again \textsc{LCLint}
performs many checks which are done by compiler nowadays. Besides that, it can
catch buffer overflow vulnerabilities. It also interprets \emph{annotated
code} and tries to avoid false positives that way (it allows turning on more
checks which would generate many false positives without the annotations).

Later, tools dedicated for a \emph{single purpose or project} emerged.
Specifically for the Linux kernel purposes, \textsc{Sparse}~\cite{sparse} was
developed.  For example it watches pointer usage on user- and kernel-space
interface (they must not be intermixed) or uses of remapped I/O memory
(have to be dereferenced via special helpers) and similar, kernel specific,
properties.  All the mentioned has no use other than in the kernel.
\textsc{Sparse} is widely used by kernel developers and many sparse errors are
fixed in each kernel release cycle\footnote{For instance, over 20 of sparse
errors were fixed (since 2.6.31) in 2.6.32.}. Let me note, that since this
tool is able to check whole kernel, it is thus capable to parse most of GNU~C
extensions described in \cite{gcc}.

From a different point of view, static checkers need not to be only
\emph{dedicated programs}. \textsc{clang} \cite{clang} compiler contains a
built-in non-configurable static analyzer. With a use of internal compiler
structures, it looks for typical programming errors like buffer overruns,
using of deprecated (insecure) functions, a sort of memory leaks etc.

All previous open-source tools are configurable only to some extent. There is
no way of specifying new check but changing the sources.
\textsc{Uno}~\cite{uno} is a \emph{fully configurable} tool. It allows user to
specify for what type of bugs should the checker look for. Of course there are
predefined ones (e.g. use of uninitialized variables), but they can be easily
augmented.

Still, some tools are hard to use. There is one exception called
\textsc{FindBugs}~\cite{findbugs}. To the best of my knowledge it is the most
known code-checking tool for Java sources. \textsc{FindBugs} is capable to
find hundreds types\footnote{\url{http://www.ibm.com/developerworks/%
java/library/j-findbug1/}} of errors in whole Java project. It has a
\emph{graphical user interface} (GUI) and in fact it is a single button tool.

\paragraph{Stanse} I slightly described \stan\ in the introduction. It is
developed at ITI by J. Obdržálek, J. Slabý and M. Trtík (alphabetical order,
I will describe my role later) with a help of master students. For the ease of
programming, the tool is written mostly in Java. Some parts are pure C and the
rest (scripts) uses \textsc{Perl}.

\stan\ tries to pick all the pros from previously named tools and also add
some novelty. It consists of three checkers.
\begin{itemize}
\item AutomatonChecker -- technique based on finite automata. I will describe
the technique in detail in Chapter~\ref{automata}.
\item ThreadChecker -- Jan Kučera developed the checker in the scope of his
master thesis. The checker looks for thread-creation functions and statically
follows the threads. Namely, it watches lock operations and builds
\emph{locksets}~\cite{lockset}, represented as resource allocation graphs
(RAG). The checker combines RAGs of all threads and tries to find cycles
there. If it is successful with the search, an error is reported\footnote{For
example one is reported at \url{http://lkml.org/lkml/2009/4/14/527}.}.
\item ReachabilityChecker -- It is a short checker (200 lines of code) which
performs reachability analysis of statements to reveal unintentional dead
code. It catches errors like omitted semicolons after the if branch, so that
the branching is not done.
\end{itemize}

I did not mention what type of code walking all the tools support. \stan\
performs simple inter-procedural analysis. It binds together a CFG node of a
call and CFG of called function.

\begin{figure}[htb]
\begin{center}
\begin{tabular}{l|ccccl}
\multirow{2}{*}{Tool name} &
\multirow{2}{*}{Open} &
Confi- &
Batch &
Inter- &
\multirow{2}{*}{Languages} \\
%Tool name        & Open & Confi-       & Batch   & Inter-     & Languages
		  &      & gurable      & support & procedural & \\ \hline
\textsc{Coverity} & no   & yes          & yes     & yes        & C, C++, C\#, Java \\
\textsc{Klocwork} & no   & yes          & yes     & yes        & C, C++, C\#, Java \\
\textsc{Splint}	  & yes  & no           & no      & no         & C \\
\textsc{Sparse}   & yes  & no           & yes     & no         & C \\
\textsc{clang}	  & yes  & no           & no      & no         & C, C++, Obj-C \\
\textsc{Uno}	  & yes  & yes          & no      & yes        & C \\
\textsc{Stanse}	  & yes  & yes          & yes     & yes        & C \\
\textsc{FindBugs} & yes  & no           & yes     & no         & Java \\
\end{tabular}
\caption{Overview and comparison of described tools. \emph{Configurable} here
means, that new checks may be added to the tool without a need of source code
update.}
\end{center}
\end{figure}

\subsection{Algorithms over Abstraction}
The previous chapter introduced solely to the tools currently known and used.
This chapter will cover the main algorithms implemented over abstract
interpretation in some of them.

\subsubsection{Automata}
\label{automata}


\subsubsection{Symbolic Execution}
\label{current-se}
So far I have not mentioned one special code interpretation called symbolic
execution.
\\[1cm] \noindent
\textsc{Klee}~\cite{klee}, \textsc{Pex}~\cite{pex}

\newpage

\section{Dissertation Thesis Intent}
As mentioned in the introduction, the intention is to apply symbolic execution
on a kernel. Dissertation thesis should contain all the steps needed to
perform the execution to find errors in the code. From processing the code
(compilation) towards recognizing error patterns specified by user and
reporting them. Now I will go into details of each phase and describe the idea
behind.

\subsection{Code processing}
For the convenience of users, we want to support makefile build system to
obtain source code file list. Method used in \stan\ may be reused for this
purpose to some extent. It originates from the way how
\textsc{Sparse}~\cite{sparse} tool was designed. In fact it is simple
interception of a compiler command-line invoked by the build system. It then
runs original, real, compiler with proper arguments.

Although in \stan\ we only store the intercepted command-line into a single
file, this will not suffice for out new symbolic 

Compilation gnome: \textsc{LLVM}~\cite{llvm}

Compilation problems: inlined assembly, 

Linking: \texttt{llvm-link}

Linking against unknown functions vs. userspace: \textsc{Klee}'s~\cite{klee}
solution; functions written in assembler

\subsection{Specifying Error Patterns}
Instrumented code: asserts, \texttt{\_\_chk\_user\_ptr}

\subsection{Finding Error Patterns}
Marek's dissertation, fallback to \textsc{Klee} (we have source code).

Expected Results; Schedule for Further Development; Expected Outcomes

\newpage

\bibliographystyle{plain}

\bibliography{teze}
\addcontentsline{toc}{section}{References}

\newpage

\section{Summary of Study Results}
During the first year of doctoral study I studied papers with results of past
researches in the area of static analysis. Also I read through some literature
about static analysis and compilers.

Under \emph{ITI}, we developed \stan, a static analysis tool written in Java.

\end{document}
