% vim: textwidth=78 ai spell spelllang=en sw=2
\documentclass[11pt,a4paper]{book}

%\includeonly{}

\usepackage{adjustbox}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bibentry}
\usepackage[table]{xcolor}
\usepackage[utf8x]{inputenc}
\usepackage[IL2]{fontenc}
\usepackage[final]{hyperref}
\usepackage[all]{hypcap}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{xspace}
\usepackage[final]{listings}
\usepackage{tikz}
\usetikzlibrary{arrows,fit,shapes,positioning,plothandlers}

\setlength{\marginparwidth}{0pt}
% part of marginparwidth which is 106pt
\addtolength{\oddsidemargin}{40pt}
\addtolength{\evensidemargin}{-40pt}

\tikzset{cfg/.style={thick,draw,minimum size=6mm}}
\tikzset{cfg_trian/.style={cfg,inner sep=0pt,regular polygon,regular polygon
        sides=3}}
\tikzset{cfg_start/.style={cfg_trian, regular polygon rotate=180}}
\tikzset{cfg_end/.style={cfg_trian, regular polygon rotate=0}}
\tikzset{cfg_edge/.style={<-, >=stealth', semithick}}

\tikzset{ast/.style={thick,draw,minimum size=5mm}}
\tikzset{ast_hl/.style={ast, bottom color=black!10, top color=black!10}}

\tikzset{tl_state/.style={rounded corners, thick, draw, minimum height=6mm,
	minimum width=12mm, bottom color=black!15, top color=black!5}}
\tikzset{tl_phase/.style={thick, draw, minimum size=8mm}}
\tikzset{tl_edge/.style={<-, >=stealth', semithick}}

\tikzset{pts_edge/.style={<-, >=stealth', semithick}}
\tikzset{pts_node/.style={circle, draw, thick, minimum size=7mm, inner
	sep=0pt}}

\hyphenation{Mi-chael}
\hyphenation{Sa-man}

\nobibliography*

%\setlength{\textwidth}{140mm}
%\setlength{\textheight}{220mm}
%\setlength{\oddsidemargin}{15.5pt}
%\setlength{\evensidemargin}{15.5pt}

\font\logo fi-logo at 4cm

\setcounter{tocdepth}{2}

\lstset{language=C,tabsize=2,numbers=left,showstringspaces=false,
	breaklines=true,breakatwhitespace=true,escapeinside={(*@}{@*)},
	frame=l,captionpos=b,basicstyle=\sffamily\small}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\nb}[1]{\marginpar{\scriptsize #1}}
%\newcommand{\todo}[1]{\nb{\red{#1}}}
\newcommand{\todo}[1]{}
\newcommand{\var}[1]{\ensuremath{\mathsf{#1}}}
\newcommand{\sym}[1]{\ensuremath{\underline{\var{#1}}}}
\newcommand{\true}{\ensuremath{\mathit{true}}}
\newcommand{\code}[2][]{\lstinline[columns=fullflexible, #1]$#2$}

\newcommand{\cdb}{\textsc{ClabureDB}\xspace}
\newcommand{\clang}{\textsc{Clang}\xspace}
\newcommand{\cppcheck}{\textsc{Cppcheck}\xspace}
\newcommand{\fbugs}{\textsc{FindBugs}\xspace}
\newcommand{\gcc}{\textsc{gcc}\xspace}
\newcommand{\klee}{\textsc{Klee}\xspace}
\newcommand{\ucklee}{\textsc{UC-Klee}\xspace}
\newcommand{\llvm}{\textsc{Llvm}\xspace}
\newcommand{\sage}{\textsc{Sage}\xspace}
\newcommand{\smatch}{\textsc{Smatch}\xspace}
\newcommand{\sparse}{\textsc{Sparse}\xspace}
\newcommand{\stan}{\textsc{Stanse}\xspace}
\newcommand{\symbio}{\textsc{Symbiotic}\xspace}
\newcommand{\perl}{\textsc{Perl}\xspace}
\newcommand{\xgcc}{\textsc{xgcc}\xspace}

\newtheorem{dfn}{Definition}

% stanse article
\newcommand{\TC}{\textsc{ThreadChecker}\xspace}
\newcommand{\AC}{\textsc{AutomatonChecker}\xspace}
\newcommand{\RC}{\textsc{ReachabilityChecker}\xspace}
\newcommand{\LC}{\textsc{LockChecker}\xspace}

% fmics article
\newcommand{\SA}{{\textit{SA}}}
\newcommand{\twolines}[2]{
  \noindent
  \begin{tabular}{c}
     \,\,#2\\[-1mm]
     \scriptsize #1
  \end{tabular}
}

\begin{document}
\frontmatter
\pagestyle{plain}

\begin{titlepage}
\vspace{2cm}
\begin{center}
     \textsc{Masaryk University\\Faculty of Informatics}

     \vspace{1cm}
     \logo SL

     \vspace{2cm}
     \huge{\textbf{Automatic Bug-finding Techniques\\[0.5em]
     for Large Software Projects}}
     \\[1.7cm]

     \large{\textsc{Dissertation Thesis}}
     \vfill
     \Large{\textbf{Jiri Slaby}}
     \vfill
     \large{Brno, 2013}\\[2mm]
     \small{Revision \input{revision.txt}}
     \\[1.3cm]
\end{center}
  \begin{tabular}{ll}
  \textbf{Supervisor:} & prof. RNDr. Antonín Kučera, PhD.\\
  \textbf{Co-supervisor:} & assoc. prof. RNDr. Jan Strejček, PhD.
  \end{tabular}
\end{titlepage}

\clearpage
~
\newpage

\begin{center}
~
\vfill
Copyright \copyright\ Jiri Slaby, 2013 \\
All rights reserved\scalebox{.2}{to\v{z} to ja}
\end{center}

\chapter*{Abstract}
This thesis studies the application of contemporary automatic bug-finding
techniques to large software projects. As they tend to be complex systems,
there is a call for efficient analyzers that can exercise such projects
without generating too many false positives. To fulfill these requirements,
this thesis aims at the static analysis with the primary focus at techniques
like the abstract interpretation and symbolic execution.  Since their
introduction in 1970's, their applicability is increasing thanks to current
fast processors.  Only this renders the techniques usable for larger software.

The thesis summarizes the current state of the art of chosen techniques
implemented in the current static analyzers. Further, we discuss some large
code bases which can be used as a test-bed for the techniques. Finally, we
choose the Linux kernel. On the top of the explored state of the art, we
develop our own techniques. We present the framework \stan which utilizes the
abstract interpretation and has no issues with handling the Linux kernel. The
framework also prunes false positives resulting from some checkers. The
pruning method is based on the pattern matching, but it turns out to be weak.
Hence we subsequently try to improve the situation by the symbolic execution.
For that purpose we use an executor called \klee in our new tool \symbio. But
since the symbolic execution requires a long computation time, we slice the
code by a static slicer first.

During our experiments, we cumulated knowledge about categories of bugs
usually present in code such as the Linux kernel. We investigated it more and
developed a database of known bugs in a fixed version of the kernel. It is
publicly available and contains information about errors and false positives
found by several bug-finding tools. This can help developers of new static
analyzers to compare their brand new tool against the database and see how
their tool stands with respect to false positives, missed errors and of
course error hits.

We also participated with \symbio in the Software Verification Competition.
The tool had to be adapted to the competition needs.  Since we participated
only with a too preliminary version of the tool, the results were not so
encouraging.  Nonetheless, the tool was improved later and it gives
good results now -- we collect nearly all points in some competition
categories.  We plan to compete in the competition with our tuned tool again.

\chapter*{Acknowledgement}
I would like to thank my supervisors prof.\ Kučera and assoc.\ prof.\
Strejček for their patience with my ignorance during the nice past years spent
at the university. I really enjoyed that time and was taught many new things
which would stay unrevealed for me otherwise. I thank my colleagues for their
ideas and inspiration, especially Marek with whom we started the studies and
walked most of the path.

Of course, my parents and family altogether deserve many thanks for their
eternal support. Without them and their care, I would be only a poor man on
this Earth. And last, but definitely not least, Zuzanka must not be missing on
this list. Her love, patience, and care is much appreciated. I am really glad
we met each other.

\clearpage

\tableofcontents

\clearpage

\pagestyle{headings}
\mainmatter

\include{intro}
\include{state}
\include{stanse}
\include{fmics}
\include{clabure}
\include{svcomp}

\clearpage
\chapter{Conclusion}
In this thesis, we introduce the reader to the checking of large software
projects. We emphasized the importance of evading as many errors from such
projects as possible. Otherwise, with a presence of errors, attackers have an
open door to gain sensitive data for example. As we aim specifically at large
projects, we selected the Linux kernel as a program analysis test-bed and
provided the reader with reasons of this decision.

The thesis further contains the state of the art of selected program analysis
techniques.  To recall, we provide a detailed description mainly of the
abstract interpretation, symbolic execution, and program slicing. We also
shortly introduced the pattern matching, data-flow analysis, and model
checking. All these techniques are widely spread in the area of the program
analysis. We conclude the state of the art with tools capable of checking
large projects, especially those capable of handling the Linux kernel.

There are four main contributions of this thesis. The first is the techniques
survey in the state of the art. We tried to extensively explore the current
state and document it. The other three contributions are as follows:
\begin{enumerate}
\item \stan\ -- a freely available framework for easy implementation of static
code analyzing techniques.  The framework provides needed information about
the analyzed code and it also comprises a functionality for traversing the
information. To demonstrate the functionality, \stan already ships with
checkers based on various approaches like the abstract interpretation. The
framework contains many other useful helpers such as false positive filters.

We emphasize that \stan is able to handle large software projects like the
Linux kernel. We analyzed the kernel in version 2.6.28 and reported over one
hundred real errors to the kernel developers.

\item Since techniques implemented in \stan still produce many false
positives, we introduced a new technique to lower their occurrence. The
technique is based on a combination of three known approaches:
instrumentation, slicing, and symbolic execution. It is an exact technique
thanks to the symbolic execution and applicable to real code thanks to the
slicing. We implemented the technique in the experimental tool \symbio.

\symbio was experimentally employed to classify the output of \stan from the
Linux kernel 2.6.28. We evaluated the combination like a promising technique,
but there are still a few drawbacks in the symbolic executor we use. The
drawbacks are described in other publications about the symbolic execution and
need to be tackled in the symbolic executor.

\item The results produced by our tools while checking the kernel were not
useful only to the kernel developers. After we figured out the classification
of the reports whether they are false positives or real errors, we established
the classified bug database \cdb. It can serve as a benchmark for developers
of new code analyzers. The developers can run their tool on one of the \cdb's
projects and compare their new tool how it stands with respect to hit and
missed errors, and how many false positives are produced by their tool too.

\end{enumerate}

Since \symbio is an exact tool, we enrolled it to the software verification
competition SV-COMP 2013. Despite the results were not so encouraging due to
various reasons, it helped us to improve the tool and the competition
benchmarks a lot.

\paragraph{Future Work}
We plan to contribute with tuned \symbio in the software verification
competition SV-COMP 2014. This includes a new version of \symbio. Further we
plan to maintain all \stan, \symbio, and \cdb, possibly with augmenting \cdb
by new data.



\clearpage
\bibliographystyle{alpha}

\phantomsection
\addcontentsline{toc}{chapter}{Bibliography}
\bibliography{dis}

\clearpage
\appendix
\chapter{Author's Contribution}
\section{List of Publications}
Here we list all the publications the author of this thesis contributed to
during the PhD studies. We also try to estimate the author's participation
ratio in per cents.  This is not easy as we have to consider several criteria
like who came up with the idea, implemented it, wrote up the publication,
presented it etc. Hence the per cents are only subjective approximations.

We also declare, that this thesis is based on publications~\ref{pub_svcomp},
\ref{pub_cdb}, \ref{pub_fmics}, and \ref{pub_stanse} below. All were presented
by the author of this thesis on the respective conferences or workshops.  All
of them were international and the papers underwent a review process.  Three
of the conferences/workshops are ranked by the ERA conference ranking as
\emph{A}, \emph{B}, and \emph{C}. Another two author papers which the thesis
is not based upon were published on \emph{A}-ranked conferences. Additionally,
one of them was presented by the author of this thesis too.  The author was
also \emph{invited} and pleased to talk about the work from the publication
\ref{pub_stanse} at the embedded world conference 2010. He was also invited to
talk about the Linux kernel development at the NFX workshop co-organized by
the Tomáš Baťa University in Zlín.

\begin{enumerate}
\item \bibentry{cse_atva}. [Contribution: 25\,\%]

\item \label{pub_svcomp} \bibentry{symbiotic-svcomp}. [Contribution: 75\,\%]

\item \label{pub_cdb} \bibentry{claburedb}. [Contribution: 80\,\%]

\item \label{pub_fmics} \bibentry{symbiotic}. [Contribution: 70\,\%]

\item \label{pub_stanse} \bibentry{stanse}. [Contribution: 40\,\%]

\item \bibentry{web2_isd}. [Contribution: 30\,\%]

\item \bibentry{web2_synasc}. [Contribution: 30\,\%]

\end{enumerate}

\section{List of Software}
This is a list of author's contribution to software projects. All of them are
properly described in this thesis as is denoted in the list below.

\begin{enumerate}
\item \stan (Chapter~\ref{sec:stanse})
\item \symbio (Chapter~\ref{sec:fmics}) -- The \llvm Slicer to be emphasized.
\item \cdb (Chapter~\ref{sec:clabure})
\end{enumerate}

\end{document}
