% vim: textwidth=78 ai spell spelllang=en filetype=tex sw=2

\chapter{Introduction}\label{sec:intro}
In the contemporary world, we are surrounded by devices making our lives
easier. It counts from modern lifts which speak to us by natural language to
supercomputers helping to treat cancer such as \emph{IBM
Watson}~\cite{watson-cancer}. Since Moore's law~\cite{moore_law} has proven to
hold for the past four decades, most present devices are featuring quite fast
processing units thanks to the increasing count of built-in transistors. This
allows developers to implement many interesting but complex algorithms.
However being written by human beings vastly, it may lead to errant behaviour
of the implementations as they are only as strong as the weakest link in the
imaginary chain. One example is the error that led to a crash of the
\emph{Ariane~5} rocket that should launch a satellite to the orbit of the
Earth~\cite{ariane_5}. \$370M were buried in the middle of nowhere due to a
programmer's omission to check for an overflow in the acceleration
computation.  So these errors obviously cause companies revenue losses. It is
known the later in the development cycle an error is revealed, the more
expensive it is to fix the error~\cite{topten}. The situation becomes even
worse when the software is deployed at the customer's site or in a rocket
flying 13~000~ft above the sea level already.

Companies have been always desiring for cheap solutions perfectly fitting
their business plans. And yet, solutions which are based on algorithms that
will cause no harm when deployed, completely error-free at best.  But they are
out of luck as, in general, there cannot be a tool which would tell us whether
some algorithm is error-free. If there were one, it would be in contradiction
of the \emph{Halting problem}~\cite{halting_problem}.  However while it is
impossible in theory, it does not necessarily mean we should not at least try
hard for non-general cases when dealing with algorithms implemented in
software. There, the implementations have at least finite state space thanks
to finite memory.  Even though the limited state space is still huge,
companies have been always using techniques helping to avoid and remove errors
from their programs.  Even some kind of \emph{standardization} of used coding
principles like dictated coding style may reduce the number of introduced
errors~\cite{topten}. The reason may be that a project management imposes a
duty on developers to write simpler code. For example, the management can ban
complex constructs which are known to introduce errors. Or, for some
environments like embedded systems, heap allocations can be disallowed. This
results in weeding out one whole category of programming errors -- memory
leaks. Some companies also started adopting standards like
ISO~9001:2000~\cite{iso9000} or MISRA~C~\cite{misra_c}. Their customers are
assured that the internal management of such a company conforms to a
standardized behaviour that tries to increase quality of developed products.

Standardization cannot by definition evade programming errors completely.
Since the early times of computer technology, programmers in companies, or
testers if they were distinct persons, were forced to test programs.  Testers
were running the programs with diverse input, random inclusive, to make the
program misbehave or crash in order to catch and fix program errors.  This
approach is called \emph{dynamic program testing}~\cite{sw_testing}. It is
used very often as it is relatively cheap and mostly very effective technique.
Thanks to executing the code in real-time with real data, whenever the dynamic
testing reveals an error, it is guaranteed to be genuine (if we look aside
from malfunction of broken hardware).  But it indeed takes a huge amount of
time to perform systematic testing of large projects. And yet, the dynamic
testing needs obviously a runnable executable which is not usually available
in the early phases of development.  Testers can help in that case by creating
simple binaries that exercise only finished subsystems, by so-called
\emph{unit tests}. However tests, especially the manually written ones, often
do not cover all possible paths in the program.  Actually, it was shown that a
certain set of unit tests developed for over fifteen years could be augmented
to cover sixteen more per cents of code. This was achieved by
\klee~\cite{klee} using automatic sophisticated techniques for tests
generation.

\emph{Compilers} can also help to warn about many primitive errors introduced
into a program by a programmer. But compilers are designed to be really fast
in compiling code, so their primary goal is definitely not error finding.
Hence many warnings reported by compilers are rather by-products of performed
optimizations and analyses needed for the intermediate compilation phases. We
have to emphasize that an advanced checking of code got introduced only by
highly optimizing compilers, they were not in the prior compilers at all.
Separate projects like \textsc{Lint}~\cite{lint} or
\textsc{Dave}~\cite{dave} performed this role and were widespread for checking
at that time. These tools were based on pattern matching and also on some
analyses looking at how data flow in the checked code. Mostly, these
techniques were more computationally expensive than compilers, but for that
price, the tools implementing them were giving more accurate results with more
reports overall.  These tools belong to the first attempts to use the
\emph{static program analysis}~\cite{program_analysis}. The static analysis
tries to solve the problems of finding errors in programs automatically,
possibly in the early phases of development, and more importantly in a whole
program and without the urge of the dynamic execution. But unlike the dynamic
execution, static techniques are often prone to the presence of reports that
are not real. Those are errors found in the checked code by an analyzer, but
they can never occur in practice, only the analyzer is imprecise. There are
many ways to cope with such reports.  For example, tools like \textsc{Lint}
suggest a programmer to annotate the checked code, so that the tools have a
clue and can thus generate less invalid reports.

%\todo{overview in~\cite{analysis_old_survey}}
In fact, much more sophisticated techniques like of Cousots~\cite{cousot} or
King~\cite{se_king} were described theoretically couple years before
\textsc{Lint} was presented.  However they came of real use in the static
analysis even lately, when computers become powerful enough to finish
computation of such analyses in a reasonable time.  Now, having the computation
power and multi-processor (thousand cores are not exceptions today) machines,
we witness a rapid improvement of the situation over the past two decades. We
could also see many new techniques emerging. Many of those were adopted or
even developed by industrial companies to make their products safer. For
example, Microsoft introduced \textsc{Sdv} toolkit~\cite{sdv} based on
\textsc{Slam}~\cite{slam_cav, slam_popl, slam_decade} and later \textsc{Sdv}
2.0 with \textsc{Slam2}~\cite{slam2} to improve stability of Windows
drivers.  Microsoft also released for internal use \sage~\cite{sage} which
found errors in core Windows parts like the parser of cursor files. Other
companies like Coverity, Gimpel Software, GrammaTech, Klocwork, or MathWorks
presented their products~\cite{coverity, pclint, codesonar, klocwork,
polyspace} to offer a service to check code of their customers. Some of the
companies also check open source code like the Linux kernel or core Linux
libraries. Although the companies make the results
available~\cite{coverity_scan}, developers of the kernel or libraries usually
have no control over the setup and check frequency, therefore they have no
chance to check their code on demand.

For those reasons, some groups decided to implement their own, freely
available, checkers. For example \textsc{Cppcheck}~\cite{cppcheck},
\fbugs~\cite{findbugs}, \smatch~\cite{smatch}, \sparse~\cite{sparse}, or
\stan~\cite{stanse}.  They vary in precision, speed, and of course in the used
method for finding bugs.  Although developers may use them at any time at
their will, they often get results with much higher unreal reports rate than
in results of a commercial tool.

\section{Motivation}
In an ideal world, every code checking tool is capable to handle and check any
project. Unfortunately, this is not true in the real world we live in. What
are the difficulties we can hit when analyzing large software projects? And
why should we care? Those are exactly the questions which this section gives
answers for. First, specifics of large projects are enumerated. Later, we
explain why it is important to find as many errors as possible in them.

\subsection{Requirements for Analyzers}
If code analysis tools were perfect, we could probably skip this section.
But this is not true, the tools are often incomplete in some sense when it
comes to analyzing real large-size code.  We can likely conclude that every
problem in handling of large software projects stems from a single source:
\emph{complex code}. Now let us split the source into several categories and
comment each of those. An observing reader may notice that not all of the
categories are specific solely to large projects. But we understand the list
rather as an enumeration of items which are a must for a code analyzer
evaluating large projects.
%This will happen from the most abstract level to simple code constructs.

\begin{description}
\item[Organization of Files] Source code implementing project's functionality
is often divided into files and files are organized in directories. It is
important to know where the analyzer can find all those files which constitute
the project.  Should a checker just recursively crawl a specified directory
and check all sources in there?  What if some source file includes another one
and the inner one is not compilable on its own as happens in the Linux kernel?
Even if a user specified the files manually, large projects tend to pass
special directives to a compiler. The directives often specify paths to header
files, definitions of some macros from the command line, or pre-inclusion of
specific files.  And since the user of the analysis tool need not be a
developer of the checked project, writing the directives down by hand is
infeasible for large projects.  Checking tools which work on the source level
definitely need to support \emph{build systems} and infer the files and
compiler directives automatically. The tools cannot depend on the user's
input.

\item[Source Code Language] It is not often easy to read and parse source
code of a large project. This is due to languages used for writing code of
such projects.  Even though there are standards for languages, there are many
language offshoots that ``evolved'' with respective compilers. Let us consider
C with the last standard issued in 1999~\cite{ansi_c}. There are many
deviations from the standard introduced by \gcc~\cite{gcc}, \textsc{Intel
CC}~\cite{intel_cc}, \textsc{MSVC}~\cite{msvc}, and possibly others.  Every
single of them supports a different superset of the standard.  Thus it is very
hard for analyzers to implement a parser which would be universally capable of
parsing whichever program with sources in a specified language. This causes
problems when analyzing large projects since there is a high probability of
using many features of some offshoot implemented in some compiler used in the
projects by default.
%code written by diverse people+large chance to hit something unusual in the
%large code: all diverse kind of code, whatever one can think of

\item[Much Parsed Information] When a checking tool is able to parse some
project's code, intermediate code structures are created from the code. But
information about all these structures barely fits into the memory for large
projects. We have to think how and which information to store and where. For
example dumping the information into files and restoring them later. Or
computing only minimal data which sum the effect of the parsed code and
dropping the large structures afterwards.

\item[High Computation Cost] Some techniques cannot be applied to real
software due to a high computation cost.  When developers of large project
analyzers choose techniques to implement in their analyzer, they must have in
mind the large amount of code which needs to be handled. The state space to
be explored can be larger by many orders of magnitude in comparison to an
analysis of small programs. Although computers are fast enough to deal with
more precise analyses today, we are still not there to fully prove there is no
error in some large project. The increase of computation power can be
demonstrated on the symbolic execution~\cite{se_king}. It started being used
extensively almost thirty years after its introduction because the computers
were not so powerful and nobody could really use it for anything of real size.

\item[Higher Level Analysis] The best code practices suggest to split
functionality among smaller logical pieces for better readability and
maintainability. Programmers usually put these pieces into separate functions.
But these pieces alone are not useful when finding errors. They are potential
sources of invalid reports. Consider the example of
Figure~\ref{fig:intro_need_context}.
\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}
DEFINE_LOCK(some_lock);

void my_lock() {
  lock(&some_lock);
}
void my_unlock() {
  unlock(&some_lock);
}
void fun() {
  my_lock();
  my_unlock();
}
\end{lstlisting}
\end{tabular}
\end{center}
\caption{Code to demonstrate the need of analyses to work on a higher level
than functions.}
\label{fig:intro_need_context}
\end{figure}
There, \code{my_lock} locks the lock \code{some_lock}, \code{my_unlock}
unlocks it and they are called in a sequence from \code{fun}.
If we analyzed \code{my_lock} and \code{my_unlock} alone, we would receive two
invalid reports. First, that \code{my_lock} is left with \code{some_lock}
still held. Second, unlock of an unlocked lock in \code{my_unlock}. But when
all three are analyzed together, no such report is generated. This drags in
the need of more wise analyses which look at code from a higher level.

\item[External Dependencies] Large projects tend to use some external library
very heavily, e.g.\ for reusability purposes. So to have a complete analysis
these libraries have to be analyzed too. But there is a slight problem when
the libraries are written in a different language. Still, the analyzer can be
taught to understand that language. But it can be worse, when no sources are
available at all. In this case, there are still some possible solutions.
First, we can develop an analyzer working on the binary level where no sources
are needed.  This was demonstrated in \sage~\cite{sage} for example. Second,
we can ignore calls to unknown code. But we lose precision of course. Finally,
we can solely aim at projects that are self-contained. They do not depend on
any other code.  These are kernels of operating systems for example.

\item[Wide Range of Properties] Large projects usually contain diverse parts,
which in turn contain a handful variety of errors. Not necessarily, but
analyzers are desired to report as many various reports as possible during a
single run.  Even if a project is highly specialized, i.e.\ there is some
class of errors missing, the tools still should find some other classes of
errors. Running several tools is possible and often performed, but it is much
easier to run a single tool which covers more error types.

\item[No Manual Annotations] Annotations are used sometimes to suppress
invalid reports from the output. They give clues to an analyzer about code
which is analyzed. But unlike code of smaller projects, code of large projects
cannot be annotated manually. First, because nobody really knows all
interactions fully as the projects are usually developed by independent groups
of people.  Second, it is not trivial to annotate such a large project
manually. The burden posed on developers would be unbearable.

\item[Concurrency] It is of high probability, that large projects support
creation of threads which can run in parallel. This adds another class of
errors to them: concurrency errors. These are errors that result from an
unexpected scheduling of threads, mostly because invalid or missing critical
sections (locking).  Many checked properties like invalid pointer dereferences
do not necessarily need to be thread-aware. But tracking complex locking
errors or deadlocks using the static analysis needs an analyser, that can
handle threads creation and scheduling of processes.

\item[Complex Environment] Usually, implementations of projects run in an
environment that is somehow initialized. It is a running kernel which serves
system calls, disk content, input of a user, etc. The larger a project, the
more complex environment. Analyzers have to cope with that, usually by
modelling the environment. But it is not an easy task.  If modelled
incorrectly, analyzers relying on a precise environment would produce invalid
reports.

\end{description}

There are perhaps more requirements similar to those listed above, we tried to
list especially the important ones.  We can sum up with an overall view: the
analyzers have to be as much precise as possible, without much user
intervention, but still reasonably fast to finish the analysis in the
foreseeable future. A research and implementation of such an analyzer is
indeed a challenge!

\subsection{Picking the Project}
Since this thesis deals with an analysis of large projects, we have to select
some project for those purposes.  There exist hundreds of software projects of
large size which we can consider. Let us first divide the projects roughly
into two categories.  According to the design of operating systems there are
two basic privilege levels in which programs run~\cite[pages
21--23]{os_concepts}.  Kernels of the operating systems often run at the level
with no limitations, i.e.\ in the \emph{kernel mode}.  There, the kernel can
touch any memory in a machine and can also change other possible reachable
resources.  On the other hand, user space programs tend to be restricted,
i.e.~they run in the \emph{user mode}. There, a process cannot change any
system resource or memory of others, if not explicitly allowed to, e.g.~by
shared memory~\cite[pages 124--126]{os_concepts} or special system
calls~\cite[pages 62--73]{os_concepts}.  In particular, whether some code runs
in the kernel or user mode has a direct impact on the effect caused by a
potential error in the code. A crash of the kernel means that the machine
ceases production any further actions, whereas a crash of a user space
application causes mostly only pain to its user.

Despite the urge for error-free operating system kernels, we also bear
in mind the importance of errors in user space programs. Think about a
vulnerability in a random web browser or database system.  It could indeed
lead to stolen user's personal data or even money from user's bank accounts.
The same as a problem in the kernel. And such user programs consist of millions
lines of code too, so perfectly fits our size requirements. But in our
opinion, operating system kernels still should generally be error-free more
than user space programs. We think so, because even a minor error would allow
an arbitrary attacker to obtain administrator privileges on a machine running
such a vulnerable system, or simply crash it.  For a \emph{denial of service}
(DoS)~\cite{dos} caused by a crash, it sometimes suffices to write a
``killer'' string to a special file locally or even send a packet with a
specific structure over the network. Note the attacker can do it remotely from
a place thousands miles away.  Instances of such errors were already reported
to the Linux kernel developers~\cite{exploit_vmsplice, exploit_proto_ops,
exploit_userns}, in a book about hacking~\cite[page 256]{hacking}, and more to
be found in the National Vulnerability Database~\cite{nvd}. If the attacker
succeeds beyond crashing a machine and is able to get a login there, they are
very close to stealing whatever data they want from that machine since it is
much easier to gain an administrator shell when logged on as a casual user
already (sometimes simply because remote administrator logins are not
allowed). And yet it is often easier to gain an access to other machines in
the same subnetwork than attacking them from the outside. One of the reasons
is that there are usually no managed firewalls for the traffic flowing among
computers inside the subnetwork~\cite{insider_attack}.

The arguments stated above persuaded us to aim at operating system kernels.
But we can choose from many. There are kernels of BSDs (NetBSD, FreeBSD,
OpenBSD, \dots), GNU Hurd, GNU Mach, Linux, Mac OS, Minix, NetWare, OS/2, Sun
OS, Solaris, Windows, and many others. Nonetheless, we limited ourselves
specifically to the Linux kernel. The main reasons are that the kernel:
\begin{itemize} \item is an open source project,

\item is a large code base consisting of more than fifteen millions lines of
code (version 3.10),

\item is used widely, reportedly on 42\,\% of consumer computing
devices in 2012~\cite{linux_42_percent}, and

\item is developed so fast that each version released every two--three months
contains over 700'000 changed lines of code which are across more than eleven
thousand files (averaged for all releases between 3.5 and 3.10). This means a
relatively high probability of introducing new errors.

\end{itemize}
Let us emphasize that the Linux kernel, the same as other kernels, has another
positive aspect: it is self-contained.  And there is still the author of this
thesis who is always biased when choosing a project for code analysis. He
constantly spends some time by Linux kernel development, so he knows the
processes and has some considerable knowledge about the code too.

\section{Background}
This section introduces the reader into terms which we use throughout this
thesis.  First, we enumerate important terms that describe classes of reports.
Next, since every implemented analysis has some properties, we describe
properties which we consider important. Such properties are usually looked at
when deciding which of the available code analyzers to choose for a developed
program. Finally, there is also a small introduction to the code parsing as we
deal with source code in this thesis and the code needs to be parsed in our
tools first.

\subsection{Classes of Reports}\label{sec:intro_reports_cls}
During a code analysis of some project, code analyzers produce reports about
found errors.  Not every report is equally useful. Some of the reports may be
invalid with respect to some specified property which the analyzers try to
check in code. For example, imprecise static analyzers may flag a property
violation somewhere in the analyzed code, but the property is never violated
when the code is really executed. This led to a classification of the reports
according to their usefulness. As we use the classification throughout the
thesis, it is settled down here.

\begin{itemize}
\item \textbf{Errors} -- Reports in this category are useful. We put reports
from some tool to this category, when we conclude that such reports are
genuine and can really cause a problem in the analyzed code. They obviously
require an action to be taken, e.g.\ a fix of the code.

\item \textbf{False reports} -- This category is further split into two
classes which we describe below. Both of them mean incorrect output and such
output should be minimized in real code analyzers as it bothers users. When
we mean both classes in the thesis, we use the term \emph{false reports}.

\begin{itemize}
\item \textbf{False positives} -- These are errors reported by an analyzer
about a property violation. But when such error can never happen when the
analyzed code is really executed, it is false. The analyzer should be tuned to
not produce such reports.

\item \textbf{False negatives} -- These are only imaginary reports as they are
not reported by analyzers. These are errors present in analyzed code which an
analyzer left unnoticed even though they violate a specified property.
Analyzers should be tuned to minimize false negatives, i.e.\ to report as many
property violations as possible (but not many false positives, if some at
all).

\end{itemize}

\end{itemize}

\subsection{Theoretical Properties}\label{sec:theor_props}
Analyzers produce some reports to users. As we described in the previous
section, the reports are classified to different categories. But it still
remains to describe how we can recognize from an analyzer description whether
the analyzer produces reports in any of the false categories or only real
errors. The information about the analyzer's output is interesting for users
to know which report categories are produced by their favourite checker.  In
this section, we define a few properties which characterize analyzers and
mainly their analyses with respect to the specified (checked) properties. Note
that these are not the only possible definitions of these terms. This
description reflects how we understand the terms in this thesis.

\begin{figure}[tb]
  \begin{center}
    \tikzstyle{el_node}=[draw, ellipse, align=left]
    \begin{tikzpicture}[node distance=8mm]
      \node [el_node, fill=blue!10, minimum width=11.5cm,
	minimum height=5cm] (EL1) {};
      \node [below=0.8em of EL1.north] {Sound Analysis};

      \node [el_node, fill=red!10, minimum width=8.5cm, minimum height=3.5cm,
	xshift=-8mm, yshift=-3mm] (EL2) {};
      \node [below=0.8em of EL2.north] {All Program Errors};

      \node [el_node, fill=green!10, minimum width=5cm, minimum height=1.5cm,
	xshift=-20mm, yshift=-6mm] (EL3) {Complete Analysis};

      \node [circle, inner sep=0pt, minimum size=1.5mm, fill=black,
	below left=4mm and 12mm of EL1.east] (FP) {};
      \node [above=0.5mm of FP, align=center] {False \\ Positive};

      \node [circle, inner sep=0pt, minimum size=1.5mm, fill=black,
	below left=5mm and 15mm of EL2.east] (FN) {}
	edge [dashed, >=stealth', <-] (EL3);
      \node [above=0.5mm of FN, align=center] {False \\ Negative};
    \end{tikzpicture}
  \end{center}
  \caption{Theoretical properties of analyzers demonstrated. There is also a
    false negative of a complete analysis and a false positive of a sound
    analysis.}
  \label{fig:intro_theor_props}
\end{figure}

\begin{description}
\item[Soundness]
A program analyzer is \emph{sound} when it finds no error in a program and
there is really none. In other words, it cannot omit to report an error that
can appear at runtime in some execution environment. Since sound analyzers
often over-approximate a checked property, they suffer from \emph{false
positives}. On the other hand, since sound analyzers find all errors of a
defined class, they are error eradicators.

\item[Completeness]
A program analyzer is \emph{complete} when it never produces false positives.
Complete analyzers are prone to missing errors, i.e.\ they produce \emph{false
negatives}. So obviously, complete analyzers are also those which produce only
a few definite reports or no reports at all. Although complete, these
analyzers would be useless in practice. So complete analyzers should be
understood in the context of soundness.

\item[Termination]
A program analyzer \emph{terminates} when it finishes computation in a finite
time. A subject to non-termination is some form of simulation or the symbolic
execution described in Section~\ref{sec:se} for example. When a program
contains an endless loop, analyzers based on these two techniques can simply
run forever. As a side note, some definitions also require complete analyzers
to terminate.

\end{description}

We provide an illustration of soundness and completeness in
Figure~\ref{fig:intro_theor_props} for better understanding. We depict also a
false negative of a complete (but unsound) analysis and a false positive of a
sound (but incomplete) analysis.

Ideally, we would have sound, complete and terminating analyzers. They would
return only real errors with no false positives or negatives. But computation
time as well as memory consumption of such analyzers tend to be high. This
stems from the need of too much precise computation along with precise results
which would have to be stored somewhere. So it is very common to use analyzers
that are only terminating and sound, but incomplete. They report everything
which is possibly an error, but often, we have to await false positives.
Many analyzers are only terminating and complete, but unsound. They are
precise, but do not find all errors. Further, many commercial tools are
terminating, but neither sound, nor complete. They are good at finding some
errors without introducing many false positives. Other combinations of the
properties are possible too, but only the common ones are mentioned here.

\subsection{Sensitivity of Analyzers}\label{sec:sensitivity}
Another interesting property of many analyzers is how they are handling
separate statements and functions present in analyzed code. Consider the
example of Figure~\ref{fig:lst_sens} and also consider an analysis that is
watching values stored into variables. The analysis is able to report to the
user what value can be held in a variable during an execution.
\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}[language=C]
int fun(int a) {
  int x = 1;

  if (a == 2) (*@ \label{lst:sens_cond} @*)
    x = a; (*@ \label{lst:sens_xa} @*)
  x = 3; (*@ \label{lst:sens_x3} @*)
  return x; (*@ \label{lst:sens_ret} @*)
}
void fun2(void) {
  fun(5);
}
\end{lstlisting}
\end{tabular}
\end{center}
\caption{Code used for demonstration of analyses sensitivity.}
\label{fig:lst_sens}
\end{figure}
Given the example, we describe the following three basic properties which we
usually consider when choosing an analyzer:
\begin{description}
\item[Flow sensitivity] When an analyzer analyzes some code, it may take into
account the order of program statements and remember what is followed by what.
For the code in our example this would mean the analyzer will report that the
only possible value in \code{x} at line~\ref{lst:sens_ret} is simply 3 (the
last store to \code{x} before \code{return} proper). The other option is to
completely ignore the order and with considerably less precision report that
\code{x} at line~\ref{lst:sens_ret} can be any of 1, value of \code{a}, or 3
(the analyzer is not aware of what precedes the \code{return}, hence it has to
report all the set commands).  This is called \emph{flow sensitivity} or
\emph{insensitivity}, respectively. We often speak about this property only
when dealing with one particular function.

\item[Path sensitivity] When an analyzer is flow sensitive, it may or may not
remember conditions used for branching. If it does, in our example, \code{x}
is then reported to be one of 1, 2, or 3 on line~\ref{lst:sens_ret}, because
\code{a} was noted to be 2 before the assignment on line~\ref{lst:sens_xa}. In
the other case, \code{x} is announced to be 1, 3, or whatever value the
analyzer previously computed \code{a} can be. We have just described
\emph{path sensitivity}.

\item[Context sensitivity] An analyzer is called to be
\emph{context-sensitive}, or \emph{inter-procedural} if it considers a calling
context (call-stack).  In our example, a context-sensitive analyzer would say
\code{x} can be 1 or 3 (on line~\ref{lst:sens_ret} for flow-sensitive
analyses) since the condition \code{a == 2} is false on
line~\ref{lst:sens_cond}. The insensitive counterpart would say \code{x} can
be any value due to the assignment on line~\ref{lst:sens_xa} (the analyzer has
to assume \code{a} can contain whatever value to remain sound).

\end{description}

\subsection{Types of Analyses}\label{sec:types_anal}
Depending on information provided and the way to compute it, analyses may be
categorized into sets. In this respect, there are four possible combinations
of categories described here. We present some examples of analyses for each of
the categories in Section~\ref{sec:data-flow}.

\begin{description}
\item[May/must] This property says how much definitive an answer about some
property is. Sometimes we are interested whether some result given by an
analysis must hold definitely or may hold sometimes. If we get back to the
example of Figure~\ref{fig:lst_sens}, when an analysis says \code{x}
\emph{must} be a constant on line~\ref{lst:sens_ret}, we can just return that
constant and eliminate all the preceding code. This would not be possible with
an answer ``maybe'' given by the \emph{may} analysis. Nonetheless, may
analysis is useful for example if we are interested whether a divisor in a
division may be zero sometimes, in which case we would report an error.  Must
analysis does not necessarily find this bug.

The definition provided here is how it is understood by many compilers and
analyzers. That way it is defined by Muchnick~\cite{muchnick} and possibly
others.  There exists also another loosened definition of may/must analyses.
The analyses may be understood only as super/sub-sets of real behaviour. Hence
must analysis are an under-approximation of some property, i.e.\ a must
analysis describes behaviour that definitely must hold on \emph{some} path.
Similarly for may analysis, an over-approximation of some property is
computed.

\item[Backward/forward] For flow-sensitive analyses, the way of propagation of
the information is important.  For example, we may want to know at each line
of a program, what variables are used further in the code (to see whether we
need to hold their values still in registers for example). This information is
often propagated backwards from the end towards the beginning of a function
to the point where the variable is defined.  Contrary, to see at each line,
what variables are already set to some value at that line, analyzers usually
propagate the information forwards.

\end{description}

\subsection{Other Views}
Of course the list above is not complete. There are many other categorizations
we do not list in here. In short, we can mention the white-box versus
black-box approach of analyzers. The algorithm either knows the internals of
the checked projects or not.  The black-box approach has no or little
knowledge about checked code and does not really care about the internals.
Hence it will most likely cover less code than a white-box analyzer which
cares about the internals of the checked code.  An analogy might be seen in
checking an engine prior buying a car.  Asking a seller for taking a ride,
check whatever we see and check whether an output from that is as expected
(speedup after pressing the throttle pedal for illustration) is the black-box
approach, because we are unaware about the internals. Asking a technician to
inspect the engine of the car is the white-box approach, because he has a
knowledge about the internals and will check them.  There is a mid-way called
the grey-box approach. This approach requires only an overview of the checked
code, but not full understanding of what is going on in there.

Another angle of view might be hunting functional versus performance issues.
Once a project is working nicely with respect to the demanded functionality,
we may still want to look where we can speed up the implementation. This is
often seen in operating system kernels development as behaviour of a kernel
has to be transparent and as fast as possible.

\subsection{Code Parsing}
We illustrated some terms above directly on code written in programming
languages. This is a form understandable by people when reading, but it is not
so easy to work with when writing an analyzer. Analyzers are used to work on
more convenient structures representing the source code.  There are several
structures used in code analysis and compilers.  This section covers the
process of creating these structures from source code, including a description
of the created structures proper.

First, to be able to understand source code on the input, an analyzer must
know how to interpret it. For each programming language, we have a set of
rules. They define allowed \emph{characters} (usually basic alphabet with
parenthesis, curly braces, and other symbols), how to build words from them
and a meaning of the words: what are identifiers, reserved words etc. Finally,
the rules specify how a well-defined sentence looks like, i.e.\ how to build
expressions and statements, functions and whole files. This is accompanied
with \emph{semantics} which is an explanation what each such block means. So
when we write for example ``\code{char X;}'' in C, we know we have declared a
variable which is capable to hold at least 8 bits (an ASCII character).  The
rules are defined for most languages in the form of \emph{standard}. For
instance, for both C and C++, ISO standards exist~\cite{ansi_c, ansi_cpp},
whereas for Java, the rules are defined by the Sun's implementation. Despite
Sun tried to standardize Java twice, they withdrew the drafts both
times~\cite{java_non_std}. So Java is a \emph{de facto} standard established
by its implementation.  Remember that even if a standard exists, there is no
force to push compiler implementors to adhere to that standard. This is seen
in the C family and its compilers. What is compilable by the GNU's C
compiler~\cite{gcc} is not necessarily understood by the Intel's
compiler~\cite{intel_cc} for example.

\begin{figure}[tb]
\begin{center}
    \tikzstyle{pre} = [<-,shorten <=1pt,>=stealth',semithick]
    \begin{tikzpicture}[node distance=5mm]
        \node [tl_state] (S) {Source};
        \node [tl_phase] (LEX) [right=of S, align=center] {Lexical \\ Analyzer}
          edge [pre] node {} (S);
        \node [tl_state] (TOK) [right=of LEX] {Tokens}
          edge [pre] node {} (LEX);
        \node [tl_phase] (PAR) [right=of TOK, align=center] {Syntactic \\ Analyzer}
          edge [pre] node {} (TOK);
        \node [tl_state] (AST) [right=of PAR] {AST}
          edge [pre] node {} (PAR);
        \node [tl_state] (CFG) [above right=of AST] {CFG}
          edge [pre,dashed] node {} (AST);
        \node [tl_state] (OTH) [below right=of AST] {\dots}
          edge [pre,dashed] node {} (AST);
    \end{tikzpicture}
\end{center}
    \caption{Process of creating program structures which are more convenient
    to work with than source codes.}
  \label{fig:parsing_process}
\end{figure}

With the rules, we can start writing a translator that reads source code in
the human readable form on the input and produces some structures on the
output. Such translators are called \emph{parsers}.  The whole process of
parsing is depicted in Figure~\ref{fig:parsing_process}.  As we can see, there
are several stages. We discuss them in further text one by one.  The
explanation happens on the C programming language, more concretetely on the
code of Figure~\ref{fig:intro_parsing_codes}~(a).

\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
    \begin{tabular}{ccc}
    \begin{lstlisting}[language=C]
void fun(int a)
{ /* comment */
  int b = 0;
  if (a)
    b = 5;
  a = 3 * b;
}
    \end{lstlisting} & & {
    \begin{lstlisting}[numbers=none]
key:void id:fun LPAR key:int id:a RPAR
LCUR
  key:int id:b EQ num:0 SEMIC
  key:if LPAR id:a RPAR
    id:b EQ num:5 SEMIC
  id:a EQ num:3 MULT id:b SEMIC
RCUR
    \end{lstlisting} } \\
    (a) & & (b)
    \end{tabular}
\end{tabular}
\end{center}
    \caption{(a) code example for parsing demonstration; (b) tokenized code.}
  \label{fig:intro_parsing_codes}
\end{figure}

%Some standards also contain a \emph{formal grammar}.

First, the input is read character by character by the \emph{lexical analyzer}
(lexer).  It consults with language rules what the input means and starts
creating so-called \emph{tokens}. Tokens are character sequences that are
marked by information whether they are a string, number, reserved keyword
(\code{int}, \code{const}, \dots), identifier (\code{some_variable}), or any
other kind defined in the rules. We can see how our example was tokenized in
Figure~\ref{fig:intro_parsing_codes}~(b).  Keywords, identifiers, and numbers
are prefixed by type, braces, parenthesis, and operators are converted to
special names, and so on (depending on the language rules). Comments and other
similar information are dropped.

A list of these tokens is then passed on the input of the \emph{syntactic
analyzer}. Again, it implements the language rules and looks whether the token
sequences are correct. In case they are, syntactic analyzers usually build the
\emph{abstract syntax tree} (AST). All irrelevant information is forgotten
there (abstracted from the original language) and the relevant part is stored
in a tree as is depicted in Figure~\ref{fig:intro_parsing_ast} for our
example.
\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
    \begin{tikzpicture}[level/.style={sibling distance = 2.5cm/#1,
    level distance = 1cm}]
      \node [ast] {\texttt{fun}}
	  child [sibling distance=1.8cm] { node [ast] {\texttt{void}} }
	  child { node [ast_hl] {param}
	    child { node [ast] {\texttt{int}} }
	    child { node [ast] {\texttt{a}} }
	  }
	  child [sibling distance=4.5cm] { node [ast_hl] {body}
	    child [sibling distance=2cm] { node [ast_hl] {decl}
	      child [sibling distance=1.3cm] { node [ast] {\texttt{int}} }
	      child { node [ast_hl] (L_1) {var}
		child { node [ast] {\texttt{b}} }
		child { node [ast] {\texttt{0}} }
	      }
	    }
	    child [sibling distance=2cm] { node [ast] {\texttt{if}}
	      child { node [ast] (L_2) {\texttt{a}} }
	      child { node [ast] (L_3) {\texttt{=}}
		child { node [ast] {\texttt{b}} }
		child { node [ast] {\texttt{5}} }
	      }
	    }
	    child [sibling distance=2cm] { node [ast] (L_4) {\texttt{=}}
	      child { node [ast] {\texttt{a}} }
	      child { node [ast] {\texttt{*}}
		child { node [ast] {\texttt{3}} }
		child { node [ast] {\texttt{b}} }
	      }
	    }
	  }
	  ;
	  \node [red,font=\scriptsize,above=-1mm of L_1] {1};
	  \node [red,font=\scriptsize,above=-1mm of L_2] {2};
	  \node [red,font=\scriptsize,above=-1mm of L_3] {3};
	  \node [red,font=\scriptsize,above=-1mm of L_4] {4};
    \end{tikzpicture}
\end{tabular}
\end{center}
    \caption{Abstract syntax tree of the code of
    Figure~\ref{fig:intro_parsing_codes}.}
  \label{fig:intro_parsing_ast}
\end{figure}
At this abstract point, we have no clue what the original language was.
Compilers usually have separate modules for different languages, called
\emph{frontends}. They parse the input and create one pre-defined form of the
AST. Every frontened does so, hence we have one type of the AST for all
diverse languages the frontends support.  Then, the rest of the compiler
pipeline can be the same for all of the supported languages and the compiler
can run the same optimization passes for all the languages. The AST is
held mostly in the compiler through the whole compilation process. It is usual
that further passes in the compiler pipeline add more information to the nodes
of this tree.  The very common one is a position of the nodes in the source
code for debugging purposes. Even though ASTs usually hold information about
the whole parsed source file, we omitted for simplicity from
Figure~\ref{fig:intro_parsing_ast} other functions, global variable
declarations and other constructs.  There is only a subtree corresponding to
the function from Figure~\ref{fig:intro_parsing_codes}~(a).

From the AST, we can derive much information about the parsed code. For
illustration, we can build a \emph{control flow graph} (CFG). CFG of a
function is a directed graph where nodes contain statements of that function
and edges are labeled by conditions when coming from branching statements. In
a CFG, the \emph{entry point} of a function is a successor of a triangle
rotated to have a vertex on the bottom. The \emph{exit point} (return from a
function) is a predecessor of a triangle with an edge on the bottom. A
\emph{branch} is a node with more than one successor.  As an example we built
a CFG of the \code{fun} function in Figure~\ref{fig:intro_parsing_cfg}. Nodes
of the CFG refer back to the to AST with numbers to see where the statements
come from (every node or subtree in an AST has its rule to create nodes in a
CFG).
\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
    \begin{tikzpicture}[node distance=5mm]
      \node [cfg_start] (S) {};
      \node [cfg] (L_1) [label=right:\red{1},below=3mm of S] {\code{b=0}}
        edge[cfg_edge] node {} (S);
      \node [cfg] (L_2) [label=right:\red{2},below=of L_1] {\code{a}}
        edge[cfg_edge] node {} (L_1);
      \node [cfg] (L_3) [label=left:\red{3},below left=of L_2] {\code{b=5}}
        edge[cfg_edge] node [label=above left:\code{a}] {} (L_2);
      \node [cfg] (L_4) [label=right:\red{4},below=1.3cm of L_2]
	{\code{a=3*b}}
        edge[cfg_edge] node [label=right:\code{!a}] {} (L_2)
        edge[cfg_edge] node {} (L_3);
      \node [cfg_end,below=3mm of L_4] {}
        edge[cfg_edge] node {} (L_4);
    \end{tikzpicture}
\end{tabular}
\end{center}
    \caption{Control flow graph of the code of
    Figure~\ref{fig:intro_parsing_codes}.}
  \label{fig:intro_parsing_cfg}
\end{figure}
Alternatively some people put statements on edges instead of putting them into
nodes for easier explanation of algorithms.  The control flow graph is not the
only structure that can be derived from the abstract syntax tree. We can
derive symbol tables, type definitions, perform semantic analysis to verify
more than syntactic rules and so on.

Even though, it is not easy to write a well-working parser by hand, some
projects like \gcc do so. Such parsers tend to be fast. But there also
exist tools that help to generate lexical and syntactic analyzers
automatically from a specified formal grammar.  Depending on the programmer's
preference on the type of the grammar, there are various tools. For
LL(\textasteriskcentered) grammars~\cite[Chapter 4.4]{dragon}, one of the most
known tools is \textsc{Antlr}~\cite{antlr4}.  \textsc{Flex} and
\textsc{Bison}~\cite{flex_bison} are used to generate LR parsers~\cite[Chapter
4.5]{dragon}. This introduction to parsing does not want to pretend it is
complete, but we do not need more information for the purposes of this thesis.
More information on parsers can be learnt from various literature about
compilers~\cite{dragon, muchnick}.

\subsection{Pointer Analysis}\label{sec:pointer_analysis}
Some programming languages allow for more complicated code constructs. They
require some advanced techniques to be used when analyzing code in these
languages. An example of such constructs are pointers and operations over
them. When pointers are used, it is important to know where they can point to.
Consider the integer variable \code{x} and the integer pointer \code{p} as are
in the code of Figure~\ref{fig:lst_ptsto}.
\begin{figure}[htb]
  \begin{center}
  \begin{tabular}{c}
  \begin{lstlisting}[language=C]
int fun(int z) {
  int x = 0;
  int y = 0;
  int *p = &x;
  int *q = &x;
  *p = z;
  q = &y;(*@\label{line:intro_ptsto_q}@*)
  return x;
}
  \end{lstlisting}
  \end{tabular}
  \end{center}
  \caption{A pointer analysis demonstrative example.}
  \label{fig:lst_ptsto}
\end{figure}
Without knowing that \code{p} points to and actually changes \code{x}, we
would conclude that \code{fun} always returns 0. But it returns whatever it
gets on the input as the parameter \code{z}.

Since pointers are used extensively in C and the main language we deal with in
this thesis is C, we describe here an approach to cope with this situation. It
is called \emph{pointer analysis}.  Nowadays, there are two basic and widely
used approaches to the pointer analysis: Steensgard's~\cite{steensgaard} and
Andersen's~\cite{andersen}. Both analyses are \emph{may}. But while the former
is based on equalities of the sets of possible pointer targets, the other on
their subsets. So the former makes the sets of targets equal whenever two
pointers can point to a common target variable.  The latter only copies some
(or all) members. Let us demonstrate it on the full example of
Figure~\ref{fig:lst_ptsto}. We depict the relation between pointers and
targets as directed graphs in Figure~\ref{fig:intro_ptsto_rel}. Nodes are both
pointers and targets, but edges from pointers to targets mean the pointers
point to the targets.
\begin{figure}[tb]
  \begin{center}
  \begin{tabular}{cc} \adjustbox{valign=m}{
    \begin{tikzpicture}[node distance=10mm]
        \node [pts_node] (P) {\code{p}};
	\node [pts_node, right=of P] (X) {\code{x}}
	  edge [pts_edge] node {} (P);
	\node [pts_node, below=of X] (Y) {\code{y}};
	\node [pts_node, right=of X] (Q) {\code{q}}
	  edge [pts_edge, ->] node {} (X)
	  edge [pts_edge, ->] node {} (Y);
    \end{tikzpicture} } & \adjustbox{valign=m}{
    \begin{tikzpicture}[node distance=10mm]
        \node [pts_node] (P) {\code{p}};
	\node [pts_node, ellipse, inner sep=2pt, right=of P, align=center] (XY) {\code{x} \\[2mm] \code{y}}
	  edge [pts_edge] node {} (P);
	\node [pts_node, right=of X] (Q) {\code{q}}
	  edge [pts_edge, ->] node {} (XY);
    \end{tikzpicture} } \\ \\
    (a) & (b) \\
  \end{tabular}
  \end{center}
    \caption{Pointer graphs produced by two different pointer analyses on the
    code from Figure~\ref{fig:lst_ptsto}: (a) Andersen's and (b)
    Steensgaard's.}
  \label{fig:intro_ptsto_rel}
\end{figure}
We see that the assignment to the pointer \code{q} on
line~\ref{line:intro_ptsto_q} does not lose precision in Andersen's analysis
in Figure~\ref{fig:intro_ptsto_rel}~(a), but puts the targets (\code{x} and
\code{y}) to the equal set using Steensgaard's analysis in
Figure~\ref{fig:intro_ptsto_rel}~(b). This would prevent optimizations to
remove \code{p} and \code{y} from the code even though they are actually
useless. In general, subset-based analyses are more precise than
equality-based, but they are also slower. The slowness is a result of the
count of nodes in graphs like those we have just presented.

Shapiro and Horwitz investigated 61 C programs to verify the accuracy and
speed of the two approaches of Andersen and Steensgaard. The findings were
summarized in a Shapiro and Horwitz's study later~\cite{shapiro_horwitz}. And
they also came up with an approach that tries to speed up the Andersen's
analysis while preserving its accuracy using parametrization. They allow each
node of a pointer graph to have at most degree $k$. Obviously, if $k=1$, we
are talking about Steensgaard's analysis, if $k$ is the number of possible
targets, we obtain Andersen's. If $k$ is kept reasonably small, the results are
often more accurate than Steensgaard's algorithm, but much faster than
Andersen's.

\paragraph{Other Analyses} The \emph{alias analysis}, sometimes referred as
the \emph{points-to analysis}~\cite{alias_anal} is somewhat different to the
pointer analysis above in what it is able to tell about pointers.  The alias
analysis can either conclude that two pointers point to the same location,
that they point to a different location, or the analysis cannot tell anything
about the pointers when the knowledge is insufficient. It cannot report the
set of possible targets as the true pointer analysis does. This is sometimes
sufficient and for example the compiler framework \llvm as of version 3.3
provides its optimization passes only with the alias analysis. For performance
purposes, alias analyses are often computed as a \emph{may-alias}
over-approximation or \emph{must-alias} under-approximation.

Another view to pointers is suggested by the \emph{shape
analysis}~\cite{ptr_shape}.  It is an analysis of dynamically created
structures like lists. It is able to recognize operations like list creation
and give this information to the analyzer. It can be used when finding memory
leaks or tell whether the list is created properly.

\paragraph{Field-sensitivity} There is also another property of a pointer
analyzer to be considered for languages allowing structured constructs, like
\code{struct} in C. We can either consider each of the members in such an
object when computing the pointer sets or ignore the members and work on the
object level only. Consider the example of Figure~\ref{fig:intro_pts_field}.
\begin{figure}[htb]
  \begin{center}
  \begin{tabular}{c}
  \begin{lstlisting}[language=C]
struct S {
  int x;
  int y;
  int *p;
} s;
struct T {
  int x;
  int *p;
} t;

void fun() {
  int *p = &s.x;(*@\label{line:intro_field_1}@*)
  s.p = &t.x;(*@\label{line:intro_field_2}@*)
  t.p = &s.y;(*@\label{line:intro_field_3}@*)
}
  \end{lstlisting}
  \end{tabular}
  \end{center}
  \caption{A demonstration of field-sensitivity for the pointer analysis.}
  \label{fig:intro_pts_field}
\end{figure}
Even though the use of structures in the code is heavy, there are only
disjunct assignments to pointers: all the pointers point to a different memory
location (\code{s.x}, \code{s.y}, and \code{t.x}). Hence a pointer analysis
that does not consider members of structures, would give imprecise results.
This can be seen in the graph of Figure~\ref{fig:intro_ptsto_rel2}~(a), which
is a result of the field-insensitive pointer analysis.
\begin{figure}[b]
  \begin{center}
  \begin{tabular}{ccc}
    \adjustbox{valign=m}{
    \begin{tikzpicture}[node distance=8mm]
        \node [pts_node] (P) {\code{p}};
	\node [pts_node, right=of P] (S) {\code{s}}
	  edge [pts_edge] node {} (P);
	\node [pts_node, right=of S] (T) {\code{t}}
	  edge [pts_edge, bend left] node {} (S)
	  edge [pts_edge, bend right, ->] node {} (S);
    \end{tikzpicture} } &
    \adjustbox{valign=m}{
    \begin{tikzpicture}[node distance=8mm]
        \node [pts_node] (P) {\code{p}};
	\node [pts_node, right=of P] (DX) {\code{*.x}}
	  edge [pts_edge] node {} (P);
	\node [pts_node, below=5mm of P] (DP) {\code{*.p}}
	  edge [pts_edge, ->] node {} (DX);
	\node [pts_node, right=of DP] (DY) {\code{*.y}}
	  edge [pts_edge] node {} (DP);
    \end{tikzpicture} } &
    \adjustbox{valign=m}{
    \begin{tikzpicture}[node distance=8mm]
        \node [pts_node] (P) {\code{p}};
	\node [pts_node, right=of P] (SX) {\code{s.x}}
	  edge [pts_edge] node {} (P);

	\node [pts_node, below=5mm of P] (SP) {\code{s.p}};
	\node [pts_node, right=of SP] (TX) {\code{t.x}}
	  edge [pts_edge] node {} (SP);

	\node [pts_node, below=5mm of SP] (TP) {\code{t.p}};
	\node [pts_node, right=of TP] (SY) {\code{s.y}}
	  edge [pts_edge] node {} (TP);
    \end{tikzpicture} } \\ \\
    (a) & (b) & (c) \\
  \end{tabular}
  \end{center}
    \caption{Pointer graphs produced by two different pointer analyses with
    respect to field-sensitivity on the code from Figure~\ref{fig:lst_ptsto}:
    (a) field-insensitive, (b) field-based and (c) field-sensitive.}
  \label{fig:intro_ptsto_rel2}
\end{figure}
Since structures are used heavily in C, this is definitely not only a
theoretical problem.  Therefore, field-insensitive analyzers are not of much
use there due to imprecisions. From the first graph of the illustrative
example, we can see that the imprecisions are two-fold:
\begin{enumerate}
\item Each member of a structure can be itself a pointer. This means it can
have its own pointer target set. Storing this information in the
field-insensitive sense only on the object level per each structure introduces
a big imprecision. In example of Figure~\ref{fig:lst_ptsto}, there are two
such members with their own pointer sets (\code{s.p} and \code{t.p}).

\item Given some pointer, it is always useful information to know where it can
point to exactly. Especially, if it can point to a member of some structure
and we can decide which of the members in the structure is pointed at. This
can be seen on lines~\ref{line:intro_field_1}, \ref{line:intro_field_2}, and
\ref{line:intro_field_3} in the example of Figure~\ref{fig:lst_ptsto}. We are
interested on exact members (\code{x} and \code{y}) where the pointers
(\code{p}, \code{s.p}, and \code{t.p}) point to.

\end{enumerate}

These two options are not bound rigidly together. They can be combined or
implemented separately in a \emph{field-sensitive} pointer analyzer. And yet,
the field sensitivity can be implemented in two possible ways with impact on
precision.  First, a \emph{field-based} sensitivity has only one global
information about structure members, per each member name. So when we consider
again the example of Figure~\ref{fig:lst_ptsto}, we have two structures
\code{s} and \code{t} of a totally different type. Since they both contain the
member \code{int *p}, the field-based analysis considers them to be the same
and updates \code{*.p}'s points-to set any time \code{s.p} or \code{t.p} is
set to point to a new target. This situation is depicted for our example in
Figure~\ref{fig:intro_ptsto_rel2}~(b).

The other way is a true \emph{field-sensitive} analysis. It properly handles
members of structures in the natural sense, but for a higher computation
price, of course. The pointer graph for this implementation is shown in
Figure~\ref{fig:intro_ptsto_rel2}~(c). There is no imprecision as all three
pointers with their graphs are independent.

For more information about pointer analysis, see for example Hind's
discussion~\cite{ptr_solved}.

%\todo{emphasize that this thesis deals with practical aspects of using static
%analyzers}
\section{Thesis Goals}
This thesis studies an application of contemporary automatic techniques for
finding errors in large software projects. It specifically aims at more
advanced techniques. These techniques include the abstract interpretation and
symbolic execution which we deal with in this thesis. Both of them were
designed in 1970's and were adopted in some tools since then. In the past
decade they came to interest of code analyzers implementers thanks to the
occurrence of fast computers. The thesis poses a claim to summarize the
current state of the art of chosen techniques implemented in code analyzers
which can handle large code bases. The aim of the thesis is also to include a
short description of some analyzers. On the top of the explored state of
the art, the goal is to develop our own techniques and tools which are capable
of handling the Linux kernel.  The tools are asked to produce only a low
number of false positives.

Another aim is to help others with code analyzers development and evaluation.
We established a test-bed with predefined categories of errors which are
usually present in code of projects such as the Linux kernel. For that purpose
a publicly available benchmark is to be designed. It should contain
information about errors and false positives found by several tools finding
errors in large projects.

There is also a software competition tailored to exercise code analyzers. So
another goal of this thesis is to contribute with our analyzers and compete
with others.

\section{Thesis Contribution}
We implemented \stan, a framework for developers who wish to implement some
analysis. The framework provides many helpers, e.g.\ for traversing and
checking code. As it was already prepared for an implementation of code
checking techniques, we implemented several checkers based on different
approaches. The most used one is based on the abstract interpretation. The
downside of \stan is the produced false positive rate.

We categorized some reports of \stan by a new tool called \symbio. It is based
on a combination of three techniques: instrumentation, slicing, and symbolic
execution.  First, an instrumentor injects checked properties into code that
is to be checked. Second, a static slicer removes code irrelevant to the
properties.  Finally, a symbolic executor evaluates the remaining code with
low number of false positives produced on the output.

We used reports of \stan and other available checkers to create a test-bed for
comparing the output of newly implemented code analysis tools. In particular,
it evaluates how good the reports from the tool are.  We started calling the
test-bed \cdb.

Finally, we released \symbio. We used that version of \symbio on a software
competition of code analyzers. We plan to participate with a tuned version of
\symbio again next year.

\section{Thesis Organization}
Chapter~\ref{sec:state} deals with a description of the current state of the
art of standard code analysis techniques and tools. In
Chapter~\ref{sec:stanse}, we discuss our tool \stan~\cite{stanse} and its pros
and cons. The chapter also refers to the abstract interpretation as it is
implemented in \stan.  We continue in Chapter~\ref{sec:fmics} by an approach
to eliminate false positives resulting from other analyses like the abstract
interpretation. This is implemented in a tool called \symbio~\cite{symbiotic}.
Chapter~\ref{sec:clabure} then introduces a database of known errors in large
projects like the Linux kernel.  The database is called \cdb~\cite{claburedb}.
Chapter~\ref{sec:svcomp} presents results of \symbio in a software
competition~\cite{symbiotic-svcomp}.  Finally, we conclude with summaries of
what we have done and some plans for the future research.

%TODO: nemela by v sekci 6 (nebo nekde jinde) byt nejaka evaluace \stan,
%\symbio a dalsich na zaklade benchamrku z \cdb?
