% vim: textwidth=78 ai spell spelllang=en filetype=tex sw=2

\chapter{State of the Art}\label{sec:state}
Later chapters in this thesis deal with techniques used for code analysis.
To fully understand the details behind the text, this chapter comprises a
description of code analysis techniques with their current state of the art.
Since there are many of the techniques, we present a more detailed description
only of techniques referred in later text. Especially, we outline static
analyzers which handle code written in C.  But before we can start with a
deeper explanation, we first provide a definition of how we understand the
term static analysis and what we think is its usual use. This is followed by
an explanation of simpler static analysis techniques. The explanation further
continues toward more complex ones. We conclude the chapter with summarization
of two checkers groups. One group are checkers that we consider important in
the industrial field. Second, we summarize implemented Linux kernel checkers.
Reports produced by most of the latter group are used to form a database in
Chapter~\ref{sec:clabure}.

\paragraph{Static Analysis} In our understanding the term \emph{static
analysis} means analyzing code without directly executing it on a processor.
In other words, the static analysis performs exactly the opposite of the
dynamic analysis where a program is executed and either automatically supplied
with some input in some form or tested by testers manually.  This can be
performed thoroughly only on programs which are either finished or have an
executable binary with most components plugged.  The static analysis approach
has an advantage in this respect: to analyze every part of a program
statically, an executable need not even exist.  Therefore the static analysis
is not as limited as the dynamic one, it can be used whenever it is desired in
the development cycle. This is one of the reasons which make the static
analysis so popular and utilized in several fields of the computer science.
Aside from the code analysis, we list the main fields below:
\begin{itemize}
\item \emph{Optimizing compilers} use results of the static analysis to
eliminate unused code, propagate constants where variables would have to be
used otherwise, or to remove implicitly added runtime out-of-bounds checks if
a compiler knows that the bounds are never exceeded (this is implemented in a
Java compiler).

\item \emph{Debuggers} use results of the static analysis to limit themselves
only to code that is relevant to variables a programmer is interested in.
Debuggers can also print possible sources of the value currently stored in a
tracked variable for example.

\item \emph{Program tests} can be generated using the static analysis to cover
more code. The analysis can establish some knowledge about the code and help
to find inputs that exercise more code.

\end{itemize}

Moreover, if the static analysis is well-defined according to some rules
defined e.g.\ by Cousot's framework~\cite{cousot} described in
Section~\ref{sec:ai}, we are able to decide whether the analysis is sound,
complete, and whether it terminates.  This is useful for example when building
an optimizing compiler to prove it does not generate invalid code.

%the most ready/useful\todo{XXX}

% XXX do we want a picture specification+program -> analyzer -> diagnosis

\section{Pattern Matching}\label{sec:pattern_matching}
The simplest automatic analysis we can think of is to find some predefined
patterns in code of a program.  We can take an instance of a regular
expression pattern matcher like \texttt{grep}, give it a pattern ``/ 0'' and
see if we did a primitive zero division error somewhere in the code. But this
approach may be a source of false reports. This happens when developers
comment their code with comments similar to ``/ 0 happened here''.  But there
of course exist more robust techniques based on pattern matching.  They, in
some way, parse the code (which also removes the problem with the comments
like the one above) and do pattern matching on some internal structure. If we
recall the abstract syntax tree from Figure~\ref{fig:intro_parsing_ast}, we
can design a checker which would match AST subtrees for instance. Then to
match division by zero errors, we simply try to find a subtree in an AST like
of Figure~\ref{fig:ast_to_match}.  The left leaf is left intentionally blank
to express a wildcard.
\begin{figure}[h]
\begin{center}
    \begin{tikzpicture}[level/.style={sibling distance = 1cm/#1,
    level distance = 1cm}]
      \node [ast] {\texttt{/}}
	    child { node [ast] {} }
	    child { node [ast] {\code{0}} }
	  ;
    \end{tikzpicture}
\end{center}
    \caption{A subtree of an AST to match for a primitive division by zero.}
  \label{fig:ast_to_match}
\end{figure}

One of the most known tools based on the pattern matching is
\fbugs~\cite{findbugs}. Despite it is intended to find errors in Java
programs, we mention it here, because it became very popular. This happened
thanks to the ease of use, hundreds of error types it can find, scalability
and precision. It is able to prioritize the reports according to their weight
so that it usually puts real errors on the top of the list in favor of false
reports.

\textsc{Coccinelle}~\cite{coccinelle} is another pattern matcher. The authors
call it rather a semantic patcher, because it can understand the semantics of
the checked code. It uses a special language SmPL for declaring a semantic
patch. The patch is utilized for matching against control flow graph nodes and
changing them arbitrarily. The graph is then converted back to source code.
\textsc{Coccinelle} is used mainly for modifying code to use newer interfaces
like replacing two calls to \code{kmalloc} and \code{memset} functions by a
single call to \code{kzalloc} which does the same as the two together.  Since
it is able to match the control flow graph, it can be used for error pattern
matching too~\cite{wysiwib}. As an example, we can pick missing checks for
allocations: a call to an allocation function with no condition to check the
return value afterwards.

\section{Data-flow Analysis}\label{sec:data-flow}
One of the basic static analyses is \emph{data-flow analysis}. It collects
information about some data flowing through points in code. The data can be
some information about values of variables, about statements, and so on. We
usually look at each program line separately to see what are effects of the
line and then encode the desired information into a \emph{transfer equation}.
It describes relations between what holds before and after the execution of a
particular line. Next, we stabilize a full set of program equations, usually
by computing a point where the equations no longer change. This state is
called \emph{fixed point}. The fixed point is chosen either to be the least or
greatest solution of the equations, depending on the performed analysis. When
we have final data computed by the data-flow analysis, the data can be
utilized for various code optimizations or program analyses.

Let us demonstrate the data-flow analysis on an example and then generalize
the approach. We chose one of the basic data-flow analyses: \emph{reaching
definitions}~\cite[Section 9.2.4]{dragon}. At some point $p$ in code, we are
interested in all assignments where some variable $\var{x}$ was set and was
not re-set during the control flow to the point $p$. This lets us reason about
many questions like whether a variable can be replaced by a constant as an
optimization.  We show the analysis on the code which is in the second
column of Table~\ref{tab:state_df_eq}.  Intuitively, the reaching definition
set on line~\ref{lst:state_rd_ref} is a set of four labels
$\{\ref{lst:state_rd_def1},\, \ref{lst:state_rd_def2},\,
\ref{lst:state_rd_def3},\, \ref{lst:state_rd_def4}\}$, where we define
\code{a} and \code{b}, because all of those definitions can reach
line~\ref{lst:state_rd_ref}.
%Note that line~\ref{lst:reach_def0} does not belong there because \code{a}
%was re-set later on both paths to line~\ref{lst:reach_use}.

Now we derive the information algorithmically using the data-flow analysis by
constructing the transfer equations and computing their fixed point. In the
case of the reaching definitions analysis, we can simply hold the definitions
in the equations.  That is, initially at the beginning of a function, the
equation there holds an empty set. Then this empty set is gradually
transformed according to statements and the definitions are put into the set.
Since we need to propagate information about definitions, there are two
types of propagations:
\begin{enumerate}
\item On statements which are assignments, the contents of incoming sets is
mangled. First, the equation takes the set of statements from the
predecessors. If there are multiple predecessors, we use the set union as we
want to propagate definitions from all branches. Second, the equation reduces
the predecessors' set by all statements that assign to the variable assigned
in the current statement. Finally, the equation adds the current statement.

\item Equations for other statements simply copy data from predecessors (with
a use of the set union, if there are more than one).

\end{enumerate}
We constructed the transfer equation $f_s$ for each statement $s$ of the
example in Table~\ref{tab:state_df_eq}. They are shown in the third column of
the same table.
\begin{table}[tb]
\begin{center}
\begin{tabular}{|r|l||p{0.2em}lp{0.3em}r|} \hline
\textbf{\footnotesize \#} &
  \multicolumn{1}{c||}{\textbf{Code}} &
  \multicolumn{4}{c|}{\textbf{Transfer Equations}} \\ \hline\hline
&& $f_0$ & $= \emptyset$ && \\ \hline

1 & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
a = 1;(*@ \label{lst:state_rd_def1} @*)
\end{lstlisting} } &
$f_1$ & $= \{1\}$ & $\cup$ & $(f_0 \,\setminus\, \{1,4\})$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
b = read();(*@ \label{lst:state_rd_def2} @*)
\end{lstlisting} } &
$f_2$ & $= \{2\}$ & $\cup$ & $(f_1 \,\setminus\, \{2,5\})$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
while (b) {
\end{lstlisting} } &
$f_3$ & $= \emptyset$ & $\cup$ & $(f_2 \cup f_6 \,\setminus\, \emptyset)$ \\
  \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
  a = a+5;(*@ \label{lst:state_rd_def3} @*)
\end{lstlisting} } &
$f_4$ & $= \{4\}$ & $\cup$ & $(f_3 \,\setminus\, \{1,4\})$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
  b = b-1;(*@ \label{lst:state_rd_def4} @*)
\end{lstlisting} } &
$f_5$ & $= \{5\}$ & $\cup$ & $(f_4 \,\setminus\, \{2,5\})$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
}
\end{lstlisting} } &
$f_6$ & $= \emptyset$ & $\cup$ & $(f_5 \,\setminus\, \emptyset)$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_df_comp,numbers=none,frame=none]
use(a);(*@ \label{lst:state_rd_ref} @*)
\end{lstlisting} } &
$f_7$ & $= \emptyset$ & $\cup$ & $(f_3 \,\setminus\, \emptyset)$ \\ \hline

\end{tabular}
\end{center}
  \caption{A reaching definitions data-flow analysis example.}
  \label{tab:state_df_eq}
\end{table}
If we look closer at the example, it suggests that all the equations have one
predefined form:
\begin{equation*}
f_s = gen_s \cup (pred_s \setminus kill_s)
\end{equation*}
where $gen_s$ is the set $\{s\}$ that \emph{generates} $s$ if it is an
assignment, otherwise $gen_s$ is empty for that particular equation.  $pred_s$
is a union of $f_{s'}$ for each predecessor $s'$.  Finally, if $s$ is an
assignment to \code{x}, $kill_s$ is a set that contains other assignments to
\code{x}.  Otherwise $kill_s$ is empty.

Now, we would like to compute a fixed point of the equations. Since the
reaching definitions analysis starts with an empty set and gradually augments
that, we need to compute the least fixed point. We can use
Definition~\ref{def:fixpoint} for that purpose, but we need a single composite
function $F$. We achieve that by rewriting the transfer equations to functions
taking all others functions as parameters. For all statement labels $s$, the
functions look like $f_s=F_s(f_0,\, \dots,\, f_8)$. The composite function is
then $F(\overrightarrow{f})=(F_0(\overrightarrow{f}),\, \dots,\,
F_7(\overrightarrow{f}))$. With the function $F$, we can iteratively compute
the least fixed point. The computation for our example of
Table~\ref{tab:state_df_eq} is shown in Table~\ref{tab:state_df_fp}.

\begin{dfn}[Fixed point]\label{def:fixpoint}
A \emph{fixed point} of a function $F : D \rightarrow D$ is an element $x \in
D$ such that $F(x) = x$. The set $fp(f)$ of fixed points of $F$ is $\{x \in D
\mid F(x) = x \}$.
\end{dfn}

\begin{table}[tb]
\begin{center}
\begin{tabular}{|c||c|c|c|c|c|c|c|} \hline
\multirow{2}{2.7em}{\textbf{Equa\-tion}} &
  \multicolumn{7}{c|}{\textbf{Iterations of Fixed Point Computation}}
  \\ \cline{2-8}
& 1 & 2 & 3 & 4 & 5 & 6 & 7 (fixed) \\ \hline\hline
$f_0$ & $\emptyset$ & $\emptyset$ & $\emptyset$ & $\emptyset$ & $\emptyset$ &
  $\emptyset$ & $\emptyset$ \\ \hline
$f_1$ & $\emptyset$ & $\{1\}$ & $\{1\}$ & $\{1\}$ & $\{1\}$ & $\{1\}$ &
  $\{1\}$ \\ \hline
$f_2$ & $\emptyset$ & $\{2\}$ & $\{1,2\}$ & $\{1,2\}$ & $\{1,2\}$ &
  $\{1,2\}$ & $\{1,2\}$ \\ \hline
$f_3$ & $\emptyset$ & $\emptyset$ & $\{2\}$ & $\{1,2,5\}$ & $\{1,2,4,5\}$ &
  $\{1,2,4,5\}$ & $\{1,2,4,5\}$ \\ \hline
$f_4$ & $\emptyset$ & $\{4\}$ & $\{4\}$ & $\{2,4\}$ & $\{2,4,5\}$ &
  $\{2,4,5\}$ & $\{2,4,5\}$ \\ \hline
$f_5$ & $\emptyset$ & $\{5\}$ & $\{4,5\}$ & $\{4,5\}$ & $\{4,5\}$ &
  $\{4,5\}$ & $\{4,5\}$ \\ \hline
$f_6$ & $\emptyset$ & $\emptyset$ & $\{5\}$ & $\{4,5\}$ & $\{4,5\}$ &
  $\{4,5\}$ & $\{4,5\}$ \\ \hline
$f_7$ & $\emptyset$ & $\emptyset$ & $\emptyset$ & $\{2\}$ & $\{1,2,5\}$ &
  $\{1,2,4,5\}$ & $\{1,2,4,5\}$ \\ \hline
\end{tabular}
\end{center}
  \caption{The fixed point computation of reaching definitions for example of
    Table~\ref{tab:state_df_eq}.}
  \label{tab:state_df_fp}
\end{table}

There is another requirement to ensure a fixed point exists and that some
data-flow analysis finds one and terminates. The function $F$ has to be
monotone and has to work on an ordered set, lattice in particular.  Since we
have not defined lattices yet, this is described in more detail later in
Section~\ref{sec:ai}. The requirement stems from works of Kleene~\cite[pages
348--349]{kleene} and Tarski~\cite{tarski}.

Now, we generalize the approach of the data-flow analysis. In the data-flow
analysis, we have a pair of equations for each statement $s$ of a program:
\begin{equation} \label{eq:data_flow}
\begin{aligned}
out_s & = trans_s(in_s) \\
in_s & = join_{p\in pred_s}(out_p)
\end{aligned}
\end{equation}
We define a transfer function $trans_s$ to transform the input set $in_s$ of a
statement $s$ to the output set $out_s$. $trans_s$ is defined according to the
effect of $s$ in a particular analysis.  The join operator $join$ takes output
sets from all nodes $pred_s$ and combines them together. $join$ is usually a
set union or intersection depending whether the analysis is may or must (cf.\
Section~\ref{sec:types_anal}). The nodes $pred_s$ are either predecessors or
successors, depending whether the analysis is forward or backward
(Section~\ref{sec:types_anal}).  The direction of the analysis also affects
the position where we put $in_s$ and $out_s$ sets. For a forward analysis, the
$in_s$ set is bound to an incoming control flow graph edge(s) of the node $s$
and the $out_s$ set to the outgoing one(s). It is other way round for a
backward analysis. As we can imagine, $trans_s(in_s)$ is exactly the transfer
equation $f_s$.

Now we can define our example of the reaching definitions above using the
notion of Equation~\ref{eq:data_flow}. The join operator can be defined as the
set union over all incoming sets ($\bigcup$). The transfer function can be set
up to hold line numbers of statements. Therefore, if we let $S$ be the set of
all statements in an analysed program, the transfer function can be as is
depicted in Figure~\ref{fig:st_rdef_trans}.

\begin{figure}[tb]
\begin{align*}
trans_s(in_s) & = gen_s \cup (in_s - kill_s) \\
gen_s & =
\begin{cases}
\{s\}     & \text{$s$ is an assignment} \\
\emptyset & \text{otherwise}
\end{cases} \\
kill_s & =
\begin{cases}
\{x \in S \mid x \text{ assigns to \code{a}} \} & s \text{ is an assignment}
  \\
& \text{(let } s \text{ assign to \code{a})} \\
\emptyset & \text{otherwise}
\end{cases}
\end{align*}

  \caption{The transfer function defined for the reaching definitions
  data-flow analysis.}
  \label{fig:st_rdef_trans}
\end{figure}

Thanks to the simplicity of the data-flow analysis, there are many diverse
data-flow analyses. In Table~\ref{tab:data_flow_examples}, we suggest some
examples of other well-known basic data-flow analyses as they are defined
according to the forward/backward and may/must criteria.
\begin{table}[tb]
\begin{center}
\begin{tabular}{|c||c|c|} \hline
\textbf{Analysis}& \textbf{May}& \textbf{Must} \\ \hline\hline
\multirow{2}{*}{\textbf{Forward}}
  & Reaching    & Available   \\
  & definitions & expressions \\ \hline
\multirow{2}{*}{\textbf{Backward}}
  & Live        & Very busy   \\
  & variables   & expressions \\ \hline
\end{tabular}
\end{center}
    \caption{Examples of data-flow analyses according to may/must and
    forward/backward rules.}
  \label{tab:data_flow_examples}
\end{table}
These are the most known analyses which we believe the reader knows. If this
is not the case, the analyses are defined in every superior literature about
compilers like of Aho et al.~\cite{dragon} or Muchnick~\cite{muchnick}.

The simplicity is also a reason, that the data-flow analysis is used in
many tools including checkers.  The very common place of extensive use are
compilers. Apart from generating binaries, they generate warnings or errors
when some suspicious code is found. A good example is \clang~\cite{clang}. It
is a C/C++ compiler from the \llvm toolkit~\cite{llvm} and it contains also a
static checker framework.  There are checkers for finding errors in memory
handling, locking, or pointer operations. Sadly, the results are mostly
imprecise and contain many false reports.  \textsc{Cppcheck}~\cite{cppcheck}
works on the same principles so it checks for a very similar set of errors.
Earlier, while compilers were still not so evolved, other data-flow analyzers
were used to fulfill this role. One of the most known analyzers at that time
was \textsc{Lint}~\cite{lint}. It was capable of finding errors like division
by zero, constant conditions which always evaluate either to true or false, or
use of uninitialized variables.  Later, \textsc{Lint} resulted in some clones
like \textsc{LCLint}~\cite{lclint} which were augmented to catch errors like
buffer overflows. \textsc{LCLint} was later renamed to \textsc{Splint}.  These
tools mostly have a support for annotations so more speculative checks can be
enabled. In that case, the code has to be heavily annotated, otherwise the
false positive rate is unbearable. Nonetheless, the development of
\textsc{Splint} stagnated over the past decade due to obvious reasons: the
most errors and warnings \textsc{Splint} was used to report are reported by
modern compilers nowadays. It was also thanks to the work of Reps et
al.~\cite{dataflow_large}. They showed that it is possible to compute the
data-flow sets precisely for a whole program inter-procedurally in polynomial
time. They presented their discovery on an analysis that finds possibly
uninitialized variables.  There exist also analyzers that are not so powerful
to handle large code bases or require high user intervention to annotate the
checked code. As an example, we give \textsc{Aspect}~\cite{aspect}, which
compares a program specification of an input-output behaviour provided by a
user to data-flow models extracted from analyzed code.

As we use the data-flow analysis only as a support for more complex analyses,
this section is only a short introduction to the data-flow analysis.  More
information and many examples are available in the literature~\cite{dragon,
muchnick}.



\section{Abstract Interpretation}\label{sec:ai}
The data-flow analysis is only a limited tool. It just collects information
about data and how they propagate through the control flow of a program.
Therefore it is not that useful for designing more complex analyses which are
usual for program analysis and possibly elsewhere. There was a desire to
formalize the process of designing such analyses. Cousots have been dealing
with this for a long time: their work on this topic stretches back to the
1970s. Back then, they described a framework called \emph{abstract
interpretation}~\cite{cousot}. It is designed to allow users of the formalism
to easily design an analysis, prove whether it is correct and sound, to
consistently formulate more precision levels of the analysis and more. The
formalism is so generic, that it allows us to interpret analyses like the
data-flow analysis as a special instance of the abstract interpretation
too~\cite{ai_intro}.

As the name of the analysis suggests, the data domains used in analyzed code
are replaced by abstract ones. The approach requires an abstract domain to be
a \emph{complete lattice}~\cite{lattice1,lattice2}. We formally define it in
Definition~\ref{def:colattice}. Since the analyzed program is interpreted on a
new (abstract) domain, we have to provide a new semantics of the program
instructions over the domain.  There are no limits in theory, the semantics
can be given by arbitrary \emph{monotone functions} over the abstract lattice.
The functions are assigned to each instruction in an analyzed program to
transform the program states over a particular instruction. We then compose
these functions into a single monotone function over the whole program (over a
product lattice as we will see later).  Thanks to the work of
Tarski~\cite{tarski} and Kleene~\cite[pages 348--349]{kleene}, we can always
compute a fixed point of the single monotone function transferring states over
the product lattice.

\begin{dfn}[Complete lattice]\label{def:colattice}
A \emph{complete lattice} $(L,\, \leq,\, \bot,\, \top,\, \sqcup,\, \sqcap)$ is
a partially ordered set $(L,\, \leq)$ where every set $X \subseteq L$ has the
least upper bound (supremum) $\sqcup X$ and the greatest lower bound (infimum)
$\sqcap X$. In particular, $L$ has the least element $\bot = \sqcap L$ and the
greatest element $\top = \sqcup L$.
\end{dfn}

The use of lattice(s) was not a random choice. Lattices are effective, their
join and meet operators can be used when an analyzer is interpreting a
program.  For example, when a branching occurs, the branch join operator can
simply take the values present on the outputs of both branches and performs
the lattice join ($\sqcup$).

Let us demonstrate the abstract interpretation on an example and then define
the approach formally.  We work with the \emph{sign analysis} for
demonstration purposes.  The sign analysis is able to predict a sign of a
variable \code{x} by looking only on signs of constants in an expression
assigned to \code{x}.  Consider for example the expression $\var{y} = -123
\cdot (678 + 5298)$.  Without really looking at the actual numbers, the
analysis can conclude that the value stored in \code{y} is negative (in the
usual understanding of the addition and multiplication operators). Note that
the analysis is not able to answer every such query due to the abstraction.
For example, we cannot decide the sign of \code{z} in the expression $\var{z}
= 3 + (-5)$ when considering only signs (a negative number added to a positive
one can generally give all a negative, positive, or zero result). In our
further explanation, we limit ourselves only to the addition and
multiplication operators for easier understanding. We also consider only a
single variable in a program.

First, in our example, we have to establish an abstract domain which the
abstract interpretation works with. There are five possible values which can
occur in the sign analysis.  Either an expression is positive ($\oplus$),
negative ($\ominus$), zero ($0$), or the result is unknown as for \code{z}
above ($\top$). Finally, there is a value to represent an uninitialized
variable ($\bot$). If it appears in some subexpression after the abstract
interpretation is computed, the analyzer should report an error.  The complete
lattice for the set of these elements $\mathbf{Sign} = \{\oplus,\, \ominus,\,
0,\, \top,\, \bot \}$ is depicted in Figure~\ref{fig:state_ai_sign_lat}.
\begin{figure}[htb]
  \begin{center}
    \begin{tikzpicture}[node distance=5mm and 5mm]
      \node (TOP) {$\top$};
      \node [below right=of TOP] (PLUS) {$\oplus$}
	edge node {} (TOP);
      \node [below=of TOP] (ZERO) {$0$}
	edge node {} (TOP);
      \node [below left=of TOP] (MINUS) {$\ominus$}
	edge node {} (TOP);
      \node [below=of ZERO] (BOT) {$\bot$}
	edge node {} (PLUS)
	edge node {} (ZERO)
	edge node {} (MINUS);
    \end{tikzpicture}
  \end{center}
  \caption{A complete lattice used for the sign analysis.}
  \label{fig:state_ai_sign_lat}
\end{figure}

Next, we need to cope with subexpressions and compute a result of a whole
expression assigned to a variable. For that purpose, we define two binary
(overloaded) functions $+$ and $\cdot$ over abstract values, for addition and
multiplication respectively. Their type is obviously $\mathbf{Sign} \times
\mathbf{Sign} \rightarrow \mathbf{Sign}$.  These two functions are defined in
Table~\ref{tab:state_ai_addmult}.
\begin{table}[tb]
  \begin{center}
    \begin{tabular}{|c||c|c|c|c|c|}\hline
      $+$       & $0$       & $\ominus$ & $\oplus$ & $\bot$    & $\top$ \\ \hline \hline
      $0$       & $0$       & $\ominus$ & $\oplus$ & $0$       & $\top$ \\ \hline
      $\ominus$ & $\ominus$ & $\ominus$ & $\top$   & $\ominus$ & $\top$ \\ \hline
      $\oplus$  & $\oplus$  & $\top$    & $\oplus$ & $\oplus$  & $\top$ \\ \hline
      $\bot$    & $0$       & $\ominus$ & $\oplus$ & $\bot$    & $\top$ \\ \hline
      $\top$    & $\top$    & $\top$    & $\top$   & $\top$    & $\top$ \\ \hline
    \end{tabular} ~~
    \begin{tabular}{|c||c|c|c|c|c|}\hline
      $\cdot$   & $0$ & $\ominus$ & $\oplus$  & $\bot$    & $\top$ \\ \hline \hline
      $0$       & $0$ & $0$       & $0$       & $0$       & $0$    \\ \hline
      $\ominus$ & $0$ & $\oplus$  & $\ominus$ & $\ominus$ & $\top$ \\ \hline
      $\oplus$  & $0$ & $\ominus$ & $\oplus$  & $\oplus$  & $\top$ \\ \hline
      $\bot$    & $0$ & $\ominus$ & $\oplus$  & $\bot$    & $\top$ \\ \hline
      $\top$    & $0$ & $\top$    & $\top$    & $\top$    & $\top$ \\ \hline
    \end{tabular}
  \end{center}
  \caption{The behaviour of arithmetic addition and multiplication for the
  sign analysis.}
  \label{tab:state_ai_addmult}
\end{table}
Recalling the example above, $\var{y} = -123 \cdot (678 + 5298)$ is abstracted
to $\var{y} = \ominus \cdot (\oplus + \oplus)$ and using the two functions, we
can evaluate the sign of \code{y} to $\ominus$.

We know how the values are computed in an expression, but we still need to
define a \emph{transfer function} over lines of code. Similar to the data-flow
analysis, the transfer function $f_s$ for a statement $s$ takes the states
incoming to $s$ and changes them according to the effect of $s$. The type of
$f$ is $f^n \rightarrow \mathbf{Sign}$, where $n$ is the number of lines. In
the following text, we demonstrate the interpretation on example code from the
second column of Table~\ref{tab:state_ai_sign_ex}.
\begin{table}[tb]
  \begin{center}
  \begin{tabular}{|r|l||p{2.4em}l|} \hline
\textbf{\footnotesize \#} &
  \multicolumn{1}{c||}{\textbf{Code}} &
  \multicolumn{2}{c|}{\textbf{Transfer Function}} \\ \hline\hline
1 & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
void fun(int n) {
\end{lstlisting} } &
  $f_1\left(\overrightarrow{f}\right)$ & $= \bot$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
  int x = 0;(*@ \label{lst:state_ai_x0} @*)
\end{lstlisting} } &
  $f_2\left(\overrightarrow{f}\right)$ & $= 0$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
  x = x + 1;(*@ \label{lst:state_ai_xp1} @*)
\end{lstlisting} } &
  $f_3\left(\overrightarrow{f}\right)$ & $= f_2\left(\overrightarrow{f}\right) + \oplus$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
  while (--n)
\end{lstlisting} } &
  $f_4\left(\overrightarrow{f}\right)$ & $= f_3\left(\overrightarrow{f}\right) \sqcup
    f_5\left(\overrightarrow{f}\right)$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
    x = x * 5;(*@ \label{lst:state_ai_xm} @*)
\end{lstlisting} } &
  $f_5\left(\overrightarrow{f}\right)$ & $= f_4\left(\overrightarrow{f}\right) \cdot \oplus$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
  x = x + (-1);(*@ \label{lst:state_ai_xp2} @*)
\end{lstlisting} } &
  $f_6\left(\overrightarrow{f}\right)$ & $= f_4\left(\overrightarrow{f}\right) + \ominus$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_ai_comp,numbers=none,frame=none]
}
\end{lstlisting} } &
  $f_7\left(\overrightarrow{f}\right)$ & $= f_6\left(\overrightarrow{f}\right)$ \\ \hline

\end{tabular}
  \end{center}
  \caption{Example to demonstrate the sign abstract interpretation. We only
  track the sign of the variable \code{x}.  $\protect\overrightarrow{f}$ is
  the tuple $f_1,\, \dots,\, f_7$.}
  \label{tab:state_ai_sign_ex}
\end{table}
In the example, we want to track the sign of the variable \code{x} using the
abstract interpretation. Transfer functions for this interpretation are
shown in the third column for each of the lines. Note that we denote the tuple
of all $f_s$ as $\overrightarrow{f}$.

As we mentioned earlier, a single monotone function is needed for the fixed
point calculation. The function is composed of the transfer functions and it
works on a whole program using a \emph{product lattice}. A product lattice is
the complete lattice $(L^n,\, \sqsubseteq)$, where $(L,\, \leq)$ is a complete
lattice and
\begin{equation*}
\left(\left(x_1,\, \dots,\, x_n\right) \sqsubseteq \left(y_1,\, \dots,\,
  y_n\right)\right) \Leftrightarrow
  \left(x_1 \leq y_1 \wedge \dots \wedge x_n \leq y_n\right).
\end{equation*}
Given the product lattice $(L^n,\, \sqsubseteq)$, we define the composed
function $F: L^n \rightarrow L^n$ of monotone functions $f_1,\, \dots,\, f_n$
as
\begin{equation*}
F\left(\overrightarrow{f}\right) =
  \left(f_1\left(\overrightarrow{f}\right),\, \dots,\,
  f_n\left(\overrightarrow{f}\right)\right).
\end{equation*}
Note that $F$ is also a monotone function in $(L^n,\, \sqsubseteq)$. Thanks to
that, the least fixed point exists~\cite{kleene, tarski}.  We iteratively
compute the least fixed point in Table~\ref{tab:state_ai_sign_fp} for our
example of Table~\ref{tab:state_ai_sign_ex}.
\begin{table}[tb]
  \begin{center}
\begin{tabular}{|c|c||c|c|c|c|c|c|} \hline
\multicolumn{2}{|c||}{\textbf{Transfer}} &
  \multicolumn{6}{c|}{\textbf{Iterations of Fixed Point Computation}}
  \\ \cline{3-8}
\multicolumn{2}{|c||}{\textbf{Function}} &
  ~~~0~~~ & ~~~1~~~ & ~~~2~~~ & ~~~3~~~ & ~~~4~~~ & 5 (fixed) \\ \hline\hline
\multirow{7}{*}{$F$} &
  $f_1$ & $\bot$ & $\bot$   & $\bot$   & $\bot$   & $\bot$   & $\bot$ \\
  \cline{2-8}
& $f_2$ & $\bot$ & $0$      & $0$      & $0$      & $0$      & $0$ \\
  \cline{2-8}
& $f_3$ & $\bot$ & $\oplus$ & $\oplus$ & $\oplus$ & $\oplus$ & $\oplus$ \\
  \cline{2-8}
& $f_4$ & $\bot$ & $\bot$   & $\oplus$ & $\oplus$ & $\oplus$ & $\oplus$ \\
  \cline{2-8}
& $f_5$ & $\bot$ & $\bot$   & $\bot$   & $\oplus$ & $\oplus$ & $\oplus$ \\
  \cline{2-8}
& $f_6$ & $\bot$ & $\bot$   & $\bot$   & $\top$   & $\top$   & $\top$ \\
  \cline{2-8}
& $f_7$ & $\bot$ & $\bot$   & $\bot$   & $\bot$   & $\top$   & $\top$ \\
  \hline
\end{tabular}
  \end{center}
  \caption{Fixed point computation for example of
  Table~\ref{tab:state_ai_sign_ex}.}
  \label{tab:state_ai_sign_fp}
\end{table}
We can see that the analysis can definitely decide about the sign of a value
stored in \code{x} until the assignment on line~\ref{lst:state_ai_xp2}. After
that line, we lose precision due to the abstraction.

As we proceeded with the example, we informally defined the abstract
interpretation by:
\begin{enumerate}
\item A complete lattice $(L,\, \leq,\, \bot,\, \top,\, \sqcup,\, \sqcap)$
which describes abstract values.

\item Operations over the lattice definining the behaviour on subexpressions.

\item A transfer monotone function $F$ which defines the interpretation
of a program.

\end{enumerate}
Formally, the abstract interpretation is the tuple $I=\langle L,\, \circ,\,
\leq,\, \top,\, \bot,\, F\rangle$. $\circ$ is either the supremum ($\sqcup$)
or the infimum ($\sqcap$) lattice operator depending whether the results
should be the least or the greatest fixed point, respectively.

The abstract interpretation is a very flexible tool for designing analyses. It
does not require the lattices to have a finite height. Even if the lattice is
infinitely high, a fixed point still can be computed in a finite time (but
need not be the least).  Consider the lattice of
Figure~\ref{fig:state_lat_int}. It is used for abstracting values of integer
variables in programs.  The lattice is composed of intervals to remember
possible bounds of some variable's value.  This is a very useful domain
especially when analyzing indices of arrays to ensure they never exceed the
length of the indexed array.
\begin{figure}[tb]
\begin{center}
  \tikzstyle{invisible} = [minimum size=6mm]
  \begin{tikzpicture}
    \node (N_7) at (0,1) {$[-\infty,\infty]$};

    \node[invisible] (N_6X) at (-3,0) {}
      edge[dashed] node {} (N_7);
    \node[invisible] (N_6A) at (-2,0) {}
      edge[dashed] node {} (N_7);
    \node[invisible] (N_6B) at (-1,0) {}
      edge[dashed] node {} (N_7);
    \node[invisible] (N_6C) at (1,0) {}
      edge[dashed] node {} (N_7);
    \node[invisible] (N_6D) at (2,0) {}
      edge[dashed] node {} (N_7);
    \node[invisible] (N_6Y) at (3,0) {}
      edge[dashed] node {} (N_7);

    \node (N_5) at (0,-1) {$[-2,2]$}
      edge[dashed] node {} (N_6B)
      edge[dashed] node {} (N_6C);

    \node[invisible] (N_5X) at (-2,-1) {};
    \node[invisible] (N_5Y) at (2,-1) {};

    \node (N_4A) at (-1,-2) {$[-2,1]$}
      edge node {} (N_5)
      edge[dashed] node {} (N_5X);
    \node (N_4B) at (1,-2) {$[-1,2]$}
      edge node {} (N_5)
      edge[dashed] node {} (N_5Y);

    \node[invisible] (N_4X) at (-3,-2) {};
    \node[invisible] (N_4Y) at (3,-2) {};

    \node (N_3A) at (-2,-3) {$[-2,0]$}
      edge node {} (N_4A)
      edge[dashed] node {} (N_4X);
    \node (N_3B) at (0,-3) {$[-1,1]$}
      edge node {} (N_4A)
      edge node {} (N_4B);
    \node (N_3C) at (2,-3) {$[0,2]$}
      edge node {} (N_4B)
      edge[dashed] node {} (N_4Y);

    \node[invisible] (N_3X) at (-4,-3) {};
    \node[invisible] (N_3Y) at (4,-3) {};

    \node (N_2A) at (-3,-4) {$[-2,-1]$}
      edge node {} (N_3A)
      edge[dashed] node {} (N_3X);
    \node (N_2B) at (-1,-4) {$[-1,0]$}
      edge node {} (N_3A)
      edge node {} (N_3B);
    \node (N_2C) at (1,-4) {$[0,1]$}
      edge node {} (N_3B)
      edge node {} (N_3C);
    \node (N_2D) at (3,-4) {$[1,2]$}
      edge node {} (N_3C)
      edge[dashed] node {} (N_3Y);

    \node[invisible] (N_2X) at (-5,-4) {};
    \node[invisible] (N_2Y) at (5,-4) {};

    \node (N_1A) at (-4,-5) {$[-2,-2]$}
      edge node {} (N_2A)
      edge[dashed] node {} (N_2X);
    \node (N_1B) at (-2,-5) {$[-1,-1]$}
      edge node {} (N_2A)
      edge node {} (N_2B);
    \node (N_1C) at (0,-5) {$[0,0]$}
      edge node {} (N_2B)
      edge node {} (N_2C);
    \node (N_1D) at (2,-5) {$[1,1]$}
      edge node {} (N_2C)
      edge node {} (N_2D);
    \node (N_1E) at (4,-5) {$[2,2]$}
      edge node {} (N_2D)
      edge[dashed] node {} (N_2Y);

    \node [invisible,left=of N_1A] (N_1X) {};
    \node [invisible,right=of N_1E] (N_1Y) {};

    \node (BOT) at (0,-7) {$\bot$}
      edge[dashed] node {} (N_1X)
      edge node {} (N_1A)
      edge node {} (N_1B)
      edge node {} (N_1C)
      edge node {} (N_1D)
      edge node {} (N_1E)
      edge[dashed] node {} (N_1Y);
  \end{tikzpicture}
\end{center}
\label{fig:state_lat_int}
\caption{Lattice for intervals.}
\end{figure}

Obviously, the Kleene's sequence is not necessarily finite for this lattice.
Therefore, the computation of a fixed point may not finish. To overcome the
problem, Cousots~\cite{cousot} introduced the \emph{widening} operator
$\triangledown$. Informally, the sequence is sped up by widening, so that it
converges eventually.  Imagine that we have an interval lattice and an
infinite loop where we increment a variable that was initially set to zero.
The infinite sequence of intervals computed for that variable would be:
$[0,0],\, [0,1],\, \dots,\, [0,\infty]$. We can define the widening operator
to set one of the interval bounds to infinity after some time (e.g.\ 100
iterations) to force the sequence to be finite like: $[0,0],\, [0,1],\,
\dots,\, [0, 99],\, [0,\infty]$. This makes the Kleene's sequence finite.
There exists also the \emph{narrowing} operator $\vartriangle$ which is used
to make abstract values more precise. The reader is pointed to Cortesi and
Zanioli~\cite{ai_widening} or Cousots~\cite{cousot_galois} who explain
widening and narrowing formally and demonstrate the operators on more extended
examples.

Intervals are not the only shapes to represent abstract values of numbers.
Octagons, elipses, polyhedra, and other shapes are used to represent relations
between valuse of two or more variables~\cite{ai_octagons, ai_polyhedra,
astree}. Several abstractions can be also combined to communicate with each
other and improve thus accuracy or any other aspect of the
analysis~\cite{ai_product}. This was utilized in tools like
\textsc{Astree}~\cite{astree} and \textsc{Clousot}~\cite{clousot}.
Strings~\cite{ai_strings1, ai_strings2} and foating point
numbers~\cite{ai_float1} can be abstracted too.  Further reading on the
abstract interpretation is an overview~\cite{cousot_progress} describing the
progress in the area of the abstract interpretation over 25 years. A not so
usual application of the abstract interpretation can be found
too~\cite{ai_unusual1, ai_unusual2, ai_unusual3}.  Engler et al.\ brought the
abstract interpretation to the attention in 2000 by a technique based on a
meta compilation in \xgcc~\cite{metacomp}. There, program developers specify
properties of the checked system in \textsc{Metal} language~\cite{metal},
compile it into an internal representation and the tool applies the
specification on all executable paths in a particular function.  Violation of
the specification on a path is then reported as an error.  \xgcc suffers from
false positives despite the use of false positive suppression algorithms like
killing variables and expressions, synonyms, false path pruning, and others.
Besides the suppression algorithms, error reports from the tool are further
ranked according to their probability of being real errors. There are generic
and statistical ranking algorithms ordering error reports.  An extension
introduced later~\cite{lockchecker} provides an automatic inference of some
temporal properties based on statistical analysis of assumed programmer's
beliefs.  \stan~\cite{stanse} is built on a similar base as \xgcc. We discuss
\stan in Chapter~\ref{sec:stanse}.  It is uncertain whether some industrial
tool adopts the abstract interpretation too, see the discussion in
Section~\ref{sec:industrial}.

The correctness and soundness (cf.\ Section~\ref{sec:theor_props}) of an
analyzer based on the abstract interpretation is very prone to selection of
the abstract domain. While many employed abstractions are sound (produce all
possible errors), they are not so often complete (false positives occur). For
example, this was demonstrated on several abstractions of numbers which we
discussed above~\cite{ai_octagons}. It is important to note that a properly
selected abstract domain, possibly combined with the widening, ensures that
the analysis always terminates.




\section{Symbolic Execution}\label{sec:se}
In the previous section we described the abstract interpretation which is a
framework for fast, but typically imprecise analyses. As a result of sound,
but incomplete abstract analyzers, we often get a pile of false positives on
their output.  There also exist techniques which cannot formally generate
false positives, but are often slower than abstract interpreters or sometimes
even unsound.  One of precise and sound techniques is the \emph{symbolic
execution}. The technique was introduced in 1970's and described by several
authors concurrently~\cite{se_king, se_select, se_dissect, se_clarke}. The
symbolic execution represents program inputs by symbols.  The program is then
executed by an executor which manipulates expressions over symbols rather than
concrete values. Hence it does not need to execute the program many times with
concrete, initialized, data.  Let us demonstrate the symbolic execution on an
example depicted in Figure~\ref{fig:se_example}.
\begin{figure}[htb]
\begin{center}
\begin{tabular}{c}
\begin{lstlisting}
int fun(void) {
  int a = read() + 1; (*@\label{line:se_ex_sset}@*)
  int b = 0;

  if (a == 0x12345678) (*@\label{line:se_ex_if1}@*)
    b = 1; (*@\label{line:se_ex_set}@*)

  if (a == 0x12345678) (*@\label{line:se_ex_if2}@*)
    a = 10 / b; (*@\label{line:se_ex_div}@*)

  return a;
}
\end{lstlisting}
\end{tabular}
\end{center}
  \caption{Example to demonstrate the symbolic execution.}
  \label{fig:se_example}
\end{figure}

The example is two-fold. First, there is a read from a user. It would take a
long time to exercise the program to get to the true branch of the \code{if}s
with dynamic testing.  With \code{int} being usual 32 bits, there is only one
successful attempt out of $2^{32}$ overall. Of course, with a static analyzer
based on the data-flow analysis or abstract interpretation, those true
branches are usually analyzed quite quickly. But this brings us to the second
point. Many complete analyzers generate a report about division by zero
possible on line~\ref{line:se_ex_div}. This is indeed a false positive because
\code{b} is set to \code{1} on line~\ref{line:se_ex_set} in case \code{a} is
\code{0x12345678}. The source of the false positive is that the analyzers do
not often remember conditions on branchings.

The symbolic execution tackles these two problems. All inputs from the user
are replaced by \emph{symbols}. For a variable \code{a}, we denote the symbol
corresponding to \code{a} as $\sym{a}$. We allow symbols to be assigned to
variables, similar to constants and other expressions. In our example, the
assignment on line~\ref{line:se_ex_sset} would set \code{a} to $\sym{a}+1$.
Obviously, we have to teach the symbolic executor to deal with symbols in
arithmetic operations and branching statements.

A symbolic executor, to be fully working, has to remember a \emph{program
state}. The executor represents the state as a triple and updates its members
while executing a program. The triple $(l,\, \theta,\, \varphi)$ contains:
\begin{itemize}
\item \emph{Location} $l$: the symbolic executor needs to remember the
currently evaluated program location in a control flow graph.

\item \emph{Symbolic memory} $\theta$: the executor needs a storage. The
memory $\theta$ is a function that assigns to each variable its current
symbolic value. So after evaluating the statement on
line~\ref{line:se_ex_sset}, $\theta(\var{a}) = \sym{a}+1$.

\item \emph{Path condition} $\varphi$: to be precise in the execution of code,
the executor also has to remember the branches which it chose to jump to at
branching points. The path condition is a first-order formula~\cite{fol}
representing a necessary and sufficient condition over symbols to drive the
execution along the currently executed path to the actually evaluated point.
$\varphi$ is set initially to $\true$ and is evaluated and potentially updated
at every branch.

\end{itemize}

In our example, the path condition is used and possibly updated on
lines~\ref{line:se_ex_if1} and~\ref{line:se_ex_if2}.  On
line~\ref{line:se_ex_if1}, we check whether $\varphi$ implies validity of the
condition instantiated by the current state of $\theta$. More concretely, the
executor checks whether either of the following two implications hold (at most
one can~\cite{se_king}):
\begin{align}
\varphi & \rightarrow \left( \theta(\var{a}) = \var{0x12345678} \right) \label{equ:se_fi} \\
\varphi & \rightarrow \left( \theta(\var{a}) \neq \var{0x12345678} \right) \label{equ:se_se}
\end{align}
There are three possibilities that can happen:
\begin{itemize}
\item \textbf{Implication \ref{equ:se_fi} is valid:} the executor simply
continues execution by jumping directly to the \emph{true} branch.

\item \textbf{Implication \ref{equ:se_se} is valid:} The executor simply
continues execution by jumping directly to the \emph{false} branch.

\item \textbf{Neither is valid:} This means that both branches are possible,
depending on the symbolic input.  So to retain soundness, the symbolic
executor has to fork the execution of the whole program to explore both
branches. In that case, the path condition $\varphi$ has to be augmented in
the respective branches. Since we have $\theta(\var{a})=\sym{a}+1$, a new path
condition in the true branch is $\varphi \wedge (\sym{a}+\var{1} =
\var{0x12345678})$, for the other branch $\varphi \wedge (\sym{a}+\var{1} \neq
\var{0x12345678})$.

\end{itemize}
In our example, on line~\ref{line:se_ex_if1} neither implication is valid, as
$\theta(\var{a})=\sym{a}+1$ cannot prove validity of $\theta(\var{a}) =
\var{0x12345678}$ or $\theta(\var{a}) \neq \var{0x12345678}$. So the execution
forks.  The situation is different on line~\ref{line:se_ex_if2}, nonetheless.
There, we can get into two states, by the two previous forked executions. In
one, the path condition contains $\theta(\var{a}) = \var{0x12345678}$ and
symbolic memory assigns \code{1} to \code{b}. In the other $\theta(\var{a})
\neq \var{0x12345678}$ and $\theta(\var{b}) = 0$. From the first one, the
executor can induce Implication~\ref{equ:se_fi}, so it continues by jumping
into the true branch.  And since $\theta(\var{b}) = 1$, no division by zero
can happen.  Similarly for the second state: the executor jumps to the false
branch of the \code{if}, i.e.~it skips the \code{if} completely. Note that
this is one of the very cases where other types of analyses including the
abstract interpretation produce false positives.

Due to the forks, the execution of a program is usually depicted as a
\emph{symbolic execution tree}. Nodes of the tree are labeled by program
states computed during the execution.  Edges of the tree correspond to
transitions between program states labeling their end nodes.  For
demonstration, we use the code from the data-flow example as was in the second
column of Table~\ref{tab:state_df_eq}. A control flow graph of the example is
shown in Figure~\ref{fig:state_se_cfg}.
\begin{figure}[htb]
\begin{center}
    \pgfdeclarelayer{background}
    \pgfsetlayers{background,main}
    \begin{tikzpicture}[node distance=5mm,font=\small]
      \node [cfg_start] (S) {};
      \node [cfg] (ASS1) [label=right:\red{1},below=3mm of S] {\code{a = 1}}
	edge[cfg_edge] node {} (S);
      \node [cfg] (ASS2) [label=right:\red{2},below=of ASS1] {\code{b = read()}}
	edge[cfg_edge] node {} (ASS1);
      \node [cfg] (LOOPH) [label=right:\red{3},below=of ASS2] {\code{b}}
	edge[cfg_edge] node {} (ASS2);
      \node [cfg] (ASS3) [label=right:\red{4},below=8mm of LOOPH,xshift=-3mm]
        {\code{a = a + 5}}
	edge [cfg_edge, bend right=10] node [label=right:\code{b}] {} (LOOPH);
      \node [cfg] (ASS4) [label=left:\red{5},below left=of ASS3]
        {\code{b = b - 1}}
	edge [cfg_edge, bend right=25] node {} (ASS3)
	edge [cfg_edge, ->,bend left=40] node {} (LOOPH);
      \node [cfg] (USE) [label=right:\red{6},right=8mm of ASS3]
	{\code{use(a)}}
	edge [cfg_edge, bend right=15] node [label=above right:\code{!b}] {} (LOOPH);
      \node [cfg_end,below=3mm of USE] {}
	edge[cfg_edge] node {} (USE);

      \begin{pgfonlayer}{background}
	\node[fill=blue!20, inner sep=-1pt, rotate=35, xshift=-2mm,
	  yshift=-4mm, ellipse, fit=(LOOPH) (ASS3) (ASS4)] {};
      \end{pgfonlayer}

    \end{tikzpicture}
\end{center}
  \caption{A control flow graph of code from Table~\ref{tab:state_df_eq}. The
  loop is highlighted.}
  \label{fig:state_se_cfg}
\end{figure}
A symbolic execution tree of the control flow graph is depicted in
Figure~\ref{fig:state_se_tree}.
\begin{figure}[htb]
\begin{center}
    \tikzstyle{loc} = [thick,draw,minimum size=5mm]
    \tikzstyle{pre} = [<-,shorten <=1pt,>=stealth',semithick]
    \tikzstyle{symmem}=[draw, loosely dashed, rectangle, inner sep=8pt,
      rounded corners=12pt]
    \tikzstyle{symlink}=[-, dashed, in=-110, out=0]
    \tikzstyle{pathcond}=[symmem, densely dotted]
    \tikzstyle{pathlink}=[symlink, densely dotted]
    \pgfdeclarelayer{background}
    \pgfsetlayers{background,main}
    \begin{tikzpicture}[->, >=stealth', circle, node distance=5mm, font=\small]
      \node [cfg_start] (S) {}
        child { node [loc,below=4mm of S] (L_1) {\red{1}}
	  child { node [loc, below=of L_1] (L_2) {\red{2}}
	    child { node [loc,below=of L_2] (L_3) {\red{3}}
	      child { node [loc,below=of L_3] (L_4) {\red{4}}
		child { node [loc,below=of L_4] (L_5) {\red{5}}
		  child { node [loc,below=of L_5] (L_6) {\red{3}}
		    child { node [loc,below=of L_6]
			   (L_7) {\red{4}}
		      child { node [inner sep=0pt, below=of L_7] (L_8) {$\vdots$} }
		      edge from parent node[left] {$\sym{b}-\var{1} \neq \var{0}$}
		    }
		    child { node [loc,right=of L_7] (F_2) {\red{6}}
		      child { node [cfg_end,below=4mm of F_2] {} }
		      edge from parent node[right,yshift=2mm] {$\sym{b}-1=0$}
		    }
		  }
		}
		edge from parent node[left] {$\sym{b} \neq \var{0}$}
	      }
	      child { node [loc,right=of L_4] (F_1) {\red{6}}
		child { node [cfg_end,below=4mm of F_1] {} }
		edge from parent node[right,yshift=2mm] {$\sym{b}=0$}
	      }
	    }
	  }
	};

      \node [symmem, left=2cm of L_2] (MEM1) {
        \begin{tabular}{rl}
	  $\theta(\var{a})=$ & $\var{1}$ \\
	  $\theta(\var{b})=$ & $\sym{b}$ \\
        \end{tabular}
      } edge [symlink] node {} (L_2);

      \node [symmem, left=1.5cm of L_5] (MEM2) {
        \begin{tabular}{rl}
	  $\theta(\var{a})=$ & $\var{6}$ \\
	  $\theta(\var{b})=$ & $\sym{b} - 1$ \\
        \end{tabular}
      } edge [symlink] node {} (L_5);

      \node [pathcond, yshift=4mm, left=15mm of L_4] (PC1) {
	$\varphi:\,\sym{b}\neq\var{0}$
      } edge [pathlink, in=180, out=0] node {} (L_4);

      \node [pathcond, yshift=-4mm, right=5mm of F_1] (PC2) {
	$\varphi:\,\sym{b}=\var{0}$
      } edge [pathlink, in=0, out=180] node {} (F_1);

      \node [pathcond, yshift=-4mm, left=5mm of L_7] (PC3) {
	$\varphi:\,(\sym{b} \neq \var{0}) \wedge (\sym{b}-\var{1} \neq \var{0})$
      } edge [pathlink, in=180, out=0] node {} (L_7);

      \node [pathcond, yshift=-4mm, right=5mm of F_2] (PC4) {
	$\varphi:\,(\sym{b} \neq \var{0}) \wedge (\sym{b}-\var{1} = \var{0})$
      } edge [pathlink, in=0, out=180] node {} (F_2);

      \begin{pgfonlayer}{background}
	\node[fill=blue!20, rounded corners=12pt, rectangle, inner sep=4pt,
		fit=(L_3) (L_4) (L_5) (L_6) (L_7) (L_8)] {};
      \end{pgfonlayer}

    \end{tikzpicture}
\end{center}
  \caption{A symbolic execution tree of the control flow graph of
  Figure~\ref{fig:state_se_cfg}. The loop remained highlighted. Selected nodes
  are depicted along with the symbolic memory (dashed; after assignments) and
  path conditions (dotted; after branchings).}
  \label{fig:state_se_tree}
\end{figure}
For readability of figures, nodes are not labeled by full program states, but
only by numbers representing the current location in the control flow graph.
In addition, we label
branching edges by instances of the corresponding branching conditions in the
flow graph. These labels allow us to reconstruct the path condition for each
node in the tree: it is the conjunction of labels of all edges along the path
from the root to the particular node. We present them on selected nodes in the
picture as well as the contents of the symbolic memory.

We have not discussed performance of the symbolic execution yet. There are
two main sources of potential performance bottlenecks. First, symbolic
executors use \emph{Satisfiability Modulo Theory} solvers~\cite{smt} to reason
about validity of formulae like~\ref{equ:se_fi} and \ref{equ:se_se}.  Whereas
performance of individual solvers vary, complex queries are not easy to answer
for any solver. So it usually takes a long time for solvers before they
answer such queries. And worse, some queries can be also out of decidable
theories.  Second, the symbolic execution suffers from the \emph{path
explosion problem}, i.e.\ an exponential growth of paths that need to be
explored. Every branching in the code potentially multiplies the possible
number of paths through the program by two. It is even worse for loops, where
the symbolic execution tree can become huge or infinite.  Hence, for
real-world programs, the symbolic execution is barely usable without
incorporating other techniques to reduce the count of explored paths.  Some
tools try to tackle this issue by traversing more paths at once to
substantially reduce the size of the execution tree~\cite{cse_atva}.
Others~\cite{pex_ddse, G07}, compose symbolic executions of several functions
using summaries which results in an inter-procedural analysis. This approach
limits the need of execution to separate functions only.
\textsc{Exe}~\cite{exe}, \textsc{Klee}~\cite{klee}, and
\textsc{RWset}~\cite{rwset} are based on recording of already seen behavior
and pruning its repetition. Tools like Girgis'~\cite{se_limited} and
\ucklee~\cite{ucklee} also limit the tree depth by a hard-coded constant, to
obtain a finite tree for the price of unsoundness.  Some works aim at guessing
the number of loops needed to reach some target location to help analyzing
large programs with many nested loops~\cite{partial_loop}.

Despite King originally introduced the symbolic execution for symbolic
debugging~\cite{se_king}, the symbolic execution is currently widely used for
generating unit tests which cover more paths in code than tests written
manually~\cite{se_generate_test1, se_generate_test2, se_generate_test3,
se_generate_test4, se_generate_test5, exe, klee}. Offutt and
Seaman~\cite{se_generate_test2} use the symbolic execution also for re-writing
program internal variables as functions of the input values and then generate
tests using constraint-solving. The symbolic execution is usable as a partial
correctness prover~\cite{se_corr1, se_corr2}, as a reverse engineering
engine~\cite{se_reverse}, or for refactoring software to a one which is more
specialized~\cite{se_spec}. It can be extended to handle heap~\cite{se_heap,
se_gener}, summarize effects of libraries~\cite{se_libs}, or to extract
dependencies of outputs on inputs~\cite{se_output1}. It was used to find bugs
in aircraft control and nuclear power systems~\cite{se_critical} or
filesystems by generating malicious disks~\cite{exe_fs}. Even though, some do
not explicitly mention the symbolic execution, they in fact implement it, for
example to find errors in memory management~\cite{prefix}.
\textsc{PathFinder-SE}~\cite{pathfinder_se} is an extension to the model
checker \textsc{PathFinder} which enables the model checker to perform the
symbolic execution.  They further improved the tool to handle e.g.\ concurrent
programs, heap structures and data.  \textsc{IntScope}~\cite{intscope}
decompiles x86 binaries to an intermediate representation form and runs the
symbolic execution to find integer overflows.
%
%\cite{active_props}

\subsection{Concolic Execution}
Many modern tools combine the symbolic execution with dynamic (concrete)
analysis. This combination of analyses is called \emph{concolic
execution}. Tools implementing concolic execution achieve better scalability
and higher coverage of analyzed code. \textsc{PathCrawler}~\cite{pathcrawler}
was the first executor to combine symbolic and dynamic execution. It does not
use concrete values in the symbolic execution to simplify complex constraints
nonetheless. This was first performed only by \textsc{Dart}~\cite{dart}.  The
primary goal is to execute \emph{all}, or at least as much as possible
feasible program paths, in a systematic way. It starts dynamically with random
input and symbolically tracks branching conditions.  The symbolic execution
and a solver is then used to change the input so that it traverses another
paths when executed dynamically again. This happens until the code is covered
fully or an user stops the process.  \textsc{Dart} can propagate concrete
values from the dynamic runs to the symbolic run when the solver is incapable
to reason about some values. Especially, when pointer constraints are used,
\textsc{Dart} gives up the symbolic execution and switches to standard random
testing. This was improved in another executor \textsc{Cute}~\cite{cute}. It
handles pointers better and is able to analyze concurrent programs. So except
it generates the inputs systematically, it also makes diverse schedule plans
of individually spawn threads.  \textsc{Cute} cannot still handle symbolic
offsets of pointers. Very close tool to \textsc{Cute} is
\textsc{Crest}~\cite{crest}, except it is freely available, source
codes inclusive. Since every concolic execution tool is prone to selected
strategy that decides which states are explored next, the authors of
\textsc{Crest} studied and compared the strategies too.  \textsc{Check 'n'
Crash}~\cite{check_crash} does quite the opposite to \textsc{Cute}. First, it
runs a coarse symbolic execution to find a suspicious location in code. Then
the tool generates a test case and runs the program dynamically to verify
whether the code location really contains an error.

\paragraph{\textsc{Exe} and Derivatives}
At the same time as \textsc{Dart} and \textsc{Cute} emerged, Cadar et al.\
came up with \textsc{Egt}~\cite{egt} for generating test cases from symbolic
inputs.  \textsc{Egt} was a predecessor of \textsc{Exe}~\cite{exe} which
finally introduced a precise handling of pointers with symbolic offsets.
\textsc{Exe} was later completely redesigned and the new tool
\textsc{Klee}~\cite{klee} was put to execute \llvm code. \textsc{Klee}
optimizes solver queries, has better search heuristics, and other improvements
like a model to simulate a filesystem without really involving it.  This
allowed for a successful error finding in \textsc{Coreutils}, the basic set of
utilities most Linux users work with. \textsc{Klee} can be used also for
checking a tool equivalence. It takes two implementations of the same
functionality and symbolically compares the results. Several deviations from
the standard behaviour of some utilities were found that way. This was later
improved more in \ucklee~\cite{ucklee}. First, it compares memory contents of
both implementations. Second, it makes the memory symbolic when accessed
(on-demand). Finally, it checks that both implementations behave the same with
respect to error conditions like division by zero. \textsc{Klee} and
\textsc{UC-Klee} are put into concolic executors since they allow to invoke
library functions dynamically.

Instead of generating only a single input, \sage~\cite{active_props, sage}
defines the term \emph{generational search} to generate as many inputs as
possible in a single symbolic run. This is unimportant for cases, where the
whole state space is explored. But this is not the case of real programs as
they are barely covered fully because the symbolic execution is very
expensive. So generating more inputs often helps more coverage. As it is a
concolic executor, it records branching conditions from the first dynamic run.
But it negates \emph{all} of them one by one and asks a solver for a solution.
For each positive reply the solver gives, \sage generates a separate input.
The whole tool does not use any intermediate language, it symbolically
executes directly the x86 binary code. So it can be used also in cases where
source code is not known.  The whole approach of \sage is called
\emph{automated whitebox fuzz testing}.  \textsc{SmartFuzz}~\cite{smartfuzz}
works very similarly, but the set of errors the tool aims at is limited only
to integer errors. That is, \textsc{SmartFuzz} captures signed/unsigned
conversions for example. The generated inputs are filtered, so more eager
heuristics that allow more inputs generation are allowed.

None of the tools above supports automatic bi-directional conversion across
the boundary of the symbolic and concrete execution. This was one of the two
novelties in \textsc{S${}^2$e}~\cite{s2e}: \emph{selective symbolic
execution}. It was published by some of the authors earlier to solve the
problem of path explosion in libraries and system calls~\cite{se_selective}.
As we are mainly interested only in some parts of the analyzed code, the
symbolic execution does not fork in, let us say, libraries. Only when the
execution gets back to our code from the libraries, it can fork again.  The
second novelty is a true concolic execution because it runs the \textsc{Klee}
symbolic executor and a modified \textsc{Qemu} virtual machine in parallel.
Those two communicate with each other and exchange information, especially the
symbolic state is updated from the dynamic execution and concrete values
instantiated from the symbolic run on the fly. Converters in both directions
from x86 code to \llvm have to be in place to have both executions running.
\textsc{S${}^2$e} was used to find errors in Windows network drivers, and
allegedly it was the first one to employ the symbolic execution for
performance analysis too.  \textsc{S${}^2$e}'s parallel engine works on the
top of \textsc{Ddt${}^+$}, which is a reimplementation of a closed source tool
\textsc{Ddt}~\cite{ddt}. A very similar to \textsc{S${}^2$e}, but less generic
framework is \textsc{BitBlaze}~\cite{bitblaze}. It analyses binary files and
creates intermediate code with a reduced instruction set. The instructions are
executed symbolically.  One of unusual uses of the concolic execution is
generating exploits by a comparison of an unpatched and patched version.
\textsc{BitBlaze} was used for checking HTTP and NTP protocol
servers~\cite{bitblaze_net}.  \textsc{Bitscope}~\cite{bitscope} is built on
the same cornerstone as \textsc{BitBlaze}, but provides understanding of
dependencies between program inputs and outputs. Both were used for
recognizing malware in software.  Another true concolic executor is
\textsc{Sjpf}~\cite{sjpf} stemming from
\textsc{PathFinder-SE}~\cite{pathfinder_se}. The difference between
\textsc{Sjpf} and \textsc{PathFinder-SE} is that \textsc{Sjpf} symbolically
executes the Java bytecode instead of source code, therefore performs the
symbolic execution more directly. Nevertheless, a \textsc{Sjpf}'s downside is
that it still does not track constraints in both directions as the executors
above: it tracks the constraints only when switching from the concrete
execution to the symbolic one.

Besides the aforementioned papers, there are also overviews containing a
summary of some symbolic execution techniques we have
mentioned~\cite{se_review88, se_review09, se_review10, se_review13}. And note
that we also employ the symbolic execution in our tool described in
Chapter~\ref{sec:fmics}.

% SADAT and IVS are missing
%EL1~\cite{se_el1}



\section{Program Slicing}\label{sec:slicing}
As was described in the previous section, the symbolic execution is prone to
slowness when analyzing programs with loops or many branchings. We mentioned
some techniques to cope with this problem in the previous section too, but we
have not enumerated them all of course. There is another approach that can be
used in an analyzer to help the symbolic execution. The requirement is to have
a set of code locations, e.g.\ lines with potential bugs. In that case, the
analyzer can simply remove all the code which does not affect the flow to
those code locations in any way and the analyzer can evaluate only the
relevant part.  This is exactly a task that can be performed by \emph{program
slicing}.  This is a technique that was described by Weiser in his PhD thesis
in 1979~\cite{slicing_phd} and later also published as an
article~\cite{slicing}.  As the slicing leaves only the code that really
affects particular code locations, the slicing may bring a considerable
speed-up to techniques working with code in general, not only to the symbolic
execution.

\newcommand{\newsym}[1]{\ensuremath\boldsymbol{\mathsf{#1}}}

When we want to slice a program, we have to specify what behaviour we want to
preserve in the program first. In the slicing, this is performed by defining
\emph{slicing criteria}. These are chosen lines of the program accompanied
with variables we are interested in at those lines. Note that there is no
limitation on the variables like whether they are set or used on those lines.
Such defined criteria ensure that values of variables at those lines must be
equal in both original and sliced program for any execution. We can slightly
loosen the requirement of ``any execution'' to ``a specified execution'' as we
will see later nonetheless.  Thereafter, when a slicer is given the slicing
criteria, it processes functions in a program independently, in an
intra-procedural way. The slicer starts with some function and assigns a set
of \emph{relevant variables} to each line of the function. For lines with
some slicing criteria, the variables from the criteria are put into the
relevant variables. Otherwise the relevant sets are left empty.  Further,
according to the semantics of program commands, the slicer propagates relevant
variables among function lines and inserts discovered variables into relevant
sets on corresponding lines. This happens as follows:
\begin{itemize}
\item All relevant variables from a successive line are copied to the current
line's set except for those which are set on the current line. This is a
simple information \emph{propagation}.
\item If the current line sets some variable which is relevant on the next
line, the slicer adds all variables which the current line references.  I.e.\
the slicer performs computation of a \emph{direct dependency}.
\item If the current line is a branching node (or loop head) and some line in
the branch (or loop body) sets a relevant variable, the slicer adds all
variables referenced by the current line to the current line's relevant set.
This is \emph{indirect dependency} and serves the purpose of leaving in place
branches and loops which indirectly affect slicing criteria.
\end{itemize}
The slicer continues the evaluation of the function statements repeatedly
until the sets stop increasing in size, i.e.\ a fixed point is reached.  From
the content of the sets and the program semantics we can safely claim, which
program lines have no impact on the initial slicing criteria.
Those are exactly the lines, where the intersection between variables that are
set on that line and relevant variables on the \emph{successive} line(s) is
empty.  Slicers also have to leave lines with branches or loop heads that
affect some unsliced lines from the previous sentence. The rest can be removed
from the body of the function.  Let us demonstrate the computation on the
example of Table~\ref{tab:state_df_eq}.  Suppose we want to slice the program
according to line~\ref{lst:state_sl_crit} and variable \code{a} there. The
computation of relevant sets denoted as $R_{7,a}$ is demonstrated in
Table~\ref{tab:state_slicing}.
\newcommand{\eset}{$\emptyset$}
\begin{table}[htb]
\begin{center}
\begin{tabular}{|r|l||c|c||c|c|c|} \hline
\multirow{2}{*}{\textbf{\footnotesize \#}} &
  \multicolumn{1}{c||}{\multirow{2}{*}{\textbf{Code}}} &
  \multicolumn{2}{c||}{\textbf{Variables}} &
  \multicolumn{3}{c|}{\textbf{Fixed Point Iteration}} \\ \cline{3-7}
&& \textbf{Set} & \textbf{Read} & 0 & 1 & 2 \\ \hline\hline

1 & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
a = 1;
\end{lstlisting} } & $\{\var{a}\}$ & \eset   & \eset & \eset   & \eset \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
b = read();
\end{lstlisting} } & $\{\var{b}\}$ & \eset   & \eset &
  $\{\newsym{a_{P}}\}$ & $\{\var{a_P}\}$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
while (b) {
\end{lstlisting} } & \eset   & $\{\var{b}\}$ & \eset &
  $\{\newsym{a_{P}}\}$ & $\{\var{a_P},\newsym{b_I}\}$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
  a = a+5;
\end{lstlisting} } & $\{\var{a}\}$ & $\{\var{a}\}$ & \eset &
  $\{\newsym{a_{D}}\}$ & $\{\var{a_D},\newsym{b_P}\}$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
  b = b-1;
\end{lstlisting} } & $\{\var{b}\}$ & $\{\var{b}\}$ & \eset &
  $\{\newsym{a_P}\}$   & $\{\var{a_P},\newsym{b_P}\}$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
}
\end{lstlisting} } & \eset   & \eset         & \eset &
  $\{\newsym{a_P}\}$   & $\{\var{a_P},\newsym{b_P}\}$ \\ \hline

\thelstnumber & {\begin{lstlisting}[name=state_sl_comp,numbers=none,frame=none]
use(a);(*@ \label{lst:state_sl_crit} @*)
\end{lstlisting} } & \eset   & $\{\var{a}\}$ & $\{\newsym{a_C}\}$ &
  $\{\var{a_C}\}$  & $\{\var{a_C}\}$ \\ \hline

\end{tabular}
\end{center}
\caption{Program slicing of example in Table~\ref{tab:state_df_eq}. The fixed
point computation of relevant variables is sped up.}
\label{tab:state_slicing}
\end{table}
Note that the particular iterations of the fixed point calculation are sped up
a little. We allow the propagation rule to apply many times in one column for
the presentation purposes. Also, we make newly added relevant variables in an
iteration bold.  We note the rule which applied to add the variable too: the
variables are marked with one of \var{C}, \var{P}, \var{D}, \var{I} in
subscript. It means their source is slicing criteria (\var{C}), propagation
rule (\var{P}), or direct (\var{D}) and indirect (\var{I}) dependency. The
result of the slicing is that we cannot slice anything from the program,
because \code{a} on line~\ref{lst:state_sl_crit} depends on all lines. As we
can see, the intersection of relevant and set variables is never empty.

To extend the slicer to be inter-procedural, we have to change the slicer to
pass the relevant variables among functions: down to a called function at
every \emph{call} instruction and back up from the called function. As relevant
variables propagate from the bottom up, the slicer has to pass the variables
from a call site to return points of the called function and when the slicer
is done with the computation of the relevant variables through the whole
function to its beginning, the variables are transferred back to the call
instruction. Of course the slicer has bind actual parameters with formal ones
and take care of the relationship between the return value and the variable
where the result is stored into. Finally, we have to ensure that the set
variables of the call instruction proper are those which the called function
modifies, so that the call instruction is not sliced when the called function
modifies some relevant variable.  The same as above, the process repeats until
a fixed point is reached. The inter-procedural slicing was discussed by
Horwitz et al.~\cite{slic_inter1} and Gallagher~\cite{slic_inter2}.

Originally, there were only slicers implementing the technique as described
above. These are \emph{static} slicers and they slice such that any possible
input to a program is handled the same way. But slicing is often used for
debugging purposes~\cite{slicing_debugging} and there, we are usually
interested in a value of some particular variable only when some specific
input was passed to the program.  Imagine we have a bug in the program and
know what was passed on the input. We also know that a variable printed on the
output has an invalid value. The slicer can be therefore limited to this
input. In particular, the input need not trigger visiting all bodies of
branches or loops, so this allows for much more code removal than static
slicing.  This approach is called \emph{dynamic} slicing~\cite{slicing_dyn}.
Note the slicer has to look at conditions and track a path condition and this
is usually slow. So this obviously results in a drawback: computing a dynamic
slice is more computationally expensive than computing static slices. Since
control flow graphs are used to compute slices (not lines like in the original
algorithm), there appeared techniques to overcome the problem by pruning of
the control flow graphs. They can significantly decrease the computation time
of dynamic slicing~\cite{slic_prunecfg1, slic_prunecfg2}, but the result is
less precise.  \emph{Hybrid} slicing tries to tackle this problem by combining
the dynamic slicing with dynamic runs.  Using breakpoints, slicers record
which code locations are visited with which input and compute the dynamic
slice according to the locations.  It is both faster and less space consuming
than the dynamic slicing.  The dynamic slicing can be improved such that it
works with more than one input value~\cite{slic_quasi, slic_union}, which can
be generalized by describing input values via
constraints~\cite{slic_constraint, slic_constraint_bw}.

There is also an orthogonal method to the static slicing. We may be
interested in all program statements which \emph{are affected} by some
variable at some line. Since the standard slicing is sometimes called
\emph{backward} slicing, the orthogonal approach was named \emph{forward}
slicing~\cite{slic_forw}. The forward slicing is used for dead code removal
or to show subsequent uses for debugging purposes for example. Both the
forward and backward slicing together can be generalized as
\emph{chopping}~\cite{slic_chopping}. The chopping requires two distinct sets
of slicing criteria: one about what a variable affects (forward slicing), the
other about what affects that variable (backward slicing). The slicing can be
abstracted a bit more~\cite{slic_abstract}: it can answer what \emph{values}
in a variable affect statements given as slicing criteria.

When taking over and reusing a library, a programmer can reduce the library
code size using slicers. Programmers just need to specify the parts of the
library which are used, and the library is sliced~\cite{slic_interface}.
The authors suggest also use in the reverse engineering -- slicing closed code
by some guessed criteria helps recognition tools to find patterns and reverse
engineer the code. Further, some variables like temporaries or iterators are
unwanted in slices when debugging. They can be disabled to appear in the
slice~\cite{slic_stop}. Similar is the \emph{barrier}
slicing~\cite{slic_barrier}. Barriers are statements which should be ignored
by a slicer including their successors.  So barriers are often at interfaces
of massive use, and therefore likely error-free libraries.  The slicing can
be also used for finding errors using \emph{dicing}. We have to know two
locations in code. On one location, there is a variable which has a correct
value. On the other there is a variable with a wrong value. We slice the code
separately with those two variables and locations as criteria and the slices
are compared. Intersection of the slices are lines of code which are likely
correct. The error has to be likely in the rest of the incorrect slice as this
was not a part of the correct slice. Slicers can work also on the semantic
level, to not preserve syntax, but leave semantics with respect to
criteria~\cite{slic_amorphous}. The slicing can reason about some paths in a
program~\cite{slic_path} such as why a statement on a particular path is not
executed? It can be used for proving correctness~\cite{slic_mc} and for
concurrent programs too~\cite{slic_concur}.  And finally, Weiser suggested to
use the slicing for transfomation of a program to a couple of parallel
programs: slices can run concurrently on separate processor without shared
memory~\cite{slicing}.  More details about slicing and its variations and uses
can be seen in various surveys and reviews~\cite{slic_overview95,
slic_overview96, slic_overview98, slic_overview01a, slic_overview01b,
slic_overview05, slic_overview12}.

We use slicing extensively to reduce code size of the code that is
symbolically executed by our tool. More insight can be seen in
Chapters~\ref{sec:fmics} and~\ref{sec:svcomp}.

\section{Model Checking}
This is a bit different technique from static analysis we are dealing in this
thesis with. It does not check the program proper, but instead tries to prove
some properties on a model of the program. The model is represented by some
finite state machine and the properties are a specified behaviour which has to
be satisfied in time. For example a property can be ``This message has to come
as a reply in the future.'' Properties are described by logic formulas,
usually in the temporal logic like LTL or CTL (see more about temporal logics
in {\O}hrstr{\o}m and Hasle's book~\cite{temporal}). Correctness of the system
is then established, if every run of the system corresponds to a particular
formula. If it does not for some case, a counterexample consisting of system
states is generated. Obviously, even if no counterexample is generated, the
system still might not be correct in case its model or checked properties are
incorrect. And yet, according to Ball et al., developing a thorough and
precise model may take years~\cite{mc_years}.

There are two basic approaches to model checkers depending on how the states
are stored: \emph{explicit} and \emph{symbolic}. Explicit model checkers
enumerate the states one by one as was implemented e.g.\ in
\textsc{Divine}~\cite{divine}. Symbolic model
checking~\cite{mc_symbolic1,mc_symbolic2} started representing sets of states
by boolean functions. For example if we have two 1-bit variables \code{a} and
\code{b} and their values are always the same (1 and 1 or 0 and 0 pairwise),
we can represent it by a formula \code{a} $\Leftrightarrow$ \code{b} instead
of storing all the values. The formulas can be represented in a space-saving
way by \emph{Reduced Ordered Binary Decision Diagrams}~\cite{bdd} (ROBDD or
simply BDD).  Despite the exponential complexity still persists (due to the
path explosion problem), BDDs helped to push the model checking usability
boundary a lot.  Both the symbolic model checking and BDDs allowed more
complex systems to be checked on the same hardware (cf.~\cite{mc_symbolic1}
again).  Symbolic model checking was later improved to defer the path
explosion problem more by techniques like \textsc{Cegar}~\cite{cegar}.
\textsc{Cegar} starts with an abstraction of the model and makes it more
precise whenever a \emph{spurious} counterexample is produced.  And only in
that part of the model which is too abstract. Examples of symbolic model
checkers are \textsc{NuSmv}~\cite{nusmv} or \textsc{Bebop}~\cite{bebop}.

In parallel to the BDDs-based model checking, \emph{bounded model checking}
exists. Instead of exploring the whole state space of the system model, there
is a predefined constant \textit{k} which determines the maximum length of
execution to be explored. If a bounded model checker explores all these
\textit{k}-prefixes, either \textit{k} is incremented and the checking
continues, or the model checker is stopped. It can have two reasons: the user
set the top bound for \textit{k} (the analysis is of course incomplete in that
case), or the execution is infeasible, because it runs out of resources. Few
representatives are \textsc{Esbmc}~\cite{esbmc} and
\textsc{Llbmc}~\cite{llbmc}. Model checking and bounded model checking are
orthogonal techniques. It was shown in~\cite{bmc_survey} that both of them
reason about different class of properties. (Nonetheless there can be an
intersection.)

Due to the path explosion problem and relatively slow computers in the past,
use of the model checking was limited rather to check small reactive systems
in industry like manufacturers of processors and automotive control systems.
Later, there started appearing more successful tools to check large software
projects. For example model checker \textsc{Blast}~\cite{blast} failed first
due to inability to reason about pointers in the Linux kernel
drivers~\cite{blast_kernel}. It was shown later, that \textsc{Blast} needs an
instrumented code and then works significantly better.  So it is currently
successfully used in the Linux Driver Verification toolkit~\cite{ldv} to
verify Linux device drivers. \textsc{Blast} was also extended to handle
concurrent data accesses~\cite{blast_concur}.  \textsc{Spin}~\cite{spin} is
aware of concurrent data accesses too. So \textsc{Spin} checked a model of the
Linux common filesystem layer~\cite{spin_vfs}. Earlier, some individual file
systems which use this layer were checked by Yang et al.\
too~\cite{mc_kernel_fs}. Other subsystems like the networking stack were
also checked~\cite{mc_kernel_ip}.  \textsc{DDverify} verifier~\cite{ddverify}
and \textsc{Saturn}~\cite{saturn} were used for automated verification of the
Linux kernel drivers.  \textsc{Undertaker}~\cite{undertaker} is a model
checker for the Linux kernel configuration system. It validates dependencies
in the \texttt{Kconfig} configuration files.  There are also model checkers
aimed specifically at the Linux kernel like \textsc{Predator}~\cite{predator}.

Similar to the attempts on the Linux kernel, Microsoft started adopting their
model checker \textsc{Slam}~\cite{slam_cav, slam_popl, slam_decade} and
\textsc{Slam 2}~\cite{slam2} in their driver verifier \textsc{Sdv}~\cite{sdv}
for checking Windows drivers. They apparently succeeded as they require every
driver to be checked by this toolkit with over 160 predefined temporal rules
before the driver is accepted by Microsoft.  \textsc{Sdv} is a proprietary
software with rules tailored to Windows drivers.  Therefore it cannot be used
for checking open source projects. So Clarke et al.~\cite{cbmc} adapted
Microsoft's approach using \textsc{Cbmc} model checker and defined at least
seven rules to check Linux kernel drivers. The rules were later extended by
Post and K\"{u}chlin~\cite{slicx}.  There exist also successful tools for
modelling and verifying Java programs like
\textsc{PathFinder}~\cite{pathfinder}.

As this thesis does not deal with the model checking, curious readers are
pointed to a survey~\cite{mc_survey} and books~\cite{mc_book1, mc_book2} about
model checking.  There is also a survey about bounded model
checking~\cite{bmc_survey} available. To the best of our knowledge, there are
no later surveys or books covering the topics in that scale as these four do.

\section{Industrial Tools}\label{sec:industrial}
There are many successful industrial tools based on the static analysis.
Unfortunately, for some of them, it is not really known what technique they
are based on. It is due to secrets the companies want to hold, so that their
competitors will not find out. From what can be found in the reference
materials, the tools often use combination of the techniques we have already
described.  They work on the top of various pattern matching, data-flow
analyses, and some of them maybe use the abstract interpretation.  Even though
the internals are not known exactly, we at least try to summarize the tools
and the little information which their providers make available.  Prevalently,
the tools below are capable to handle C, some of them C++ and other languages
like C\# or Java.

The data-flow analyzer \textsc{Lint} was an inspiration for Gimpel's
\textsc{PC Lint/FlexeLint}~\cite{pclint} and Cleanscape's
\textsc{LintPlus}~\cite{lintplus}. The same as their predecessor, they aim at
finding suspicious code constructs. MathWorks'
\textsc{PolySpace}~\cite{polyspace} was one of the first abstract interpreters
capable of handling real code. But according to the authors, the tool's upper
limit is still approximately fifty thousand lines of code that can be
analyzed.  The tool was inspired by the crash of Ariane~5 (see the
introduction).  Another abstract interpreter which can handle large scale
programs is AbsInt's \textsc{Astree}~\cite{astree}. It can be tuned to be
sound, i.e.\ to produce no false positives.  \textsc{Astree} also combines
several abstractions to perform more accurate analysis~\cite{astree_combine}.

\xgcc was a predecessor of Coverity's \textsc{Prevent}~\cite{coverity}, and
\textsc{Metal} preceded Coverity's \textsc{Extend}. \textsc{Prevent} is run
regularly to check the Linux kernel and the results are provided to kernel
developers.  KlocWork's \textsc{Insight}~\cite{klocwork} and GrammaTech's
\textsc{CodeSonar}~\cite{codesonar} are specialized to finding security and
reliability issues. \textsc{Insight} was originally intended as an optimizer
for C code. IBM's \textsc{Apollo} aims at checking web
applications~\cite{apollo}, more precisely at \textsc{PHP} scripts.

Microsoft developed many tools based on the symbolic execution.  Even though
\textsc{PREfix}~\cite{prefix} explicitly does not mention the symbolic
execution., it is used there to find errors in memory management for example.
\textsc{PREfix} has no problems analyzing large projects thanks to a
computation of function summaries, but needs user intervention to model
library functions.  This was improved in \textsc{Archer} of Xie et
al.~\cite{archer}. A similar use as IBM's \textsc{Apollo} above has
Microsoft's \textsc{Pex}~\cite{pex}. It is a symbolic executor intended for
checking of .NET web applications. Microsoft developed also a fuzz tester
\sage~\cite{sage}.  It is a concolic executor working directly on the x86
assembly code. It is therefore able to work also with libraries whose source
code was not made public.  \textsc{Yogi}~\cite{yogi} is a tool which combines
symbolic execution with dynamic execution. It was experimentally used to
replace \textsc{Slam}~\cite{slam_decade} in \textsc{Sdv}~\cite{sdv} with
comparably better results. Microsoft also released \textsc{PREfast}, a tool
that contains a set of rules that a code should fulfill. \textsc{PREfast}
checks code against the rules with a use of annotations. Its main use is to
disallow developers to commit changes that contain certain classes of errors
such as inter-procedural integer overflows.

%\url{http://spinroot.com/static/}
%\url{http://web.archive.org/web/20120316050109/http://www.testingfaqs.org/t-static.html}
%\url{http://samate.nist.gov/index.php/Source_Code_Security_Analyzers.html}
%\url{http://web.archive.org/web/20050405224446/http://www.programming-x.com/programming/analyzer.html}

\section{Summary of the Linux Kernel Checkers}
Some of the tools above can analyze kernels or only drivers of operating
systems. In this section, we specifically sum those which are capable to
handle the Linux kernel. We discuss or use many of them in the following
chapters, hence it we mention them here explicitly.

One of the basic ones is based on the pattern matching and every developer
is supposed to use it before sending patches to the kernel maintainers. It is
named obviously after what it is intended to do as
\textsc{checkpatch}~\cite{checkpatch}. It only checks for patterns like if
somebody introduces a whitespace to the code or if the code is properly
indented. Some better analysis can be performed by a semantic parser called
\sparse~\cite{sparse}. It is still a tool working with some sort of patterns
matched against the parse tree.  It was written by the founder of the kernel
and is tailored exclusively to the kernel as the checks have only a little
sense elsewhere.  The checks include basic intra-procedural checks for proper
locking, use of zero instead of \code{NULL}, or extended type checking using
the annotated code in the kernel. Yet another pattern matcher is Dan
Carpenter's source matcher \textsc{Smatch}~\cite{smatch}. Formerly, it was a
set of \perl programming languages scripts. They were later turned into C code
and \sparse was used as a base. \textsc{Smatch} added many more checks
regarding the Linux kernel like doing direct memory accesses on a stack or
passing incorrect arguments to picked functions. Even though Coverity's
\textsc{Prevent} is an industrial tool, it is constantly checking the kernel
and reports are made available. The reports are of diverse kinds like dead
code, null pointer dereferences, use before test and so on. Many of them are
getting fixed. These four tools are extensively used by the kernel developers
and errors reported are getting fixed constantly~\cite{linux_faults}.

There are also tools that can be used, but are not used that massively by so
many developers. These include \clang~\cite{clang} and
\textsc{Cppcheck}~\cite{cppcheck}.  Historically, \xgcc was used to find
errors in the kernel~\cite{metal_kernel}, mainly for errors like if there are
proper checks of untrusted data coming from the network or an user to the
kernel. It found over one hundred of such errors. As \stan~\cite{stanse} is a
tool based on a similar technique, it was used the same way and found over one
hundred another errors.

There are also projects based on model checking which are successful to check
the Linux kernel. Linux Driver Verifier (\textsc{Ldv})~\cite{ldv} is one of
those. The same as \textsc{Prevent}'s reports, \textsc{Ldv}'s reports are sent
to the subsystem maintainers so that they can look at them and fix them.

%\textsc{Herodotos}~\cite{herodotos} compares several version of checked
%software and is able track errors how are they propagated over time.

\paragraph{On Incompleteness} The survey of the techniques and algorithms is
by no means intended to be complete. There are still approaches which we have
not described for some reasons. Be it the obsoleteness of the approaches,
limited use, or finally the limited scope of this thesis. Should the reader
ever come over a tool that has to be within this thesis, please contact the
author to fix the work accordingly. He is allowed to do so even after the
thesis is in the peer review process.
