% vim: textwidth=78 ai spell spelllang=en filetype=tex sw=2

\chapter{Participating SV-COMP}\label{sec:svcomp}
At some point, we had a couple of tools, we were checking the Linux kernel and
finding real errors. We had not considered any other use for our tools until
we were pointed by the authors of \textsc{Predator}~\cite{predator} to the
past \emph{Competition on Software Verification} (SV-COMP)
2012~\cite{sv-comp12} in which they participated. We decided to participate in
the next year's competition with our tool \symbio (see
Chapter~\ref{sec:fmics}). To introduce the reader, we describe the competition
in this chapter first.

We had to improve the tool to fulfil the requirements of the competition. The
tool was also unable to handle some benchmarks from the competition. Also it
gave incorrect results about some other benchmarks. It had several reasons we
had to cope with and they are dealt with in this chapter too.

Participation in this competition is documented in a paper published in the
Springer's Lecture Notes in Computer Science~\cite{symbiotic-svcomp} and
presented on the competition as a part of the 19th International Conference on
Tools and Algorithms for the Construction and Analysis of Systems. There is
also a competition report from 2013 by Beyer~\cite{sv-comp13}.  And yet, since
one of authors of every tool is asked to be in the competition jury, the
author of this thesis was pleased to take that role.

\paragraph{Chapter Organization} In the following
Section~\ref{sec:svcomp_competition} we describe the competition proper. In
Section~\ref{sec:svcomp_participation} we characterize our path with \symbio
through the competition. Section~\ref{sec:svcomp_results} shows some results
from the competition and compares them to the current, improved results of
\symbio.  Finally, we draw some conclusion about our participation in
Section~\ref{sec:svcomp_conclusion}.

\section{The Competition}\label{sec:svcomp_competition}
First, let us introduce to the competition. To its aims, procedure, rules,
and so on. Although there existed some competitions prior SV-COMP, they were
only limited in what they demand by the competitors according to the
SV-COMP organizers. On the contrary, SV-COMP was established to attract
developers of code analyzers and to push them to improve their tools. The
tools should be fully automatic, able to handle real code, contain
state-of-the-art approaches, improve the state-of-the-art, and finally the
tools should present themselves to a wider audience too.

In the 2012's competition, the benchmarks for exercising tools were required to
be in \emph{C intermediate language} (CIL). CIL is a subset of the C language
and the requirement was a bit in a contradiction to the requirement on
handling large, real projects. So this was changed in 2013's competition and
the benchmarks are now in CIL or ANSI~C~\cite{ansi_c}. The benchmarks are
divided into several categories. The basic benchmarks split is
\emph{concurrency}, \emph{memory safety}, and the rest. The rest comprehends
\emph{bit vectors}, \emph{control flow integers}, \emph{device drivers},
\emph{feature checks}, \emph{heap manipulation}, \emph{loops}, \emph{product
lines}, and \emph{system C}. Concurrency and memory safety benchmarks are
special as they need specific tools. One needs a tool that can find errors in
parallelism, the other which can find memory errors. The rest of the
benchmarks contains a line marked by the \code{ERROR} label. The tools should
answer whether it is reachable. Overall, some of the benchmarks are small,
simple code, but many are also large code bases, e.g.\ taken from the Linux
kernel.

There are three phases during the competition. The first one is a
\emph{submission of benchmarks}. Everybody is allowed to submit a benchmark to
bother other's tools a bit. This phase is optional, but allows the developers
to increase chances of their tool. There is a deadline for this phase. After
that, the second phase starts. It is a \emph{training} phase. This is when
benchmarks are frozen and everybody can tune their tools to work as good as
possible for the benchmarks. This phase usually lasts two months. And at the
end of this time period, tools and papers about them are due too. This is to
formally enroll the tools. Now, the \emph{evaluation} phase starts. The tools
are run by the organizers on all benchmarks and are evaluated. After few
weeks, the results are made available to the authors and the evaluation ends.
Then, similar to other conferences, authors of accepted papers are requested
to provide a final version of the papers. They are finally presented on the
conference.

The competition has its rules for the evaluation phase. Every tool, when
it is evaluating a single benchmark should give some answer. It can find
nothing and declare the benchmark as safe. Or, it can find an error and is
expected to provide an error path through the program. Finally it can flag
``unknown'' in case it cannot decide. Every answer the tool gives is scored
according to Table~\ref{tab:svcomp_scores}.
\begin{table}[htb]
  \begin{center}
  \begin{tabular}{|l|c|c|} \hline
  \textbf{Tool's}  & \multicolumn{2}{c|}{\textbf{Benchmark}} \\ \cline{2-3}
  \textbf{Answer}  & \textbf{Safe} & \textbf{Unsafe} \\ \hline
  \textsf{Safe}    &  2 & -8 \\
  \textsf{Unsafe}  & -4 &  1 \\
  \textsf{Unknown} &  0 &  0 \\ \hline
  \end{tabular}
  \end{center}
  \caption{Scores for answers in SV-COMP.}
  \label{tab:svcomp_scores}
\end{table}
Some tools happen to compute the results for a very long time. Hence there is
a time limit and if the tool does not finish in the limit, the organizers kill
the process for that particular benchmark and consider the answer as unknown.
In the 2013's competition, the limit was fifteen minutes.

After a tool is run for all benchmarks in a category, the scores are
summarized. All tools are compared according to the sum. If there are two with
the same score, the running time is considered. Finally, when all categories
are evaluated for every tool, the overall score is computed. It is weighted by
benchmark counts over categories to ensure fairness. The formula used for
computation is in Equation~\ref{eq:svcomp_overall}, where we denote $C$ the
set of benchmark categories, $s_i$ the score in the category $i$, $n_i$ the
number of benchmarks in the category $i$, and $\overline{n}$ the average
number of benchmarks in all categories. The reasoning about the urge for this
formula is in the 2013's report~\cite{sv-comp13}.

\begin{equation}\label{eq:svcomp_overall}
s_{overall} = \left( \sum_{i \in C} \frac{s_i}{n_i}\right) \cdot \overline{n}
\end{equation}

When the final score is known, the results are given to the authors of the
tools. They have still a possibility to review the score and raise an issue
if something really unexpected happened with their tool.  And if everything is
find, they still have a possibility to \emph{opt-out}. If some score in a
category is too unfavorable, they can have no score in the category by opting
out. But that also means, they will have no score in the overall category,
i.e.\ they cannot win the category.

\section{Our Participation}\label{sec:svcomp_participation}

We decided to participate in the competition with one of our tools. And we
chose to participate with \symbio. Since we have already provided a detailed
description of the \symbio internals in Chapter~\ref{sec:fmics}, let us now only
summarize that the tool combines instrumentation, program slicing, and
symbolic execution in order to detect errors described by automata. More
precisely, we instrument a given program with code that tracks runs of
automata representing various erroneous behaviors. If an instrumented
automaton enters an error location during a program execution, then the
original program contains an error specified by the automaton. After
instrumentation, we slice~\cite{slicing} the program to reduce its size
without affecting runs of the instrumented automata.  Finally, we symbolically
execute~\cite{se_king} the sliced program to find errors in the program.

As we already know, the competition contains three sets of benchmarks. Since
\symbio handles neither concurrency nor is prepared for memory errors, we
wanted to cope with the rest of benchmarks only. As reachability of an
\code{ERROR} label is the only error considered in those benchmarks, we wrote
an instrumentor to put \code{assert(0)} function calls at \code{ERROR} labels
in the code.  This is performed by a \textsc{Bash} script with a help of the
stream editor \textsc{Sed}. We settled with six patterns which appear in the
code, see \code{instrument.sh} in the \symbio distribution for details.

Given the instrumented code, we executed \clang 3.1 to produce the \llvm
bytecode. It was in turn inter-procedurally sliced with respect to slicing
criteria, which are the instrumented \code{assert} calls. In other words we
remove all the code except that has an effect on reachability of \code{assert}
calls. The sliced \llvm bytecode is finally symbolically executed by
\klee~\cite{klee}, which itself uses the \textsc{Stp} constraint
solver~\cite{stp}. We obtained both \textsc{Klee} and \textsc{Stp} from the
respective GIT repositories and used the snapshots.

There are several possible outputs that can be given by \klee.  It can either
find a reachable \code{assert} and report it, or finish the computation
without any report, or terminate in some errant way (out of time, out of
memory, invalid memory dereference, or some internal error for example). We
map these to the demanded answers: \code{UNSAFE}, \code{SAFE}, or
\code{UNKNOWN} respectively. This is taken care of in a simple scripted
filter.  The filter also uses \textsc{Bash} with a help of \textsc{Grep}.
Also, when the answer is \code{UNKNOWN}, \klee generates a counterexample for
each error path in some \code{benchmark.c} to the directory
\code{benchmark.c-klee-out/}.

For ease of use, all parts of the tool's pipeline are one by one run by a
single script \code{runme}. It accepts exactly one parameter: the benchmark
file. The pipeline of the whole process is depicted in
Figure~\ref{fig:svcomp_pipeline}.

\begin{figure}[tb]
\begin{center}
  \begin{tikzpicture}[node distance=5mm]
    \node [tl_phase] (C) {Benchmark};
    \node [tl_state,right=of C] (SYMBIO) {\symbio}
      edge [tl_edge] node {} (C);
    \node [tl_phase, right=of SYMBIO, align=center] (KLEE_RES) {
	\klee \\
	Results
      }
      edge [tl_edge] node {} (SYMBIO);
    \node [tl_state,right=of KLEE_RES] (FILTER) {Filter}
      edge [tl_edge] node {} (KLEE_RES);
    \node [tl_phase, right=of FILTER, align=center] (FINISH) {
	\code{SAFE}/ \\
	\code{UNSAFE}/ \\
	\code{UNKNOWN}
      }
      edge [tl_edge] node {} (FILTER);
  \end{tikzpicture}
\end{center}
  \caption{Pipeline of the tool at the competition. For the internal pipeline
  of \symbio, see Figure~\ref{fig:fmics_pipeline}.}
  \label{fig:svcomp_pipeline}
\end{figure}

%}}}
%{{{ Results

\section{Results}\label{sec:svcomp_results}

The strength of the tool lies in its high precision of the answers. In theory,
the only source of incorrect answers is the slicer: it can completely remove
an infinite loop in some cases and thus an unreachable \code{ERROR} label
located below the loop may become reachable. Nonetheless, there was no such
case in the competition benchmarks. 

The main weakness of the tool is high occurrence of \code{UNKNOWN} answers.
These answers come mainly from the high computation cost of the symbolic
execution of programs with loops or recursion. This problem is relieved by
slicing, but there are still many cases where the sliced code remains complex
and the symbolic execution runs out of time or memory. Other sources of
\code{UNKNOWN} results are internal errors of \klee and general limitations of
constraint solving.

\begin{table}[tb]
\begin{center}
\begin{tabular}{|l||c|c||c|c|} \hline
\multirow{2}{*}{\textbf{Category}} &
  \multicolumn{2}{c||}{\textbf{Competition}} &
  \multicolumn{2}{c|}{\textbf{Current}} \\ \cline{2-5}
& \textbf{Incorrect} & \textbf{Score} &
  \textbf{Incorrect} & \textbf{Score} \\ \hline\hline
Bit Vectors           &   5 &   -2 &  0 & 32 \\ \hline
Control Flow          &
  \multirow{2}{*}{0} & \multirow{2}{*}{28} & \multirow{2}{*}{0} &
  \multirow{2}{*}{36} \\
Integers (Precise)    &     &      &    &    \\ \hline

Control Flow          &
  \multirow{2}{*}{4} & \multirow{2}{*}{-24} & \multirow{2}{*}{5} &
  \multirow{2}{*}{-17} \\
Integers (Simple)     &     &      &    &    \\ \hline

Device Drivers        &  23 &  870 &  4 & 1166 \\ \hline
Feature Checks        &  17 &   23 &  0 &  191 \\ \hline
Heap Manipulation     &   4 &  -12 &  6 &  -20 \\ \hline
Loops                 &  33 & -190 &  1 &   39 \\ \hline
Product Lines         & 131 & -577 &  0 &  352 \\ \hline
System C              & n/a &    0 &  0 &   12 \\ \hline\hline
\textbf{Overall}      & 217 & \textbf{-724} & 16 & \textbf{1104} \\ \hline
\end{tabular}
\end{center}
  \caption{Results of \symbio in SV-COMP 2013.}
  \label{tab:svcomp_results}
\end{table}

Finally, we present the results from the competition in the second column of
Table~\ref{tab:svcomp_results}. We do not have a number of incorrect answers
from the system C category because the benchmarks were not a part of the
preliminary results we have an access to. The overall score is computed
according to the rules, i.e.\ by Equation~\ref{eq:svcomp_overall}. We see
that the \symbio results were not encouraging at all. In the following
subsection we describe what caused the results to be so suboptimal and
gradually fix the causes.

Quite interestingly, the competition suggested that it is much easier to
verify that there is no error in a benchmark than to find an error there. In
average, it lasted 7.8 seconds to report \code{SAFE} about a safe benchmark
(median is 0.22), but it took 149.3 seconds to report \code{UNSAFE} about an
unsafe benchmark (median is 230). We have not found the source of this
phenomenon yet as it stems from the \klee internals. All the times are
dependent on a computer where the analysis runs. The competition was evaluated
on an Intel Core i7 3.4 GHz with 16 GiB of RAM.

\subsection*{Fixing the Failure}
The failure was caused by several factors. The worst failure to answer
correctly laid in the benchmarks. More than 1200 still did not adhere to the C
standard. In particular, some variables were not initialized, but were used as
conditions for example. \clang compiler decided many times not to generate
code for such conditions.  We fixed many such examples. As there were
thousands in those files, we often provided the organizers only with a
\textsc{Sed} script instead of fixing the errors on our side which lead to a
joke that we thought them \textsc{Sed} scripting. Organizers accepted many of
them, but still, we have to perform some scripting in the \code{build-fix.sh}
script to fix the rest. To confirm the script is working, we removed the
slicing phase temporarily. The symbolic execution then answered only
correctly. But timeouts happened too often without the slicer of course.

The source of the remaining incorrect answers is in the slicer. The slicer in
some cases simply slices too much code. The reasons were errors in the pointer
analysis, in incorrect specification of relevant sets of some \llvm
instructions, and some other minor issues. The pointer analysis was only
field-insensitive. But this caused a removal of some relevant code as we
enlightened about in Section~\ref{sec:fmics_slicer}. So we rewrote it to
support fields. This fixed many of the still failing cases, but there were
still some. They were caused by other minor issues which were fixed one by
one. The results of the fixed implementation are in
Table~\ref{tab:svcomp_results}. Apparently, the slicer still contains some
more errors to be fixed. Ideally, before the next year's competition starts.

%\begin{center}
%  \setlength{\tabcolsep}{3pt}
%  \begin{tabular}{|l|c|c|}
%    \hline
%    & Competition Version & Current Version \\ \hline 
%    Correct Answers \code{SAFE/UNSAFE} & 81 & 116 \\ \hline
%    Incorrect Answers \code{SAFE/UNSAFE} & 17 & 0 \\ \hline
%    \code{UNKNOWN} (including timeouts) & 20 & 2 \\ \hline
%  \end{tabular}
%\end{center}

%}}}
%{{{ Conclusion

\section{Conclusion}\label{sec:svcomp_conclusion}
We participated in the second international competition on software
verification held at TACAS 2013. We enrolled with our tool \symbio. Its
results were not encouraging due to both errors in our implementation and in
the competition benchmarks. We presented, that the competition definitely
helped us to improve the implementation by the order of magnitude. Hence we
are looking forward to the next year's competition to collect some more points
than in 2013.

%}}}
