% vim: textwidth=78 ai spell spelllang=en
\documentclass{acm_proc_article-sp}
\usepackage[utf8x]{inputenc}
\usepackage[final]{hyperref}
\usepackage{color}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{listings}

\lstset{tabsize=2,showstringspaces=false,breaklines=true,breakatwhitespace=true,
	escapeinside={(*@}{@*)},basicstyle=\footnotesize}

\newcommand\Color[2]{\textcolor{#1}{#2}}
\newcommand{\nb}[1]{\marginpar{\scriptsize #1}}
\newcommand\todo[1]{\nb{\Color{red}{#1}}}

\newcommand\st{\textsc{Stanse}\xspace}

\begin{document}

\title{Eliminating Repetitive Errors Found in Consecutive Source Releases}

\numberofauthors{1}

\author{
Ji\v{r}\'\i~Slab\'y\\
	\affaddr{Faculty of Informatics}\\
	\affaddr{Botanická 68a}\\
	\affaddr{Brno, Czech Republic}\\
	\email{xslaby@fi.muni.cz}
}

\maketitle

\begin{abstract}
With a use of contemporary static analysis tools, many errors are found in
source codes. We introduce a mechanism to prevent recurring of perpetual
errors found by the tools in different source releases. The mechanism aims
especially at those errors which were not fixed yet or are false-positives
which will likely be present with its source code forever. We discuss what
algorithms can be used to determine such errors and demonstrate them on output
from \st run on the Linux kernel.
\end{abstract}

\category{D.2.4}{Software}{Software Engineering}[Software/Program Verification]

\terms{Algorithms}

\keywords{Static analysis, error pruning, consecutive version errors}

\section{Introduction}
Recently, many new static analysis tools emerged (for example
\cite{coverity,java_bugfinding,sparse}) with a potential of finding many
long-standing errors in the source code. Mostly the problem with these tools
is that they produce a very long list of errors. The reason is much sources
present in the project (more potential places for errors) and/or
false-positives which are not about to be fixed. From the recurring error
perspective, there is a little what can be done about that in one specific
release of sources. But it becomes a burden when the developers would have to
check the very same subset of reported errors every single release of the
product.

In this case, when such an analyzer is run on a project with release cycle
short enough, it would be convenient to recognize (a) the errors which were
reported in an earlier version and were not fixed yet\footnote{There are many
reasons for this including slow responses to error reports from maintainers.
An example of this is at \url{http://lkml.org/lkml/2009/9/23/221} which had to
be sent three times and was merged even after 3 months.} and (b)
false-positives because they are persistent in time and nobody wishes to see
them again and again in every upcoming released version.

Although it may seem that there are almost no requirements on the project when
we want to use this technique, it is not that easy. Not only the fast
development and quick releases of new versions is a precondition, but also
rigid structure of files so that they don't fluctuate over the source tree.
Further, a thorough development methodology which does not allow developers to
perform inappropriate moves of code parts between files every release and so
on. Every change should have its reason, otherwise the change won't be
accepted by the project management.

One of the projects with the requirements above satisfied is the Linux kernel
for sure. Its release cycle between major versions (e.g.
2.6.33$\rightarrow$2.6.34) is relatively short, oscillating between 11-13
weeks. The changes coming into the tree are not accepted unless appropriately
reasoned, reviewed and merged by subtree maintainers and finally accepted by
L. Torvalds. Also, developers are mostly not allowed to shuffle with sources
arbitrarily over the tree, every class of drivers has its own, predetermined,
directory.

We already mentioned that there are many static analysis tools out there. To
have a fully operational tool working on the Linux kernel, \st \cite{stanse},
a tool developed at the Faculty of Informatics, is used specifically in this
paper. It contains a base with C parser, error reporting facility, user
interface, points-to analysis module and application programming interface for
checkers. A checker is a module which gets the parser output and reports
errors back to the base which are in turn presented to a user. Currently there
are four checkers implemented, in this paper we are interested exclusively in
the \textsc{AutomatonChecker} which is able to find errors described by finite
automata.

The paper is organized as follows. First, in Section~\ref{sec:motiv} we will
see why the technique is important and what benefits it brings, then
Section~\ref{sec:method} introduces the technique itself with possible
extensions and finally we evaluate it with a use of \st output collected from
the Linux kernel in Section~\ref{sec:eval}.

\begin{figure*}[tb]
\begin{center}
\begin{tabular}{|c|c|} \hline
Older version & Newer version \\ \hline \hline
\begin{lstlisting}[language=C]
static DEFINE_SPINLOCK(my_lock1);
static DEFINE_SPINLOCK(my_lock2);
static void fun(int in) {
  spin_lock(&my_lock1);
  if (in)
    return;
  spin_unlock(&my_lock1);
}
.
.
\end{lstlisting} &
\begin{lstlisting}[language=C]
static DEFINE_SPINLOCK(my_lock);
static DEFINE_SPINLOCK(my_lock2);
static void fun(int a) {
  spin_lock(&my_lock1);
  if (a) {
    spin_unlock(&my_lock2);
    return;
  }
  spin_unlock(&my_lock1);
}
\end{lstlisting} \\ \hline
\end{tabular}
\end{center}
\caption{Illustration of false-negative introduced by the \emph{Simple
Solution}}
\label{fig:fp_simple}
\end{figure*}

\section{Motivating Example}
\label{sec:motiv}
Before we dive into the details of the algorithms in the
Section~\ref{sec:method}, consider the following piece of code extracted
(and also slightly manually macro-expanded for clarity) from
\texttt{lib/locking-selftest.c} obtained from the Linux kernel:
\begin{center}
\begin{lstlisting}[language=C]
static DEFINE_SPINLOCK(lock_A);
...
static void double_unlock_spin(void) {
  spin_lock(&lock_A);
  spin_unlock(&lock_A);
  spin_unlock(&lock_A);
}
\end{lstlisting}
\end{center}

The \texttt{double\_unlock\_spin} function intentionally breaks the proper
locking policy. The \texttt{lock\_A} is locked and then unlocked twice, thus
leaving the lock in an inconsistent state. However, in this case, it is the
sole purpose of the function. It is intended for exercising a mechanism for
automatic dynamic locking violations tracking present in the
kernel\footnote{So-called \textsc{LockDep}, more information are available at
\url{http://lwn.net/Articles/185605/}.}. Thus it is impossible to change or
"fix" the code to be correct from the point of view of an automaton which
tracks lock states and watches the policy is not violated. Hence it is an
error (a kind of false positive) which will be with this part of sources
forever.

We did a complete review of errors reported in the Linux 2.6.27 to sort out
which of those are false-positives. All false-positives were marked
accordingly and true errors reported to corresponding code maintainers (some
with appropriate patches). Then we tried to perform the analysis more often,
even on release candidate kernels in a weekly period only, and when inspecting
these candidate versions, we found out that many of the errors are of the
similar kind like this example, i.e. cannot be fixed or were already reported
but not fixed yet for some reason.

To tackle down this problem, we present an algorithm to match errors from
consecutive code releases to avoid repetition of errors presented to
developers in error lists. Thus lowering the burden exposed to them when
walking through the error list generated from later versions. To the best of
our knowledge there is no white paper describing a similar technique.

\section{Methodology}
\label{sec:method}
Before we start with the methodology, we first have to know what the output
of the tools look like, especially what a \st's output is. Even though the
error formatting practices differ tool by tool, in common they contain the
source file name where the error lays with exact line too. The report also
carries an information about what kind of error is that and possibly what
module (checker) generated that. Eventually some tools report another fields
like a line column, but this is irrelevant for further considerations.

In the end, the entry (error) in a list of reports $RL$ is a tuple $E$ (e.g.
in a database as we will see later) such as:
\begin{displaymath}
E = (file, line, checker, error\_description) \in RL
\end{displaymath}

Be a start point two lists of errors $RL_1$ and $RL_2$ generated from two
different versions of the same project. $RL_1$ corresponds to an older
version, $RL_2$ to a newer one respectively.

Now, we want to eliminate such tuples $E$, that correspond to the same error.
How the "same error" can be handled is described in the following subsections.

\subsection{Simple Solution}
\label{sec:simple}
The most straightforward approach of defining sameness is just to compare each
tuple from $RL_2$ to entries in $RL_1$ and obtaining this result:
\begin{equation} \label{eq:simple}
RL^{simple}_2=\{x|x\in RL_2 \wedge x \notin RL_1\}
\end{equation}

Obviously there are several problems tightly connected to this approach.
First, not all the tuple elements are preserved over source revisions. Namely
for example \textit{line} element changes extensively version by version. To
make this technique defunct, it is actually enough to add or remove single
(even empty) line to/from the file\footnote{Technically, if the analyzer works
on the preprocessed code, it is enough that somebody changes any included
file, because the preprocessed output is shifted by the change as a whole.}.

Second, this solution can introduce false-negatives. Consider a code in
Figure~\ref{fig:fp_simple} on the left side where there is a bug in locking.
The function \texttt{fun} omits to unlock \texttt{my\_lock1} depending on the
input parameter \texttt{in}. Given this error is reported to the maintainer of
the code, he tries to fix it. Unfortunately, he fixes it wrong (the right side
of the Figure) and adds a new error by unlocking a wrong lock, but still being
on the same line.

In the end both of the cases the entry will look the same in the report list:
\begin{displaymath}
E=(file.c, 6, AutomatonChecker, Locking\_imbalance)
\end{displaymath}
Inconveniently, the outcome of this is that the second error is hidden from
programmer when the \emph{Simple Solution} is used because $x \notin RL_1$
from the equation~(\ref{eq:simple}) does not hold.

\subsection{Advanced Solution}
This is not the only issue introduced by the \emph{Simple Solution}. But let
us fix it first. The issue only means that \textit{line} is not expressive
enough to describe what is really in the code. Instead of looking at line
number, we can fetch the particular \textit{line} from \textit{file} and store
it in a newly-defined tuple:
\begin{displaymath}
E=(file,line\_contents,checker,error\_description)
\end{displaymath}
This will solve the false-negative described above, because
\textit{line\_contents} differs and so the condition $x \notin RL_1$ does
hold.

In spite of that, there are still issues caused by code shuffling both on a
in-file (functions reordering) and in-tree (file move/rename) bases. This
means the \textit{file} and/or \textit{line} component of the original tuple
changes while leaving the rest (i.e. the error type and description) of the
tuple untouched.

\section{Evaluation}
\label{sec:eval}
We exercised our algorithm on the errors found in the Linux kernel by \st. The
two versions we considered are 2.6.35-rc3-next20100618 and
2.6.35-rc4-next20100706, further referred only as -rc3 and -rc4 respectively.
That means a development versions where there are over 300 thousand lines
changed in 2654 files between them in 18 days. There are no changes in
configuration of the kernel (besides newly added drivers, which were ignored)
and neither of our tool.

The errors reported by \st are represented in an XML form by default. But as
we wanted to have a convenient way for combining and querying the data, we
decided to use databases and some SQL language for accessing the data. The
reason is also to easily represent the tuples. This way, a tuple is simply a
row in a database table.

Although there exists infrastructures for combining XML and SQL like SQL/XML
\cite{sqlxml}, we sticked to a simple database system, more concretely to
\textsc{SQLite} \cite{sqlite} which stores and references databases as single
files somewhere on a backing store.

To achieve the goal, we wrote a Perl script to transform the XML into a
\textsc{SQLite} database file. Despite each report list resides in a separate
database file, it is not a problem because \textsc{SQLite} allows combining
even tables from different databases in its SQL. Thus we can easily reference
errors from two versions and store the result into the third.

Having separate databases for each list allowed us fast deployment in our web
frontend, because it works on a version-per-file basis and needed thus no
changes with this approach.

\subsection{Using Simple Solution}
First, we will have a look at the \emph{Simple Solution} presented in
Section~\ref{sec:simple}. Overall, the results are summarized in
Table~\ref{tab:res}. The \emph{Atomicity} (e.g. forbidden sleeps inside atomic
context), \emph{Locking} (omitted unlocks, double locks, etc.),
\emph{Memory} (NULL dereferences, dangling pointers and so on) and
\emph{Pairing} (grabbed resource not given back to the system for example)
rows correspond to the errors found by the
\textsc{AutomatonChecker}\footnote{The errors found by \st (with their
description) may be seen at \url{http://stanse.fi.muni.cz/bugs.html}.}. They
demonstrate, as expected, that it does not matter on which type of errors the
technique is used. The reduction ratio is more or less the same.

We also expressed the reduction in percents. There are two columns with them,
the first when -rc4 with \emph{Simple Solution} is used, the latter for
\emph{Advanced Solution} being used.

For this solution, the SQL query, sort and insert command used is as follows:
\begin{lstlisting}{language=SQL}
ATTACH DATABASE "35-rc3.db" AS old;
ATTACH DATABASE "35-rc4.db" AS new;
INSERT INTO errors (checker, importance, error, file, line)
  SELECT checker, importance, error, file, line FROM new.errors WHERE
  file || line || checker || error NOT IN
    (SELECT file || line || checker || error FROM old.errors);
\end{lstlisting}
The two error collections are attached as \emph{new} and \emph{old} and used
in the query. The || token stands for string catenation.

The table also contains a column for brute force approach which is even more
naive than the \emph{Simple Solution}. The query above is changed so that line
number is completely removed. In this case we may lose some new bugs, i.e.
have even more false-negative than earlier, but it still seems to be enough in
cases where \st is run on after each kernel release -- on a daily basis.

Time by time (e.g. each version release), the more sophisticated methods
should be used to not lose real bugs.

\begin{table}
\begin{center}
\begin{tabular}{|l|r|c|c|c|c|c|} \hline
          & \textbf{-rc4} & \textbf{-rc4$^{s}$} & \textbf{\%} &
	  \textbf{-rc4$^{b}$} & \textbf{-rc4$^{a}$} & \textbf{\%} \\ \hline
Atomicity & 23   &                 &    &              &    \\ \hline
Locking   & 179  &                 &    &              &    \\ \hline
Memory    & 217  &                 &    &              &    \\ \hline
Pairing   & 81   &                 &    &              &    \\ \hline
Summary   & 500  & 50              &    &              &    \\ \hline
\end{tabular}
\end{center}
\caption{Errors count using none, simple and advanced solutions.
-rc4$^{s}$ expresses the \emph{Simple Solution}, while -rc4$^{a}$ stands for
the \emph{Advanced Solution}.}
\label{tab:res}
\end{table}

\subsection{Using Advanced Solution}
As we already noted, the results are presented for this algorithm in
Table~\ref{tab:res} as well.

\section{Conclusion}
We presented an algorithm for pruning errors which are reported repeatedly
from consecutive versions of a software project by static analysis tools. We
introduced two variants of the algorithm that may be used and reasoned what it
can lead to when only the simpler one is engaged.

The method was implemented and then evaluated on the Linux kernel. We found
out that the need of walking through the errors is rapidly lowered by the
algorithm and thus the person-hours needed are reduced, hence potentially
the overall cost of the product being checked.

\section*{Acknowledgements}
The author is supported by the research centre Institute for Theoretical
Computer Science (ITI), project No. 1M0545.

\bibliographystyle{plain}
\bibliography{inter-errors}

\end{document}
