% vim: tw=78 ai spell spelllang=en_gb sw=2
% syntax spell toplevel

\documentclass[10pt]{llncs}
\usepackage{color}
\usepackage{colortbl}
\usepackage{url}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{xspace}
\pagestyle{plain}

\usetikzlibrary{arrows, positioning, decorations.pathreplacing}

\lstset{language=C,tabsize=2,showstringspaces=false,
	breaklines=true,breakatwhitespace=true,escapeinside={(*@}{@*)},
	captionpos=b,basicstyle=\sffamily\small}
%numbers=left,frame=l

\newcommand{\code}[2][]{\lstinline[columns=fullflexible, #1]$#2$}
\newcommand{\file}[2][]{\code[language={}, #1]{#2}}
\newcommand{\commit}[1][]{\texttt{#1}}

\newcommand{\xes}{\texttt{x86}\xspace}
\newcommand{\xesl}{\texttt{x86\_32}\xspace}
\newcommand{\xesh}{\texttt{x86\_64}\xspace}

\setcounter{tocdepth}{1}

\begin{document}

\title{Stable Patches and SUSE Trees}
\author{Jiri~Slaby}
\institute{Suse Labs \\
\email{jslaby@suse.cz}}

\maketitle

\begin{abstract}
  Here, we provide a short summary of the past and current habits of merging
  stable patches into the SUSE kernel trees. Starting with the sorted series
  section, the policy undergoes a change, so we will sum up what is happening
  in SLE15 and SLE15-SP1 now and what is our plan for future releases.
\end{abstract}

\section{Introduction} \label{sec:intro}
SUSE is maintaining kernels for their products in a git tree~\cite{git_suse}.
Each product has a dedicated branch called after it. Every such branch is
based on some picked upstream kernel version which is immutable over time. On
the top of it, the branches contain from tens (master and stable branches) to
tens of thousands (SLE, openSUSE Leap) of patches we maintain in them. To be
able to apply the patches -- to define the order -- we also store their list
in a file called \file{series.conf}.

\begin{figure}[bt]
  \begin{center}
    \begin{tabular}{l}
      \begin{lstlisting}[language=,columns=fullflexible]
# latest standard kernel patches
patches.kernel.org/5.2.1-001-crypto-lrw-use-correct-alignmask.patch
patches.kernel.org/5.2.1-002-crypto-talitos-rename-alternative-AEAD-algos.patch
...
patches.kernel.org/5.2.6-020-Linux-5.2.6.patch

# Bug workarounds for binutils

# Scheduler / Core
patches.suse/setuid-dumpable-wrongdir
...
# VM/FS patches
patches.suse/vfs-add-super_operations-get_inode_dev
patches.suse/readahead-request-tunables.patch
...
      \end{lstlisting}
    \end{tabular}
  \end{center}

  \caption{Example series file of stable.}
  \label{fig:series_stable}
\end{figure}

In master and stable branches, these patches are placed into
\file{series.conf} in an order depending on a person who did the backport
(cf.\ Fig.\ \ref{fig:series_stable}). The patches are divided into so called
\emph{sections} there. There are sections for filesystems, x86, TTY, sound,
and alike. For tens of patches it is quite bearable to maintain the patches in
this form. And we also used to use this schema for SLE before SLE 12 SP2.
However, lately, we switched to a \emph{sorted section}.

Sorted section is a section in \file{series.conf} where all the patches are
sorted according to their appearance in the upstream kernel (see Fig.\
\ref{fig:series_sle15}). This makes lives of developers easier by the order
of magnitude. One reason are tools we have to maintain series
file now~\cite{sort_series}. Another main reason is that there are no
interdependencies between patches which would be placed in different sections
with the old schema. These dependencies sometimes created large conflicts when
applying patches.  Therefore, it also made the patches to differ from the
upstream ones.  This made the situation even worse with more and more patches
coming (and differing from upstream).

One exception from this sorted-section rule are stable patches. In upstream,
for each kernel release, some later minor releases with non-intrusive fixes
are released. So for example, for a released kernel 4.12, we have 4.12.1,
4.12.2, etc. These stable releases are composed of many patches and we add
them in a single-commit-per-file manner (like any other patch). For all
branches master, stable, SLE, and Leap, we put these stable patches to the
very front of the series file. And their order is defined by the stable
upstream.  Obviously, this creates exactly the same interdependencies as
non-sorted sections schema described earlier. Note that other (non-sorted)
sections were still left in place even in the new schema. They are used for
patches which were not accepted upstream yet. One example is
\file{setuid-dumpable-wrongdir} patch which is present in both stable and
SLE15 as can be seen in both Figures\ \ref{fig:series_stable} and
\ref{fig:series_sle15}.

Note that stable releases for a particular kernel are not released infinitely.
SLE15 and SLE15-SP1 is based on 4.12. But only fourteen stable versions were
released for 4.12. These were put to the SLE15 branch to the front of its
series file. After that, we only add some patches collected from the upstream
stable list along with those reported by our tools like
git-fixes~\cite{git_fixes}.  All these are put to the sorted section, despite
they have the same origin and reasoning as stable patches from stable releases
before. In other words, we continue collecting stable patches, but we do put
them to the sorted section, not to the front of series file.

\begin{figure}[bt]
  \begin{center}
    \begin{tabular}{l}
      \begin{lstlisting}[language=,columns=fullflexible]
# latest standard kernel patches
patches.kernel.org/4.12.1-001-driver-core-platform-fix-race.patch
patches.kernel.org/4.12.1-002-RDMA-uverbs-Check-port-number.patch
...
patches.kernel.org/4.12.14-054-Linux-4.12.14.patch

# Non-upstream fixes for stable patches
patches.suse/mlxsw-spectrum-Forbid-linking-to-devices-fix.patch
...
# sorted patches
patches.drivers/0001-uuid-afs-move-struct-uuid_v1-back-into-afs.patch
patches.drivers/0002-xfs-use-uuid_copy-helper-to-abstract-uuid_t.patch
... over 20000 patches are here, sorted by the upstream appearance
patches.kabi/kabi-fix-s390-vtime-steal-time-exponential.patch
# end of sorted patches

# Scheduler / Core
patches.suse/setuid-dumpable-wrongdir
...
      \end{lstlisting}
    \end{tabular}
  \end{center}

  \caption{Example series file of SLE15.}
  \label{fig:series_sle15}
\end{figure}

\section{Proposal} \label{sec:proposal}
SLE15-SP2 will be based on a fresh upstream kernel. Likely on 5.3. Here, we
propose that every single stable patch that is released in any later 5.3.x
stable release would be placed to the sorted section too. So we will stop
including stable releases as they are, and will take only SHA identifiers of
the commits. Upstream patches with these SHAs will be taken instead. The same
route as any other patch we backport. As well as putting them to the sorted
sections. This will further reduce the burden as interferences with other
patches will be quite lower.

\section{Discussion} \label{sec:discussion}
With the proposal above, there are few open questions which we discuss here.
In particular, how to generate kernel-vanilla now, directory for placing the
stable patches, and minor numbering of our kernel.

\subsection{Vanilla Flavour}
For every SLE or openSUSE, we also provide users with a so-called vanilla
kernel (named kernel-vanilla). It is a kernel released by upstream. So for
SLE15, that would be pure 4.12.14 with few additional patches required for
proper packaging. This flavour can be useful for customers. If they find a bug
in our kernel, we can ask them whether it is an upstream issue by checking the
presence of the bug in kernel-vanilla. If it is not there, then it is a bug in
our later backports, e.g. in the sorted section.

Currently, kernel-vanilla is generated such that:
\begin{itemize}
  \item stable patches are taken from \file{patches.kernel.org} directory and
  \item patches fixing the build are taken from \file{patches.rpmify}
    directory.
\end{itemize}
Only these are combined and source code built. Remember that both of them are
applied first for any flavour (order in series file). If we put patches coming
from stable into the sorted section, this will no longer be possible.

There are two solutions to this: (a) do not build kernel-vanilla at all, and
(b) build kernel-vanilla differently from other flavours. As per (a), it is
unclear whether kernel-vanilla is of any use in the first place. This shall be
discussed further.

As per (b), kernel-vanilla would be built using upstream stable kernel with the
stable patches we applied to the sorted section. On the top of it, we would
apply \file{patches.rpmify}.  Here, we risk build breakage as
\file{patches.rpmify} might not apply cleanly any more.

\subsection{Directory for Stable Patches}
It is only a technicality, but since stable patches are now in the sorted
section, it makes a little sense (only for documentation purposes) to put the
patches to \file{patches.kernel.org} directory. They can be among other
patches in \file{patches.suse}. Or not?

\subsection{Minor Numbering}
In stable releases, there are special patches (like
\file{4.12.14-054-Linux-4.12.14.patch} from Fig.\ \ref{fig:series_sle15})
which only increase a minor release number (here, from 4.12.13 to 4.12.14). We
are not increasing minor numbers in SLE15 and SLE15-SP1, despite we apply
stable patches. So we do not necessarily need to increase minor numbers and
can skip these patches. We can do so only for consistency purposes. This is
also open to discussion.

\section{Conclusion} \label{sec:concl}
We proposed adding stable patches to the sorted sections and discussed what it
brings and what needs to be considered when doing so. There are still open
questions to be discussed during the presentation.

\bibliographystyle{plain}
\bibliography{paper}

\end{document}
