% vim: tw=78 ai spell spelllang=en_gb sw=2
% syntax spell toplevel

\documentclass[10pt]{llncs}
\usepackage{color}
\usepackage{colortbl}
\usepackage{url}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{tikz}
\usepackage{xspace}
\pagestyle{plain}

\usetikzlibrary{arrows, positioning, decorations.pathreplacing}

\lstset{language=C,tabsize=2,showstringspaces=false,
	breaklines=true,breakatwhitespace=true,escapeinside={(*@}{@*)},
	captionpos=b,basicstyle=\sffamily\small}
%numbers=left,frame=l

\newcommand{\code}[2][]{\lstinline[columns=fullflexible, #1]$#2$}
\newcommand{\codeasm}[2][]{\code[language={[x86masm]Assembler},
	morekeywords={rbp,rsp},#1]{#2}}
\newcommand{\file}[2][]{\code[language={}, #1]{#2}}
\newcommand{\commit}[1][]{\texttt{#1}}

\newcommand{\xes}{\texttt{x86}\xspace}
\newcommand{\xesl}{\texttt{x86\_32}\xspace}
\newcommand{\xesh}{\texttt{x86\_64}\xspace}

\newcommand{\dwarf}{\textsc{Dwarf}\xspace}
\newcommand{\gcc}{\textsc{Gcc}\xspace}
\newcommand{\objtool}{\textsc{Objtool}\xspace}
\newcommand{\orc}{\textsc{Orc}\xspace}

\setcounter{tocdepth}{1}

\begin{document}

\title{Unwinding the State of Unwinder}
\author{Jiri~Slaby}
\institute{Suse Labs \\
\email{jslaby@suse.cz}}

\maketitle

\begin{abstract}
  We have been using our own implementation of \dwarf unwinder in every SUSE
  and openSUSE release in the past decade. First, it makes the kernel faster
  by approximatelly 10\,\% as it removes the need for enabled frame pointers.
  Second, it prunes the call traces from false positives as it is precise
  opposing to frame pointers. Unfortunately, despite attempts to make this
  upstream, this was an out-of-tree solution most of the time.  And given the
  progress in the live patching area in upstream, there is a call for some
  reliable unwinder in upstream now too.  This paper describes the path of our
  \dwarf unwinder to upstream, its rejection twice, proposed solution, and
  an implementation which finally went upstream. And thanks to the good
  timing, this implementation will be a part of the upcomming SUSE Linux
  Enterprise 15 too.
\end{abstract}

\section{Introduction} \label{sec:intro}
\begin{figure}[bt]
  \begin{center}
%RIP: 0010:crash_now+0xae/0x170 [crasher]
%RSP: 0018:ffffb82980d1bc80 EFLAGS: 00010282
%RAX: 0000000000000023 RBX: 0000000000000000 RCX: 0000000000000000
%RDX: ffff902dd3296888 RSI: ffff902dd328e568 RDI: ffff902dd328e568
%RBP: ffffb82980d1bc88 R08: 0000000000000000 R09: 0000000000000000
%R10: 0000000000000000 R11: 0000000000000000 R12: ffffffffc04a8000
%R13: ffff902dcb5c8e20 R14: 0000000000000001 R15: ffff902d7580f880
%FS:  00007fa4fa242040(0000) GS:ffff902dd3280000(0000) knlGS:0000000000000000
%CS:  0010 DS: 0000 ES: 0000 CR0: 0000000080050033
%CR2: 0000000000000000 CR3: 0000000089138000 CR4: 00000000000006e0
%Call Trace:
    \begin{tabular}{c}
      \begin{lstlisting}[language=,columns=fullflexible]
crasher_init+0x4d/0x1000 [crasher](*@\tikz[remember picture,baseline=-.5ex]{\node [coordinate] (LSTSTMODINIT) {};}@*)
(*@\tikz[remember picture,baseline=-1.5ex]{\node [coordinate] (LSTSTINITTOP) {};}@*)do_one_initcall+0x50/0x1a0
do_init_module+0x5f/0x1f0
(*@\tikz[remember picture,baseline=0ex]{\node [coordinate] (LSTSTINITBOT) {};}@*)load_module+0x1694/0x1ea0
(*@\tikz[remember picture,baseline=-1.5ex]{\node [coordinate] (LSTSTMODPTOP) {};}@*)SYSC_finit_module+0xd7/0xf0
(*@\tikz[remember picture,baseline=0ex]{\node [coordinate] (LSTSTMODPBOT) {};}@*)SyS_finit_module+0xe/0x10
entry_SYSCALL_64_fastpath+0x23/0xc2(*@\tikz[remember picture,baseline=-.5ex]{\node [coordinate] (LSTSTSYSCALL) {};}@*)
      \end{lstlisting}

  \tikzstyle{ptr} = [->,>=stealth',shorten > =3pt]
  \begin{tikzpicture}[remember picture,overlay,node distance=1.5em]
    \node [right=of LSTSTMODINIT] { module's init function }
      edge[ptr] (LSTSTMODINIT);

    \node [coordinate,right=4.4cm of LSTSTINITTOP] (ITOPSHIFTED) {};
    \node [coordinate,right=4.4cm of LSTSTINITBOT] (IBOTSHIFTED) {};
    \draw[-,thick,decorate,decoration={brace,amplitude=4pt}] (ITOPSHIFTED) -- (IBOTSHIFTED)
      node[midway] (MID) {};
    \node [right=1mm of MID] { load and init module };

    \node [coordinate,right=4.7cm of LSTSTMODPTOP] (MTOPSHIFTED) {};
    \node [coordinate,right=4.7cm of LSTSTMODPBOT] (MBOTSHIFTED) {};
    \draw[-,thick,decorate,decoration={brace,amplitude=3pt}] (MTOPSHIFTED) -- (MBOTSHIFTED)
      node[midway] (MID) {};
    \node [right=1mm of MID] { \code{modprobe} system call };

    \node [right=of LSTSTSYSCALL] { system call }
      edge[ptr] (LSTSTSYSCALL);
  \end{tikzpicture}
    \end{tabular}
  \end{center}

  \caption{Example of a call trace. Read from bottom to top.}
  \label{fig:call_trace}
\end{figure}

Vast majority of us already met a kernel crash, \textit{BUG}, or at least a
kernel \textit{WARNING}. All such reports produced by the kernel are
accompanied with sequence of functions which were visited up to the point
of failure. Among other data, this is an important piece of information that
helps developers to understand the issue. Without that, such a report would be
mostly helpless. This sequence of functions is called a \emph{call trace} and
for demonstration, we attach one in Fig.~\ref{fig:call_trace}.  A developer
can take reports like these and start debugging the issue using disassembler,
\textsc{Gdb}, debug-enabled kernels, and so on. As we can see, there are also
numbers after the plus sign in the output. These are offsets (of total
function size after the slash) in the functions, so that developers can
precisely track which \codeasm{call}s were invoked in every function of the
chain.

\subsection{Frames and (Optional) Frame Pointers} \label{sec:fp}
There are many approaches on where and how to extract call trace entries.
In the kernel, we use the fact that for every \codeasm{call} to a function, a
CPU stores a \emph{return address} to the stack before yielding the execution
to the function. Similarly, every \code{return} causes the CPU to read this
address and remove it from the stack. Every \codeasm{call} is said to create a
\emph{stack frame} for a particular function call. And every \code{return}
destroys it.  There is an example of a program and its stack in
Fig.~\ref{fig:stack_frame}. On the left side, there are three functions which
cause the program to crash by \code{abort} eventually.  When that is about to
happen -- right before the call to \code{abort} itself -- the possible state
of the stack is depicted on the right side.

\begin{figure}[tb]
  \begin{center}
      \tikzstyle{ptr} = [->,>=stealth',shorten > =3pt,semithick]
      \tikzstyle{border} = [thick]
      \tikzstyle{leftbrace} = [-,semithick,decorate,
	decoration={brace,mirror,amplitude=5pt,raise=2mm}]
      \tikzstyle{lefttext} = [left=7pt of MID,anchor=south,rotate=90,
	align=center]
      \newlength{\hw}
      \setlength{\hw}{1.7cm}
      \begin{tikzpicture}[node distance=2mm]
	\node (PRE) { \vdots };

	\node[below=of PRE,align=center] (CFA1) { return address \\
	  from \code{fun1} (to \code{main}) };
	\node[below=of CFA1] (SRBP1) { saved \code{BP} of \code{main}};
	\node[below=of SRBP1] (X) { variable \code{xx} };

	\node[below=of X,align=center] (CFA2) { return address \\
	  from \code{fun2} (to \code{fun1}) };
	\node[below=of CFA2] (SRBP2) { saved \code{BP} of \code{fun1}};
	\node[below=of SRBP2] (Y) { variable \code{yy} };

	\node[below=of Y,align=center] (POSTCOM) { stack grows here \\
	  up to \code{abort}};
	\node[coordinate,below=6mm of POSTCOM] (POST) {}
	  edge[ptr,<-,shorten > =0pt,shorten < = 2pt] (POSTCOM);

	\node [left=\hw of PRE.north] (TL) {};
	\node [right=\hw of PRE.north] (TR) {};
	\node [below left=1mm and \hw of POST.south] (BL) {};
	\draw[border] (TL) -- (BL) node[midway] (LCEN) {};
	\node [below right=1mm and \hw of POST.south] {}
	  edge[border] (TR);

	\foreach \from/\to/\left/\right/\line in
	    {PRE/CFA1/MAINFUN1/CALLFUN1/thick, CFA1/SRBP1/D1/D2/dashed,
	    SRBP1/X/D1/ORBP/dashed, X/CFA2/FUN1FUN2/CALLFUN2/thick,
	    CFA2/SRBP2/D1/D2/dashed, SRBP2/Y/D1/RBP/dashed,
	    Y/POSTCOM/FUN2ABORT/RSP/thick} {
	  \path (\from) -- (\to) node[midway] (MID) {};

	  \node [coordinate,left=\hw of MID] (\left) {};
	  \node [coordinate,right=\hw of MID] (\right) {}
	    edge[semithick,\line] (\left);
	}

	\node[right=7mm of RBP] { \code{BP} }
	  edge[ptr] (RBP);
	\node[right=7mm of RSP] { \code{SP} }
	  edge[ptr] (RSP);
	\node[above right=-1mm and 11mm of CALLFUN1,align=center] { called \\
	  \code{fun1} }
	  edge[ptr,dashed] (CALLFUN1);
	\node[above right=-1mm and 10mm of CALLFUN2,align=center] { called \\
	  \code{fun2} }
	  edge[ptr,dashed] (CALLFUN2);
	\node[below right=8mm and 6mm of RSP,align=center] {
	  \code{abort} to \\ be called }
	  edge[ptr,dashed] (RSP);

	\draw (SRBP2.east) edge[ptr,in=0,out=30] (ORBP);

	\node[coordinate,above=23mm of ORBP] (TMP) {};
	\draw (SRBP1.east) edge[ptr,in=0,out=30] (TMP);

	\draw[leftbrace] (TL) -- (MAINFUN1) node[midway] (MID) {};
	\node[lefttext] { \code{main}'s \\[-.4ex] frame };

	\draw[leftbrace] (MAINFUN1) -- (FUN1FUN2) node[midway] (MID) {};
	\node[lefttext] { \code{fun1}'s \\[-.4ex] frame };

	\draw[leftbrace] (FUN1FUN2) -- (FUN2ABORT) node[midway] (MID) {};
	\node[lefttext] { \code{fun2}'s \\[-.4ex] frame };

	\draw[leftbrace] (FUN2ABORT) -- (BL) node[midway] (MID) {};
	\node[lefttext] { so far \\[-.4ex] unused };

	\node[left=3cm of LCEN] (CODE) {
	  \begin{lstlisting}
int main()
{
  fun1();
}

void fun1(void)
{
  long xx;

  fun2();
}

void fun2(void)
{
  long yy;

  abort();(*@ \normalfont $\leftarrow$ we are here@*)
}
	  \end{lstlisting}
	};

      \end{tikzpicture}
  \end{center}

  \caption{Example of program and its possible stack. The stack grows down, so
    the events happened from the top to bottom (as well as the code).
    \code{SP} = current stack pointer, \code{BP} = current base/frame
    pointer.}
  \label{fig:stack_frame}
\end{figure}

More precisely, when \code{fun1} is called from \code{main}, the address of
the instruction right after the \codeasm{call} is stored on the stack to know
exactly where to return. And \code{fun1} now continues the execution.
Traditionally, the first thing a function performs is (a) storing the caller's
base/frame pointer (\code{BP}) to the stack (for later restore), and (b)
making \code{BP} to point to \code{SP} -- to the current ``bottom'' of the
stack, i.e.\ the current frame.  This setup happens in a so-called
\emph{prologue}.  Further, given \code{fun1} also contains a local variable
\code{xx}, which is to be allocated on the stack, this is done now too. Then
there is a call to \code{fun2} and all these steps are repeated until the
\code{abort}'s body causes the program to exit.

If \code{fun2} ever returned, the \code{fun1}'s function body would be over
and code at the very end of the function (\emph{epilogue}) would set \code{SP}
back to the value in \code{BP}. I.e.\ to a value which was there after the
prologue.  Finally, the function would restore the old (caller's) value of
\code{BP} and invoke \code{return}. This effectively means pop the return
address off the stack and jump there.

Conclusively, \code{BP} is somewhat static inside a function and changes only
by \codeasm{call}s and \code{return}s. Exactly, \code{BP} points to the
current frame plus 2 entries: return address and caller's/saved \code{BP}.
This renders \code{BP} to be only a \emph{convenient} way of restoring
\code{SP} in the epilogue.  If the code properly tracks what it performs with
the stack throughout
the function body, it can easily restore the stack state (\code{SP}) in the
epilogue based only on the tracked operations.  And pushes to and pops from
the stack are relatively easy to count, especially when the code is generated
by a compiler. Hence, when the code is compiled by \gcc with optimizations
turned on, by default, \gcc does not use \code{BP} as a frame/base pointer
storage.  This behaviour can be disabled by \code{-fno-omit-frame-pointer}, if
needed, however. And the kernel does so on many configurations,
leaving frame pointers in place.

Now, consider that both \code{SP} and \code{BP} are usually stored in
registers. On \xesh, these are \codeasm{rsp} and \codeasm{rbp}
(\codeasm{esp}/\codeasm{ebp} on \xesl), respectively. If we avoid using
\code{BP} for stack tracking, we free up one general purpose register for
arbitrary use.  This indeed leads to a speed-up, see
Section~\ref{sec:unw_fp} for more information.

\subsection{Tail calls} \label{sec:tail_calls}
When the code in Fig.~\ref{fig:stack_frame} is compiled by an optimizing
compiler, there are not two actual \codeasm{call}s and two \code{return}s.
\code{main} still calls \code{fun1} the usual way. But since the invocation of
\code{fun2} is the last thing to do in \code{fun1}, \code{fun1} will not do
this by a \codeasm{call}, but by a \code[morekeywords=jump]{jump}. This means
that no new stack frame is created for \code{fun2}. Therefore, \code{return}
in \code{fun2} will pop the return address from the \code{fun1}'s stack frame,
i.e.\ returns directly to \code{main}. This kind of optimization is called
\textit{tail calls} or \textit{sibling calls} and avoids one call frame
setup+destroy.  This is very common in the code and stack unwinders have to
take this into account when creating call traces.

\subsection{Use of Call Traces and Our Motivation} \label{sec:uses}
Crashes (including BUGs, WARNINGs, and similar) are obvious users of call
traces. They print out the information so that developers can debug the
problem. There are also other users like \textsc{FTrace} (function tracer),
\textsc{KMemLeak} (kernel memory leak detector), \textsc{KMemCheck} (kernel
memory accesses checker), or \textsc{LockDep} (locking issues detector).  They
all do use the call traces to store information about locations where function
calls, memory accesses, allocations, and locks take place respectively. These
traces are reported to the user only if the user request them or some problem
is detected by the tools. And if the latter happens, it is very useful to know
where some memory access or lock took place retrospectively.

We can consider the all above as for debugging purposes. There is one more not
so obvious user of call traces: \textit{Kernel Live Patching} (KLP). Although
we started providing live patching since SUSE Linux Enterprise 12 (SLE), we
have not been using stack traces for anything until the upcoming release:
SLE15. With SLE15, we are very likely to provide the upstream solution which
evolved in the meantime. And it utilizes call traces to check if a process is
somewhere in a function that is to be patched. Including those up to the call
chain towards userspace. Hence we are very interested in fast and precise
call traces too.

\section{Unwinding} \label{sec:unwinding}
Now that we know the stack contains stack frames with return addresses,
stored base pointers (sometimes), and other potentially useful information, we
can make use of it. The tools to extract stack contents and build call traces
are called \emph{stack unwinders}.  All unwinders described in this section
work similarly. They somehow scan the stack and return a call trace. It is
important to note, that there is not a single stack in the kernel. Whenever an
interrupt or exception occurs, the stack is switched to the interrupt, NMI,
exception, or other stack. So in fact, not only a single stack is scanned, but
all of them are scanned up to the point where userspace invoked the kernel
e.g.\ by syscall. Nevertheless, we consider a single stack in the text below
for simplicity.

\subsection{Simple Approach} \label{sec:unw_guess}
The simplest method used in the kernel is to scan the stack by 8-byte chunks
and put every value which is a function address to the returned call trace. A
function address is easy to be identified in the kernel. It is an address in
well-defined address ranges: kernel code, modules, and other minor code. There
is even a helper to decide this. It is called
\code{\_\_kernel\_text\_address}.

This method is complete in the sense that it surely reports all really visited
functions to the resulting call trace. To prove it, it is enough to realize,
that the whole stack is scanned and the stack contains return addresses.  And
of course they are reported as well as other random material found on the
stack which by coincidence fits in the kernel code space too.

This approach is used whenever all of the sophisticated methods described
below are unavailable/disabled. Since this is the most simple and naive
unwinder, it is also the fastest one.  But \emph{least precise}, of course.
The implementation can be found in \file{arch/x86/kernel/unwind\_guess.c}.

\subsection{Frame Pointers Approach} \label{sec:unw_fp}
When base/frame pointers are enabled (\code{CONFIG\_FRAME\_POINTER} turned
on), they can be used to identify the start of the current stack frame.
Afterwards, the unwinder can easily locate a return address in them precisely.
Since the pointer to the previous stack frame is stored right next to the
return address in the current frame (cf.\ Fig.~\ref{fig:stack_frame}), the
unwinder uses the pointer to unwind one level up. Then it extracts another
return address from there. And another frame pointer. This repeats until the
bottom of the stack is reached.

There is a run-time \emph{penalty} stemming from enabled frame pointers.
Functions have to contain prologues and epilogues to maintain the frame
pointer (\code{BP}). It means (a) longer execution time needed, and (b) more
instruction cache misses due to larger code.  The overall slowdown measured is
up to 10\,\% -- depending on the load~\cite{fp_slowdown}.  The code is larger
about 3.2\,\%~\cite{orc_doc}.  The implementation is in
\file{arch/x86/kernel/unwind\_frame.c} and is still relatively simple.

For the unwinder to be correct, every function has to handle the frame pointer
correctly. Especially assembler coders have to write the prologues and
epilogues explicitly. To ensure correctness for all functions, there is a tool
called \objtool. It walks all functions and reports any inconsistencies in
frame pointer handling, including completely omitted handling.

Even if every function contains correct frame pointer handling, the unwinder
still can miss an entry. Imagine that a function \code{A} calls a function
\code{B}. A return address from \code{B} to \code{A} is pushed on the stack.
But if an interrupt comes at this very moment, i.e.\ before a prologue of
\code{B} is executed, unwinder has hard times to unwind. It is because the
return address is on the stack, while the frame pointer is not yet. It is
feasible to deal with this situation in most cases. The kernel currently
cannot handle interrupted functions which are the first on the stack -- e.g.\
the syscall handler.

\subsection{\dwarf Approach} \label{sec:unw_dw}
There was an ongoing effort to use unwinding in the Kernel Live Patching to
conclude no task is in a patched function. While there were successful
attempts to make frame-pointer-based unwinding reliable, it still drags in
the run-time overhead and imprecisions described earlier. Hence, we also
proposed on the Linux kernel mailing list to use \dwarf-based unwinding in the
kernel~\cite{patch_add_dwarf}.

Note that the unwinder was already a part of the Linux kernel since
2.6.18~\cite{git_dwarf_upstream_add}. But it was dropped by Linus later in
2.6.20~\cite{git_dwarf_upstream_rm} due to various issues. Especially, the
code was not so nice and readable and it also depends on generated data, which
were allegedly not always correct. Noteworthy, we have actually never seen a
single issue during our use of the unwinder in SUSE. Nonetheless, our newly
submitted cleaned-up version tried to address most of the problems. We also
suggested to use userspace programs during build to check and correct the data
if they were not correct. This approach was more or less accepted, until the
new \orc unwinder was proposed. See Section~\ref{sec:unw_orc} for the details.

\dwarf~\cite{dwarf} works on a completely different basis than frame pointers.
It needs no additional in-code operations.  Instead, when a stack is to be
unwound, this is performed solely on the out-of-band data. Look at the example
in Fig.~\ref{fig:dwarf}.  The assembly code in ``(b)'' saves the previous base
pointer to the stack and replaces it with the current one. Then the code
reserves \code{0x10} (i.e.\ 16) bytes from the stack, moves \code{2} to
variable \code{xx} (on stack) which is right below the base pointer.
Afterwards, \code{fun2} is called. When the CPU returns back to \code{fun1},
the reserved stack space is removed by the \codeasm{leave} instruction. It is
equivalent to restoring \codeasm{rsp} from \codeasm{rbp} and popping the old
value of \codeasm{rbp} from stack, but faster. This indeed very corresponds to
the stack content that we depicted earlier in Fig.~\ref{fig:stack_frame}.

\begin{figure}[tb!]
  \begin{center}
    \begin{tabular}{cp{4em}c}
      \begin{lstlisting}
void fun1(void)
{
  long xx = 2;

  fun2();
}
      \end{lstlisting}
      &&
      \begin{lstlisting}[language={[x86masm]Assembler}]
<fun1>:
 0:  push   %rbp
 1:  mov    %rsp,%rbp
 4:  sub    $0x10,%rsp
 8:  mov    $0x2,-0x8(%rbp)
10:  call   fun2
15:  nop
16:  leave
17:  ret
      \end{lstlisting} % stopzone
      \\
      (a) C code && (b) Assembly (offsets in hexa) \\
    \end{tabular}
      \\[1ex]
    \begin{tabular}{c}
      {
	\begin{lstlisting}[language=,numbers=left]
DW_CFA_def_cfa: r7 (rsp) ofs 8
DW_CFA_offset: r16 (rip) at cfa-8
DW_CFA_advance_loc: 1 to 0000000000000001
DW_CFA_def_cfa_offset: 16
DW_CFA_offset: r6 (rbp) at cfa-16
DW_CFA_advance_loc: 3 to 0000000000000004
DW_CFA_def_cfa_register: r6 (rbp)
DW_CFA_advance_loc: 19 to 0000000000000017
DW_CFA_def_cfa: r7 (rsp) ofs 8
	\end{lstlisting} % stopzone
      }
      \\ \\
      (c) \dwarf data (line numbers on the left)
    \end{tabular}
      \\[1ex]
    \begin{tabular}{c}
      {
	\begin{lstlisting}[language=]
.text+0:  sp:sp+8  bp:(und)     type:call
.text+1:  sp:sp+16 bp:prevsp-16 type:call
.text+4:  sp:bp+16 bp:prevsp-16 type:call
.text+17: sp:sp+8  bp:(und)     type:call
.text+18: sp:(und) bp:(und)     type:call
	\end{lstlisting}
      }
      \\ \\
      (d) \orc data
    \end{tabular}
  \end{center}
    \caption{Demonstration of \dwarf and \orc data. \dwarf was generated by
    \gcc from the above C code. It was dumped by \code{readelf -wf} and
    slightly modified for simplicity. \orc was generated by \code{objtool orc
    generate} and dumped by \code{objtool orc dump}.  The assembly generated
    by \gcc is also above and is an output of \code{objdump -d}, altered a
    bit.}
  \label{fig:dwarf}
\end{figure}

Now have a look at the \dwarf data in Fig.~\ref{fig:dwarf} ``(c)''. The data are
encoded \dwarf instructions~\cite{dwarf_insn} and every function has assigned
a block of such instructions. An instruction changes the state of a \dwarf
state machine.  Every line of the assembly can have attached some state of the
state machine.  The machine keeps states of registers at that location,
especially the stack pointer (\codeasm{rsp}). If we look closer at the Figure,
we can spot \code{cfa} there. In the language of \dwarf, it denotes a
\emph{call frame address} and is considered to be another name for the start
of the stack frame. Further, the instructions are supposed to be read in a
top-down fashion.  So we are at the start of the function and the first line
of the DWARF data (\code{def\_cfa}) defines that \code{cfa} equals to the
stack pointer with offset 8 (\codeasm{rsp+8}).  That is exactly right before
the stored return address as we also learnt earlier.  The next line says, that
the value of \codeasm{rip} is stored on the stack 8 bytes after \code{cfa}.
These two lines are common for every standard function. And again, they comply
to earlier Fig.~\ref{fig:stack_frame}.

That was all to be done at the beginning of the function. Now, the line three
instructs us to move to the next line in assembly (\code{advance\_loc}), to
offset 1. At this point, there is a \codeasm{push} which stores something
(\codeasm{rbp}) to the stack. Hence, \dwarf data at line 4 have to tell us
that \code{cfa} offset changed and is now 16 from the current stack pointer
(\codeasm{rsp}).  Coincidently, the \codeasm{push} stored \codeasm{rbp} on the
stack, so line 5 also says that the value of this register can be found at
\code{cfa-16}. The next two \dwarf lines (6--7) force us to use \codeasm{rbp}
as \code{cfa} from this position on, as (a) \code{rsp} was subtracted to
allocate some space, and (b) we apparently have frame pointers turned on and
they are exactly in \codeasm{rbp}. When leaving from the function at assembly
line 17, \codeasm{rbp} was restored by \codeasm{leave} and we should start
using \codeasm{rsp} as \code{cfa} again -- according to \dwarf lines 8 and 9.

So the task of the \dwarf unwinder is always read this data and update the
\dwarf state machine as we did in the previous paragraphs. Finally, the return
address is read from the stack from \code{cfa} with the computed offset. The
unwinder performs this for every function on the call trace. The unwinder
starts in the current function and ``executes'' the \dwarf instructions up to
the actual CPU's \textit{instruction pointer} (\code{rip}) register. Then it
reads the return address from the stack and repeats these steps from the start
of a function containing the address up to the address proper. This happens
until the bottom of the stack is reached.

Last but not least, the \dwarf data are stored in an ELF section called
\code{.eh\_frame}. Originally, this \textit{exception handling} section used
to serve for C++ exceptions, but was later extended for use in C for
unwinding.  There can also be an index of unwinding data in another section
called \code{.eh\_frame\_hdr}.  It speeds up the lookup of data for an address
in some function. The format of both of them is specified in LSB Core --
Generic~\cite{LSB_generic}. Our vmlinux from Tumbleweed's kernel-default has
1.78\,MiB of the data and 290\,KiB of the index. For example, the i915 module
contains 178\,KiB of the data and no index.

\paragraph{Drawbacks} \dwarf needs a compiler to generate the data. It is fine
for C -- \gcc handles that perfectly -- but manual annotations are needed for
assembly. And doing so manually is error-prone as was proven by the history.
As one example, see commit \commit{e0a5a5d9b006} (``x86, 64-bit, dwarf2: push
pushes 8 bytes and popf pops 8'') by Alexander van Heukelum. So despite the
annotations were part of the kernel even though the unwinder was not, they
were dropped for good in commit \commit{131484c8da97} (``x86/debug: Remove
perpetually broken, unmaintainable dwarf
annotations'')~\cite{git_dwarfinfo_upstream_rm}.  Since then, the \dwarf
unwinder cannot proceed though interrupts, system call handlers, and other
code written in assembly.

The implementation can be found in some our new kernels in
\file{kernel/dwarf.c} (the unwinder) and
\file{arch/x86/kernel/unwind\_dwarf.c} (a glue for the new API below). In sum,
they are 1917 lines of code.

\subsection{Rewrite of Unwinding -- the New API Upstream} \label{sec:rewrite}
While we (at SUSE) have been using the \dwarf unwinder plugged into the
upstream code for a long time, the interface was not nice and maintaining
another unwinder was quite hard. Therefore, the unwinding in the upstream
kernel was completely redesigned in 4.9 by commit \commit{7c7900f89770}
(``x86/unwind: Add new unwind interface and
implementations'')~\cite{git_unw_rewrite}. It allows for quite easy switching
from one unwinder to another. And this cleaned-up API also prepared the field
for easy use by KLP too. So now, there are three kinds of call traces from the
new API's point of view:
\begin{enumerate}
  \item In case of crash, \textit{BUG}, and similar, a call trace is dumped to
    the log. It should contain all available data from the stack. So similar
    to the Guess unwinder in Section~\ref{sec:unw_guess}, everything what is
    on the stack and is a kernel code address is printed.  This happens
    despite more precise unwinders might be available in the kernel at that
    moment. Nevertheless, if they are, the entries reported by the more
    precise unwinder are not accompanied by the `\code{?}' mark in the output.
    So in most cases, developers may considered entries with the `\code{?}'
    mark as irrelevant and can ignore them.  The reason for this behaviour is
    that unwinders might have troubles to unwind for some unpredictable
    reasons, while we still want to dump at least some call trace. This
    dump-all behaviour is invoked via \code{dump\_stack} and derivatives.
  \item \textsc{FTrace}, \textsc{KMemCheck}, \textsc{KMemLeak},
    \textsc{LockDep}, and similar tools save the stack trace for their use.
    Unlike for the previous kind, these traces contain only entries which are
    reported by the configured unwinder.  This kind of call traces is obtained
    by \code{save\_stack\_trace}.
  \item Kernel Live Patching demands only reliable (precise) call traces.  But
    unwinders sometimes put imprecise entries into call traces in the previous
    kind. So here, when the unwinder is unsure about any entry, the whole call
    trace shall be discarded and an error returned instead of some imprecise
    call trace.  A new function \code{save\_stack\_trace\_tsk\_reliable} was
    introduced for this purpose.
\end{enumerate}

In the new API, there is a common ``core'' for stack unwinding as an
underlying layer for the above.  Below that, there are implementations of the
respective unwinders. However, the current API allows only one unwinder to be
enabled when compiling the kernel. In our opinion, this is no limitation and
if users want more than one unwinder to be enabled at the same time, this can
be relatively easy changed. Having more unwinders in parallel may be good for
performance reasons -- e.g.\ use frame pointers for \textsc{LockDep} and use
\dwarf or \orc for Live Patching.

The core requires the enabled unwinder to define only few functions. The
functions and their meanings are as follows:
\begin{itemize}
  \item \code{\_\_unwind\_start} -- initialize the unwinder along with its
    internal state. This is a preparation for unwinding. Usually, the state is
    set up according to the state of the process. It can be the currently
    running process, current process with pre-defined values of registers
    (program counter, stack pointer, etc.), or a sleeping process. Tasks
    running on other CPUs are usually ignored as the call stack changes very
    quickly and it is also not easy to grab a call stack of such process.
  \item \code{unwind\_next\_frame} -- unwind to the next frame and internally
    note for later use its return address, stack pointer, and possibly frame
    pointer. This is called repeatedly by the core until enough frames are
    unwound or the bottom of the stack is reached. This function implements
    the heart of each respective unwinder described in previous sections.
  \item \code{unwind\_get\_return\_address} -- extract the return address from
    the current frame and return it.
  \item \code{unwind\_get\_return\_address\_ptr} -- return a pointer to the
    location of the return address on the stack. This is used only for
    comparison with the current stack pointer when doing \code{dump\_stack} to
    the log. If they match, the entry is printed as reliable (without the
    `\code{?}' mark.)
  \item \code{unwind\_get\_entry\_regs} -- implementations should return a
    \code{pt\_regs} structure describing CPU registers. It is dumped to the
    log.  This is useful for example when there was an interrupt and we want
    to dump the state of registers before the interrupt happened.
    Implementations return this usually only when jumping from one stack to
    another in a particular unwinding step (interrupt $\rightarrow$ process
    stack switch and similar.)
\end{itemize}

\subsection{\orc Approach} \label{sec:unw_orc}
The rewrite of the unwinder API allowed for easy plugging of new unwinders.
And since \dwarf was despised by upstream~\cite{kill_dwarf2}, a new
unwinder inspired by \dwarf emerged. It was named \orc\ -- \textit{Oops Rewind
Capability}. It uses out-of-band data too, but they are created independently
of \dwarf. They are also substantially simpler. Hence \orc cannot handle
states of all registers, neither complex state machines.  But the basic
functionality to track the stack pointer and call frames is included and is
actually enough for the Linux Kernel.

As we mentioned for \dwarf, the unwind data has to be generated. \orc
implemented this functionality into \objtool. Before \orc, \objtool served
for checking frame pointers only (c.f.\ Section~\ref{sec:unw_fp}.) Since it
already knew about many of the stack operations, it could also generate data
for \orc. But it had to be modified (a) to remember all stack operations, not
only checking prologues and epilogues, and (b) to generate the data for \orc.
Similar to \dwarf, the data are generated to two ELF sections:
\code{.orc\_unwind} and \code{.orc\_unwind\_ip}.  The former contains an array
of \code{orc\_entry} structures. The latter contains the same count of
addresses associated with those structures. Both arrays are sorted by the
latter -- the address they are bound to.

For performance reasons, there exist a third section called
\code{.orc\_lookup}: fast lookup table. This one is created at runtime after
the entries above are sorted. It serves as a pseudo hash table and makes the
search in the large data faster by searching only in a small subset of the
arrays above.

\begin{figure}[htb]
  \begin{center}
    \begin{tabular}{c}
      \begin{lstlisting}
struct orc_entry {
  s16      sp_offset;
  s16      bp_offset;
  unsigned sp_reg:4;
  unsigned bp_reg:4;
  unsigned type:2;
};
      \end{lstlisting}
    \end{tabular}
  \end{center}
    \caption{Structure \code{orc\_entry} used by \orc for states assigned to
    assembly lines.}
  \label{fig:orc_struct}
\end{figure}

The format of the \orc structure is depicted in Fig.~\ref{fig:orc_struct}. The
structure contains information about values of stack (\code{SP}) and
frame pointers (\code{BP}) before the call to the current function. More
precisely which register to read and what offset to add to the read value.
This allows recovering the values at any point in the function to the
starting ones, i.e.\ to exactly unwind one level up. If we look back at the
\dwarf example in Fig.~\ref{fig:dwarf}, we also present the \orc data in
`(d)'.  We can see similarities of \orc to \dwarf. For example, the first line
says that when the function was called, the previous stack pointer
(\codeasm{rsp}) can be computed as \codeasm{rsp+8}. After the \codeasm{push},
the previous \codeasm{rsp} is current \codeasm{rsp+16} and so on. The
\code{type} field of the \code{orc\_entry} structure allows for
differentiation of standard calls from system calls, interrupts, and
exceptions. In the latter three, we have to extract the state of previous
stack and frame registers from the stack, not from registers.

The implementation of the generation part is in \file{tools/objtool/}. The
unwinder is in \file{arch/x86/kernel/unwind\_orc.c} and is only 582 lines long
(including comments.)

\paragraph{Drawbacks} While it is fast and reliable, it relies heavily on the
output of \gcc. \gcc generates some patterns of register stores and loads
which \orc was taught to understand. When these patterns change, \orc has to
be changed too. The alternative would be to parse the \dwarf data and convert
them to \orc.

\paragraph{Upstream Status} Parts of the \orc unwinder were accepted in the
upstream kernel already.  It was especially preparatory
work~\cite{git_stack_val2}. \orc data generation and the unwinder proper are
pending in the -tip tree~\cite{git_orc_gen,git_orc_unw}. They are scheduled
for 4.14.

\section{Unwinding in openSUSE \& SLE} \label{sec:suse_unw}
This sections is devoted to the state of unwinders in SUSE kernels as it
developed over time.  Until \dwarf unwinder was written and ready to use, we
had frame pointers enabled to have at least a bit reliable call traces.

Even after the \dwarf unwinder was written and was in upstream since 2.6.18
till 2.6.20, none of the SUSE products were released in that time-frame. After
it was dropped from upstream, the authors of the unwinder (Jan Beulich and
Andi Kleen) pushed it into our trees as
\commit{d7af6f6e8a34}~\cite{git_suse_dwarf_add}.  This happened around 2.6.24,
so the first products to have the \dwarf unwinder were \textit{SUSE Linux
Enterprise 11} (SLE11) and \textit{SUSE Linux 11.1} (SL-11.1) -- a predecessor
of openSUSE.

Despite the \dwarf unwinder itself was dropped from upstream, the manual
annotations of assembly were left in-place. So we could use the unwinder
also throughout interrupts, system calls etc. But upstream decided with commit
\commit{131484c8da97} (``x86/debug: Remove perpetually broken, unmaintainable
dwarf annotations'')~\cite{git_dwarfinfo_upstream_rm}, that they are
hard to maintain and dropped them. Since that moment, we were unable to unwind
through assembly properly. This commit was a part of 4.2, so it affected
SLE12-SP2. We had to enable frame pointers in that release again, which lead
to slowdown (cf.\ Section~\ref{sec:unw_fp}).

\section{Conclusion} \label{sec:concl}
\begin{table}[tb!]
  \begin{center}
    \begin{tabular}{|l||c|l|c|l|c|l|} \hline
      \multicolumn{1}{|c||}{\textbf{Union of Call Traces}} &
	\multicolumn{6}{|c|}{\textbf{Unwinder}} \\ \cline{2-7}
      \multicolumn{1}{|c||}{\textbf{from All Unwinders}} &
	\multicolumn{2}{|c|}{\textbf{Guess}} &
	\multicolumn{2}{|c|}{\textbf{Frame Pointer}} &
	\multicolumn{2}{|c|}{\textbf{\orc}} \\ \hline\hline
      \code{crash\_now}		 & R & 0x94  & R & 0x99  & R & 0x94 \\
      \rowcolor[gray]{.9}
      \code{crash\_timer\_cb}	 &   &       &   & 0xe   &  & \\
      \code{call\_timer\_fn}	 & ? & 0x2e  &   & 0x33  &  & 0x2e \\
      \rowcolor[gray]{.9}
      \code{crash\_now}		 &   &       & ? & 0x170 &  & \\
      \code{run\_timer\_softirq} & ? & 0x40d &   & 0x414 &  & 0x40d \\
      \rowcolor[gray]{.9}
      \code{RANDOM}		 & ? & 0x1f  & ? & 0x2b  & ? & 0x1f \\
      \rowcolor[gray]{.9}
      \code{RANDOM}		 & ? & 0x38  & ? & 0x1d  & ? & 0x38 \\
      \rowcolor[gray]{.9}
      \code{RANDOM}		 & ? & 0x21  & ? & 0xc8  & ? & 0x21 \\
      \code{\_\_do\_softirq}	 & ? & 0xde  &   & 0xde  &  & 0xde \\
      \code{irq\_exit}		 & ? & 0xae  &   & 0xb6  &  & 0xae \\
      \code{smp\_apic\_timer\_interrupt} & ? & 0x39 &  & 0x3d &  & 0x39 \\
      \code{apic\_timer\_interrupt} & ? & 0x82 &  & 0x89 &  & 0x82
      \\ \hline\hline

      \multicolumn{7}{|c|}{--- IRQ ---} \\ \hline\hline

      \code{native\_safe\_halt}	 & ? & 0x2   & R & 0x6   & R & 0x2 \\
      \code{default\_idle}	 & ? & 0x1a  &   & 0x20  &   & 0x1a \\
      \rowcolor[gray]{.9}
      \code{arch\_cpu\_idle}	 &   &       &   & 0xf   &   & \\
      \rowcolor[gray]{.9}
      \code{default\_idle\_call} &   &       &   & 0x23  &   & \\
      \code{do\_idle}		 & ? & 0x16a &   & 0x178 &   & 0x16a \\
      \code{cpu\_startup\_entry} & ? & 0x5d  &   & 0x62  &   & 0x5d \\
      \rowcolor[gray]{.9}
      \code{rest\_init}		 &   &       &   & 0x84  &   & \\
      \code{start\_kernel}	 & ? & 0x422 &   & 0x429 &   & 0x422 \\
      \rowcolor[gray]{.9}
      \code{early\_idt\_handler\_array} & ? & 0x120 & ? & 0x120 & ? & 0x120 \\
      \rowcolor[gray]{.9}
      \code{x86\_64\_start\_reservations} &  &  &  & 0x29 &  & \\
      \code{x86\_64\_start\_kernel} & ? & 0x12c &  & 0x134 &  & 0x12c \\
      \code{secondary\_startup\_64} & ? & 0x9f &  & 0x9f &  & 0x9f \\ \hline
    \end{tabular}
  \end{center}
    \caption{Precision of unwinders demonstrated by \code{modprobe crasher
    call\_warn}. The table demonstrates the Guess, Frame Pointer and \orc
    unwinders. The question mark (?) means non-reliable entry. ``R'' means
    this value was in a register. The numerical columns are reported offsets
    against the start of the respective function. It can be also seen that the
    offsets of Guess and \orc unwinders are the same. I.e. \orc does not add
    any code anywhere, it relies only on out-of-band data, but still provides
    reliable call traces.}
  \label{tab:instr}
\end{table}

We can see that there are currently four unwinders available for the Linux
kernel. We described them and their upstream status. They all have their pros
and cons which we described too. In Table~\ref{tab:instr}, we sum up the
upstream unwinders and how their precision varies. We took call traces from
all three upstream ones using our crasher module. We did not include the
output of the \dwarf unwinder, as it gives exactly the same output as \orc. So
we made an union of the traces and put them into the table. We can see the
effect of putting every kernel code address in the output (see
Section~\ref{sec:rewrite}). When some unwinder is turned on, it marks unknown
addresses with `\code{?}'.

We can see that the Guess unwinder has no clue. The Frame pointer unwinder
skips most uninteresting entries, but can make a mistake. \orc is the most
precise, it leaves only entries which really belong there.

It was also shown, that after the years of \dwarf unwinder in our
repositories, we can finally switch from it to an upstream solution -- the
\orc unwinder which emerged recently and is being pushed to upstream as of
now.

\bibliographystyle{plain}
\bibliography{paper}

\end{document}
